### 2025-01-04 manuscript/manuscript.Rmd
First draft of manuscript introduction

---
title: "Databases are an effective and efficient method for storage and access of mass-spectrometry data"
author:
  - William Kumler
  - Anitra E. Ingalls
output: word_document
bibliography:
- Exported Items.bib
csl: journal-of-proteome-research.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

https://researcher-resources.acs.org/publish/author_guidelines?coden=jprobs

# Abstract

(200 words max)

# Introduction

Mass spectrometry (MS) still lacks a performant data access format. The mzML file type [@Martens2011], a result of over a decade of interlaboratory collaboration and workshopping, struggles to provide rapid computational access to the m/z and intensity tuples. This is the crucial component in nearly all mass spectrometry analysis, but mzML’s text-based XML format requires time-consuming decompression performed one scan at a time. This is largely due to its preservation of the scan as the unit of transaction while the field moves increasingly away from single-scan analysis [@Ting2015].

As a result, alternative file formats are proposed practically every year. These include direct improvements to the mzML format with indexing [@Rost2015] and better internal encoding of the data [@Bhamber2021], HDF5-based alternatives [@Bhamber2021; @Wilhelm2012; @Bilbao2023; @Tully2020; @Askenazi2017], relational databases [@Shah2010; @Bouyssie2015; @Handy2017; @Yang2022; @Beagley2009], or fully custom alternatives [@Rompp2011; @Lu2022]. Fundamentally, these alternatives implement tradeoffs between user sanity in exchange for access speed and/or size on disk with clever compression algorithms and modern data structures that move away from the human-readable format of the mzML. These optimized formats are inherently more difficult to understand and usually lack comprehensive documentation or examples (particularly across programming languages) making it difficult for new users to enjoy their benefits or extend their functionality. This steep learning curve, coupled with a lack of support in conversion tools such as Proteowizard’s msconvert [@Chambers2012], has prevented widespread adoption of these new formats despite their clear computational advantages. Such formats are also fragile in the sense that without community support, their continued development depends entirely on the original developers and easily become deprecated (i.e. YAFMS, Shaduf, mz5). A simple, speedy, and small MS data format remains very much in demand.

Relational databases are not new for MS workflows (see references above) and compete predominantly with HDF5-based methods. Both of these systems are widely used for big data and can be applied to MS data in a plethora of ways, leading to the proliferation of implementations we see today. Both backends provide excellent universality, larger-than-memory support, and rapid access to data, but HDF5-based systems excel at self-description and hierarchical structures [@Askenazi2017] while the relational database model is optimized for multi-table queries using a consistent syntax [@ref original database paper?]. Relational databases are increasingly seen in MS workflows for both raw and processed data, with SQLite backends now supported in the popular peakpicking software xcms [@Smith2006] via the Spectra package [@Rainer2022] (though in-memory and HDF5 options are also supported) and on MetabolomicsWorkbench [@Sud2016] while the development of MassQL [@Jarmusch2022] demonstrates the increasing comfort that MS analysts have with the adoption of SQL.

Relational databases also have several distinct advantages over hierarchical or text-based systems, particularly in performing searches for subsets of data via indices. Importantly, this indexing differs from the byte-offset indexes that already exist in the indexed mzML and HDF5 formats because the search for a particular subset cannot be done efficiently with a byte-offset index when the m/z data is encoded, though access to a particular scan can be incredibly rapid. Additionally, data from multiple samples can be stored together in a single table, differing from formats like mzDB, mzTree, and mzMD and allowing queries of all dataset samples to be performed without looping through each file in turn, thereby avoiding the associated computational overhead and query complexity.

SQL databases also allow mass spectrometrists to access the continual improvements and long-term stability produced by the industries who specialize in these. While HDF5 is a common scientific data format, databases are constantly under development by industry titans deeply invested in their maintenance and optimization. Online analytical processing (OLAP) methods are particularly well suited for MS data given their optimization for read speed, making modern systems such as DuckDB [@Raasveldt2019] or Apache Arrow’s columnar formats highly appealing while preserving the familiar serverless approach.

Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table [@Kumler2022] and here we logically extend that method into proper database storage and access. Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity tuples is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control. We compare the time and space required to extract a representative data subset under three conditions (single scan, ion chromatogram, and all scans within a retention time range) and perform these tests on multiple databases as well as mzML and other MS data formats across multiple MS experiments with varying magnitudes for comparison.

# Experimental section

This “vanilla” database deprioritizes the metadata associated with a given file and focuses on performance and simplicity. Thus, this method does not replace the existing vendor-specific or mzML files but instead constructs a parallel data structure and represents our intuition that metadata is rarely the main target of MS analysis and that labs typically preserve the original files anyway.

# Results

# Discussion

# Conclusion

# Acknowledgements

# Data availability

# References

# Supplement

### 2025-01-15 src/mzsql/duckdb_functions.py
Adding duckdb MS2 things

@@ -20,7 +20,8 @@ def turn_mzml_duckdb(file, outfile, ordered=None):
     conn = duckdb.connect(outfile)
     conn.execute("DROP TABLE IF EXISTS MS1")
 
-    scan_dfs = []
+    MS1_dfs = []
+    MS2_dfs = []
     for spectrum in pyteomics.mzml.MzML(file):
         if spectrum['ms level'] == 1:
             idx = int(spectrum['id'].split("scan=")[-1].split()[0])
@@ -28,13 +29,25 @@ def turn_mzml_duckdb(file, outfile, ordered=None):
             int_vals = spectrum['intensity array']
             rt_val = spectrum['scanList']['scan'][0]['scan start time']
             df_scan = pd.DataFrame({'id': idx, 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            scan_dfs.append(df_scan)
+            MS1_dfs.append(df_scan)
+        if spectrum["ms level"] == 2:
+            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
+            premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
+            mz_vals = spectrum['m/z array']
+            int_vals = spectrum['intensity array']
+            rt_val = spectrum['scanList']['scan'][0]['scan start time']
+            df_scan = pd.DataFrame({'id': idx, 'premz': premz_val, 'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+            MS2_dfs.append(df_scan)
 
-    all_pds = pd.concat(scan_dfs, ignore_index=True)
+    all_MS1 = pd.concat(MS1_dfs, ignore_index=True)
+    all_MS2 = pd.concat(MS2_dfs, ignore_index=True)
     if ordered is not None:
-        all_pds.sort_values(by=ordered, inplace=True)
+        all_MS1.sort_values(by=ordered, inplace=True)
+        all_MS2.sort_values(by=ordered, inplace=True)
 
-    conn.execute("CREATE TABLE MS1 AS SELECT * FROM all_pds")
+    conn.execute("CREATE TABLE MS1 AS SELECT * FROM all_MS1")
+    conn.execute("CREATE TABLE MS2 AS SELECT * FROM all_MS2")
+
[…]

### 2025-01-15 src/mzsql/sqlite_functions.py
Adding sqlite MS2 things

@@ -19,41 +19,35 @@ def turn_mzml_sqlite(file, outfile, ordered=None):
 
     Returns:
     - str: The path to the created SQLite database file.
-
-    Raises:
-    - sqlite3.Error: If there is an issue with SQLite operations.
-    - KeyError: If the input mzML file does not contain expected fields.
-    - ValueError: If the ordered column name is invalid.
     """
     if not os.path.exists(file):
         raise FileNotFoundError(f"Input mzML file '{file}' does not exist.")
 
     conn = sqlite3.connect(outfile)
-    try:
-        conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS2")
 
-        for spectrum in pyteomics.mzml.MzML(file):
-            if spectrum['ms level'] == 1:
-                try:
-                    idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-                    mz_vals = spectrum['m/z array']
-                    int_vals = spectrum['intensity array']
-                    rt_val = spectrum['scanList']['scan'][0]['scan start time']
-                except KeyError as e:
-                    raise KeyError(f"Missing expected key in mzML spectrum data: {e}")
-
-                df_scan = pd.DataFrame({'id': idx, 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                df_scan.to_sql("MS1", conn, if_exists="append", index=False)
-
-        if ordered in ['mz', 'int', 'rt']:
-            index_name = f"idx_{ordered}"
-            conn.execute(f"CREATE INDEX IF NOT EXISTS {index_name} ON MS1 ({ordered})")
-        elif ordered is not None:
-            raise ValueError(f"Invalid column for indexing: {ordered}. Must be one of 'mz', 'int', or 'rt'.")
-    except sqlite3.Error as e:
-        raise sqlite3.Error(f"SQLite error occurred: {e}")
-    finally:
-        conn.close()
+    for spectrum in pyteomics.mzml.MzML(file):
+        if spectrum['ms level'] == 1:
+            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
[…]

### 2025-01-15 src/mzsql/mza_functions.py
Making mza functions apply to MS1 data only

@@ -23,12 +23,13 @@ def get_chrom_mza(file, mz, ppm):
     scan_dfs = []
     file_keys = sorted(mza["Arrays_intensity"].keys(), key=lambda x: int(x))
     for index, scan_num in enumerate(file_keys):
-        scan_df=pd.DataFrame({
-            "rt": mza["Metadata"][index][6],
-            "mz": mza["Arrays_mz/"+scan_num][...],
-            "int": mza["Arrays_intensity/"+scan_num][...]
-        })
-        scan_dfs.append(scan_df)
+        if mza["Metadata"][index]["MSLevel"] == 1:
+            scan_df=pd.DataFrame({
+                "rt": mza["Metadata"][index]["RetentionTime"],
+                "mz": mza["Arrays_mz/"+scan_num][...],
+                "int": mza["Arrays_intensity/"+scan_num][...]
+            })
+            scan_dfs.append(scan_df)
     mza.close()
     
     file_df = pd.concat(scan_dfs, ignore_index=True)
@@ -72,14 +73,15 @@ def get_rtrange_mza(file, rtstart, rtend):
     scan_dfs = []
     file_keys = sorted(mza["Arrays_intensity"].keys(), key=lambda x: int(x))
     for index, scan_num in enumerate(file_keys):
-        rt_val = mza["Metadata"][index][6]
-        if(rtstart < rt_val < rtend):
-            scan_df=pd.DataFrame({
-                "rt": mza["Metadata"][index][6],
-                "mz": mza["Arrays_mz/"+scan_num][...],
-                "int": mza["Arrays_intensity/"+scan_num][...]
-            })
-            scan_dfs.append(scan_df)
+        if mza["Metadata"][index]["MSLevel"] == 1:
+            rt_val = mza["Metadata"][index]["RetentionTime"]
+            if(rtstart < rt_val < rtend):
+                scan_df=pd.DataFrame({
+                    "rt": mza["Metadata"][index][6],
+                    "mz": mza["Arrays_mz/"+scan_num][...],
+                    "int": mza["Arrays_intensity/"+scan_num][...]
+                })
+                scan_dfs.append(scan_df)
     mza.close()
     rtrange_data = pd.concat(scan_dfs, ignore_index=True)
     return(rtrange_data)

### 2025-01-15 src/mzsql/mza_functions.py
Adding MZA MS2 things

@@ -86,6 +86,77 @@ def get_rtrange_mza(file, rtstart, rtend):
     rtrange_data = pd.concat(scan_dfs, ignore_index=True)
     return(rtrange_data)
 
+def get_MS2scan_mza(file, spectrum_idx):
+    mza = h5py.File(file, 'r')
+    intensities = mza["Arrays_intensity/"+str(spectrum_idx)][:]
+    mz = mza["Arrays_mz/"+str(spectrum_idx)][:]
+    mza.close()
+    spec_df = pd.DataFrame({"fragmz":mz, "int":intensities})
+    return(spec_df)
+
+def get_MS2premz_mza(mza_file, precursor_mz, ppm_acc):
+    mzmin, mzmax = pmppm(precursor_mz, ppm_acc)
+    scan_dfs = []
+    mza = h5py.File(mza_file, 'r')
+    for entry in mza["Metadata"]:
+        if entry['MSLevel'] == 2 and mzmin <= entry['IsolationWindowTargetMz'] <= mzmax:
+            scan_df=pd.DataFrame({
+                    "rt": entry['RetentionTime'],
+                    "premz": entry['IsolationWindowTargetMz'],
+                    "fragmz": mza["Arrays_mz/"+str(entry["Scan"])][:],
+                    "int": mza["Arrays_intensity/"+str(entry["Scan"])][:]
+                })
+            scan_dfs.append(scan_df)
+    mza.close()
+    spec_data = pd.concat(scan_dfs, ignore_index=True)
+    return(spec_data)
+
+def get_MS2fragmz_mza(mza_file, fragment_mz, ppm_acc):
+    mzmin, mzmax = pmppm(fragment_mz, ppm_acc)
+    scan_dfs = []
+    mza = h5py.File(mza_file, 'r')
+
+    file_keys = sorted(mza["Arrays_intensity"].keys(), key=lambda x: int(x))
+    for index, scan_num in enumerate(file_keys):
+        if mza["Metadata"][index]["MSLevel"] == 2:
+            frag_mz_idxs = (mzmin < mza[f"Arrays_mz/{scan_num}"][...]) & (mza[f"Arrays_mz/{scan_num}"][...] < mzmax)
+            mz_vals = mza[f"Arrays_mz/{scan_num}"][...][frag_mz_idxs]
+            int_vals = mza[f"Arrays_intensity/{scan_num}"][...][frag_mz_idxs]
+            
+            scan_df=pd.DataFrame({
[…]

### 2025-01-16 src/mzsql/mzmlb_functions.py
Adding mzmlb MS2 things

@@ -68,3 +68,51 @@ def get_rtrange_mzmlb(file, rtstart, rtend):
             df_scan = pd.DataFrame({'mz':mz_vals, 'int':int_vals, 'rt':[rt_val]*len(mz_vals)})
             scan_dfs.append(df_scan)
     return(pd.concat(scan_dfs, ignore_index=True))
+
+def get_MS2scan_mzmlb(mzmlb_file, scan_num):
+    file_data = pyteomics.mzmlb.MzMLb(mzmlb_file)
+    rt_val = file_data[scan_num]['scanList']['scan'][0]['scan start time']
+    premz_val = file_data[scan_num]["precursorList"]["precursor"][0]["isolationWindow"]['isolation window target m/z']
+    return(pd.DataFrame({"rt":rt_val, "premz":premz_val, "fragmz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
+
+def get_MS2premz_mzmlb(mzmlb_file, precursor_mz, ppm):
+    mzmin, mzmax = pmppm(precursor_mz, ppm_acc)
+    scan_dfs = []
+    for spectrum in pyteomics.mzmlb.MzMLb(mzmlb_file):
+        if spectrum['ms level'] == 2:
+            rt_val = spectrum['scanList']['scan'][0]['scan start time']
+            premz_val = spectrum["precursorList"]["precursor"][0]["isolationWindow"]['isolation window target m/z']
+            if mzmin < premz_val < mzmax:
+                mz_vals=spectrum['m/z array']
+                int_vals = spectrum['intensity array']
+                df_scan = pd.DataFrame({'rt':rt_val, 'premz':premz_val, 'fragmz':mz_vals, 'int':int_vals})
+                scan_dfs.append(df_scan)
+    return(pd.concat(scan_dfs, ignore_index=True))
+
+def get_MS2fragmz_mzmlb(mzmlb_file, fragment_mz, ppm):
+    mzmin, mzmax = pmppm(fragment_mz, ppm_acc)
+    scan_dfs = []
+    for spectrum in pyteomics.mzmlb.MzMLb(mzmlb_file):
+        if spectrum['ms level'] == 2:
+            rt_val = spectrum['scanList']['scan'][0]['scan start time']
+            premz_val = spectrum["precursorList"]["precursor"][0]["isolationWindow"]['isolation window target m/z']
[…]

### 2025-01-16 manuscript/manuscript.Rmd
Adding experimental section and additional references to intro

@@
 ```{r setup, include=FALSE}
+options(timeout=3600)
+options(pillar.sigfig=7)
+options(digits=10)
 knitr::opts_chunk$set(echo = TRUE)
@@
 This is the crucial component in nearly all mass spectrometry analysis, but mzML’s text-based XML format requires time-consuming decompression performed one scan at a time.
-This is largely due to its preservation of the scan as the unit of transaction while the field moves increasingly away from single-scan analysis [@Ting2015].
+This is largely due to its preservation of the scan as the unit of transaction while the field moves increasingly away from single-scan analysis [@Rost2014; @Ting2015].
 
@@
 This steep learning curve, coupled with a lack of support in conversion tools such as Proteowizard’s msconvert [@Chambers2012], has prevented widespread adoption of these new formats despite their clear computational advantages.
-Such formats are also fragile in the sense that without community support, their continued development depends entirely on the original developers and easily become deprecated (i.e.
-YAFMS, Shaduf, mz5).
+Such formats are also fragile in the sense that without community support, their continued development depends entirely on the original developers and easily become deprecated (as is the case with YAFMS, Shaduf, and mz5, all of whom have links in their papers that currently redirect to missing webpages).
 A simple, speedy, and small MS data format remains very much in demand.
@@
 Both of these systems are widely used for big data and can be applied to MS data in a plethora of ways, leading to the proliferation of implementations we see today.
-Both backends provide excellent universality, larger-than-memory support, and rapid access to data, but HDF5-based systems excel at self-description and hierarchical structures [@Askenazi2017] while the relational database model is optimized for multi-table queries using a consistent syntax [@ref original database paper?].
+Both backends provide excellent universality, larger-than-memory support, and rapid access to data, but HDF5-based systems excel at self-description and hierarchical structures [@Askenazi2017] while the relational database model is optimized for multi-table queries using a consistent syntax [@Cobb1970].
 Relational databases are increasingly seen in MS workflows for both raw and processed data, with SQLite backends now supported in the popular peakpicking software xcms [@Smith2006] via the Spectra package [@Rainer2022] (though in-memory and HDF5 options are also supported) and on MetabolomicsWorkbench [@Sud2016] while the development of MassQL [@Jarmusch2022] demonstrates the increasing comfort that MS analysts have with the adoption of SQL.
@@
 While HDF5 is a common scientific data format, databases are constantly under development by industry titans deeply invested in their maintenance and optimization.
-Online analytical processing (OLAP) methods are particularly well suited for MS data given their optimization for read speed, making modern systems such as DuckDB [@Raasveldt2019] or Apache Arrow’s columnar formats highly appealing while preserving the familiar serverless approach.
+Online analytical processing (OLAP) methods are particularly well suited for MS data given their optimization for read speed under the assumption of infrequent transactions, making modern systems such as DuckDB [@Raasveldt2019] or Apache Arrow’s columnar formats highly appealing while preserving the familiar serverless approach.
 
-Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table [@Kumler2022] and here we logically extend that method into proper database storage and access.
+Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage and access.
 Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity tuples is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
@@
 
+We performed a literature search for mass-spectrometry data formats that have been published in the last 15 years and attempted to find or construct parsers for each format in Python, a popular high-level interpreted language.
+For formats which had multiple parsers available (mzML), we performed intercomparisons between the different parsers and used the best one available for each query type.
+
+Each parser was written to perform three common exploratory data analysis operations on full-scan data and four common operations on MS/MS fragmentation data.
+Full scan queries consisted of single scan extraction by scan number, retention time range extraction of all scans within a specified retention time range, and chromatogram extraction, which collects the ions within a specified parts-per-million (ppm) error of a known mass.
+These queries generally correspond to the methods used in @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser.
+MS/MS queries involved four relevant queries: a single scan extraction by scan number similar to that of the full scan; extraction of all the fragments associated with a precursor *m/z* within a given ppm; extraction of all fragments with *m/z* values within a given ppm; and finally a neutral loss query computed by subtracting the fragment mass from the precursor mass where the difference between the two fell within a specified ppm window around a given *m/z*.
+
+## Mass-spectrometry files
+
+Metabolights search for
+1. MS1 & MS2 data
+2. .raw or .wiff file types (mzDB doesn't support .d or .mzML or .mzXML)
+3. 100+ GB folder
+4. Positive mode only(?)
+
+```{r}
+download.file("https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS10096/FILES/DERIVED_FILES/POS/Feuille_Extrait_1_ESI_pos_01_7273.mzXML", mode = "wb", destfile = "../massive_data/Feuille_Extrait_1_ESI_pos_01_7273.mzXML")
+
+download.file("ftp://massive.ucsd.edu/v01/MSV000080030/peak/Forensic_study_80_volunteers/Forensic_Hands_mzXML/Forensic_Hands_plate_1_mzXML/Sample/P1_1D10v7_Palm_GD10_01_24008.mzXML", "../massive_data/P1_1D10v7_Palm_GD10_01_24008.mzXML", mode = "wb")
+
+download.file("https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS417/FILES/CS52684_pos_SWATH.mzXML", mode = "wb", destfile = "../massive_data/CS52684_pos_SWATH.mzXML")
+```
+
+Converted via msconvert or custom implementation
+
+```{shell}
+cd Documents\Will\mzsql
+conda activate mzsql
+
+E:
+cd mzsql\MTBLS10066
+
+msconvert *.raw --filter "peakPicking true 1-" --filter "polarity positive" 
+# Also convert neg mode to see what benefit to offer pos/neg combos
+msconvert *.raw --filter "peakPicking true 1-" --filter "polarity negative" -o "negative"
+msconvert *.raw --mzMLb --filter "peakPicking true 1-" --filter "polarity positive" 
+msconvert *.raw --mz5 --filter "peakPicking true 1-" --filter "polarity positive"
+
+for %f in (*.raw) do "C:\Program Files\ProteoWizard\raw2mzDB_0.9.10_build20170802\raw2mzDB.exe" -i %f -c 1-2 --safeMode
+
+for %f in (*.raw) do mza.exe -file %f
+```
+
+## Python-based file parsers
+
+```{r}
+git clone https://github.com/wkumler/mzsql
+```
+
+Comparisons between different mzML methods
+
+Some details about the parsers I wrote, nuances of which can be looped and which can't
+
+## Database schema
+
 This “vanilla” database deprioritizes the metadata associated with a given file and focuses on performance and simplicity.
@@
 
+Discuss how metadata fits into these parsers
+
+Discuss links between various tables via scan_idx column?
+
+$MS1
+
+$MS2
+
+$scan_info
+
+$file_info
+
+--
+
+$rt_correction
+
+$picked_peaks
+
+## Time and space testing
+
+Via %timeit% (need to redo these with randomization)
+
 # Results
 
+Which file types we were able to find/write parsers for
+
 # Discussion
+
+While the gap between data scientist and mass spectrometrist continues to narrow, current implementations of MS data storage systems almost seem designed to *increase* it.
+This is accomplished by extremely poor documentation that makes it difficult to do anything beyond the original designer's intent, typically with data structures that are not familiar to the MS expert.
 

### 2025-01-17 src/mzsql/mzml_functions.py
Adding pyopenms MS2 things (nloss needs work)

@@ -71,6 +71,54 @@ def get_rtrange_mzml_pyteomics(file, rtstart, rtend):
     return(pd.concat(scan_dfs, ignore_index=True))
 
 
+def get_MS2scan_mzml_pyteomics(mzml_file, scan_num):
+    file_data = pyteomics.mzml.MzML(mzml_file)
+    rt_val = file_data[scan_num]['scanList']['scan'][0]['scan start time']
+    premz_val = file_data[scan_num]["precursorList"]["precursor"][0]["isolationWindow"]['isolation window target m/z']
+    return(pd.DataFrame({"rt":rt_val, "premz":premz_val, "fragmz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
+
+def get_MS2premz_mzml_pyteomics(mzml_file, precursor_mz, ppm):
+    mzmin, mzmax = pmppm(precursor_mz, ppm_acc)
+    scan_dfs = []
+    for spectrum in pyteomics.mzml.MzML(mzml_file):
+        if spectrum['ms level'] == 2:
+            rt_val = spectrum['scanList']['scan'][0]['scan start time']
+            premz_val = spectrum["precursorList"]["precursor"][0]["isolationWindow"]['isolation window target m/z']
+            if mzmin < premz_val < mzmax:
+                mz_vals=spectrum['m/z array']
+                int_vals = spectrum['intensity array']
+                df_scan = pd.DataFrame({'rt':rt_val, 'premz':premz_val, 'fragmz':mz_vals, 'int':int_vals})
+                scan_dfs.append(df_scan)
+    return(pd.concat(scan_dfs, ignore_index=True))
+
+def get_MS2fragmz_mzml_pyteomics(mzml_file, fragment_mz, ppm):
+    mzmin, mzmax = pmppm(fragment_mz, ppm_acc)
+    scan_dfs = []
+    for spectrum in pyteomics.mzml.MzML(mzml_file):
+        if spectrum['ms level'] == 2:
+            rt_val = spectrum['scanList']['scan'][0]['scan start time']
+            premz_val = spectrum["precursorList"]["precursor"][0]["isolationWindow"]['isolation window target m/z']
+            chosen_frag_idxs = (mzmin < spectrum['m/z array']) & (spectrum['m/z array'] < mzmax)
[…]

### 2025-01-17 src/mzsql/mzml_functions.py
Finished mzML things

@@ -278,7 +278,7 @@ def get_MS2nloss_mzml_pyopenms(mzml_file, neutral_loss, ppm_acc):
             premz_val = spectrum.getPrecursors()[0].getMZ()
             rt_val = spectrum.getRT()
             mz_vals, int_vals = spectrum.get_peaks()
-            bet_idxs = (mzmin < premz_val-mz_vals) & (mz_vals < premz_val-mzmax)
+            bet_idxs = (mzmin < premz_val-mz_vals) & (premz_val-mz_vals < mzmax)
             if(sum(bet_idxs)>0):
                 df_scan = pd.DataFrame({"rt":rt_val, "premz":premz_val, "fragmz":mz_vals[bet_idxs], "int":int_vals[bet_idxs]})
                 scan_dfs.append(df_scan)
@@ -306,14 +306,15 @@ def get_chrom_mzml_pymzml(file, mz, ppm):
     mzmin, mzmax = pmppm(mz, ppm)
     scan_dfs = []
     for spectrum in run:
-        rt_val = spectrum.scan_time_in_minutes()
-        mz_vals = spectrum.mz
-        int_vals = spectrum.i
-        bet_idxs = (mzmin < mz_vals) & (mz_vals < mzmax)
-    
-        if(sum(bet_idxs)>0):
-            df_scan = pd.DataFrame({'mz':mz_vals[bet_idxs], 'int':int_vals[bet_idxs], 'rt':[rt_val]*sum(bet_idxs)})
-            scan_dfs.append(df_scan)
+        if(spectrum.ms_level==1):
+            rt_val = spectrum.scan_time_in_minutes()
+            mz_vals = spectrum.mz
+            int_vals = spectrum.i
+            bet_idxs = (mzmin < mz_vals) & (mz_vals < mzmax)
+        
+            if(sum(bet_idxs)>0):
+                df_scan = pd.DataFrame({'mz':mz_vals[bet_idxs], 'int':int_vals[bet_idxs], 'rt':[rt_val]*sum(bet_idxs)})
+                scan_dfs.append(df_scan)
     return(pd.concat(scan_dfs, ignore_index=True))
 
 def get_spec_mzml_pymzml(file, scan_num):
@@ -345,10 +346,65 @@ def get_rtrange_mzml_pymzml(file, rtstart, rtend):
     run = pymzml.run.Reader(file, build_index_from_scratch=True)  
     scan_dfs = []
     for spectrum in run:
-        rt_val = spectrum.scan_time_in_minutes()
[…]

### 2025-01-17 src/mzsql/mzdb_functions.py
Filtering for MS1 only now

@@ -73,6 +73,7 @@ def get_chrom_mzdb(file, mz, ppm):
     FROM run_slice, bounding_box, spectrum
     WHERE bounding_box.run_slice_id = run_slice.id
     AND bounding_box.first_spectrum_id = spectrum.id
+    AND ms_level = 1
     AND begin_mz = ?
     """
     bb_dataframe = pd.read_sql(bb_query, connection, params=(bb_id_for_chrom,))
@@ -133,6 +134,7 @@ def get_rtrange_mzdb(file, rtstart, rtend):
     SELECT DISTINCT bounding_box.data
     FROM bounding_box, spectrum
     WHERE bounding_box.first_spectrum_id = spectrum.bb_first_spectrum_id
+    AND ms_level = 1
     AND spectrum.time BETWEEN ? AND ?
     """
     bb_dataframe = pd.read_sql(bb_query, connection, params=(rtstart*60, rtend*60))

### 2025-01-31 src/mzsql/mz5_functions.py
Fixing mz5 rtrange extraction

@@ -36,14 +36,14 @@ def get_spec_mz5(file, scan_num):
     return(pd.DataFrame({"mz":mz_vals, "int":int_vals}))
 
 def get_rtrange_mz5(file, rtstart, rtend):
-    print("Warning: get_rtrange_mz5 does not yet seem to be functional")
-    print("Warning: returns scans without relevant spectra")
     mz5_file = h5py.File(file, 'r')
     scan_idxs = np.concatenate(([0], mz5_file["SpectrumIndex"][...]))
     scan_dfs = []
+    has_precursor = [len(item)>0 for item in mz5_file["SpectrumMetaData"]["precursors"]]
+    
     for index, rt_val in enumerate(mz5_file["ChomatogramTime"][...]):
         if(rtstart < rt_val < rtend):
-            if(len(mz5_file["SpectrumMetaData"]["precursors"][index])==0):
+            if(not(has_precursor[index])):
                 scan_df = pd.DataFrame({
                     "rt": rt_val,
                     "mz": np.cumsum(mz5_file["SpectrumMZ"][scan_idxs[index]:scan_idxs[index+1]]),

### 2025-01-31 manuscript/scripts/singlefile_timing.py
Initial timing from ILC

@@ -1,0 +1,119 @@
+
+# Setup things
+import numpy as np
+import pandas as pd
+import sqlite3
+import os
+from mzsql import *
+
+# Create databases from mzML files if they don't already exist
+if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite"))):
+    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
+    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.duckdb")
+    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds")
+
+# Identify the bits of information to be extracted
+conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
+# Get 10 random spectra
+rand_scan_query = """
+SELECT id 
+FROM (SELECT DISTINCT id FROM {0}) 
+ORDER BY RANDOM() 
+LIMIT 10
+"""
+rand_MS1_scans = pd.read_sql_query(rand_scan_query.format("MS1"), conn)["id"].tolist()
+rand_MS2_scans = pd.read_sql_query(rand_scan_query.format("MS2"), conn)["id"].tolist()
+# Get the top 10ish masses by intensity and their retention times
+top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
+top_masses = []
+top_rts = []
+for i in range(10):
+    top_rts.append(top_int_df["rt"].iloc[0])
+    top_mz = top_int_df["mz"].iloc[0]
+    top_masses.append(top_mz)
+    mzmin, mzmax = pmppm(top_mz, 10)
+    rtmin, rtmax = top_int_df["rt"].iloc[0]+(-1, 1)
+    top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax)) & ((top_int_df["rt"]<rtmin) | (top_int_df["rt"]>rtmax))]
[…]

### 2025-01-31 manuscript/manuscript.Rmd

@@
 
-```{r}
-download.file("https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS10096/FILES/DERIVED_FILES/POS/Feuille_Extrait_1_ESI_pos_01_7273.mzXML", mode = "wb", destfile = "../massive_data/Feuille_Extrait_1_ESI_pos_01_7273.mzXML")
-
-download.file("ftp://massive.ucsd.edu/v01/MSV000080030/peak/Forensic_study_80_volunteers/Forensic_Hands_mzXML/Forensic_Hands_plate_1_mzXML/Sample/P1_1D10v7_Palm_GD10_01_24008.mzXML", "../massive_data/P1_1D10v7_Palm_GD10_01_24008.mzXML", mode = "wb")
-
-download.file("https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS417/FILES/CS52684_pos_SWATH.mzXML", mode = "wb", destfile = "../massive_data/CS52684_pos_SWATH.mzXML")
-```
-
 Converted via msconvert or custom implementation
@@
 ```{shell}
-cd Documents\Will\mzsql
-conda activate mzsql
+cd Documents\Will
+git clone https://github.com/wkumler/mzsql
+cd mzsql
+
+conda create -n "mzsql_3_11" python=3.11 numpy=1.26.4
+pip install https://pypi.cs.uni-tuebingen.de/packages/pyopenms-3.0.0.dev20230306-cp311-cp311-win_amd64.whl#md5=cdce650718bea07a0193b3227a2c075b
+conda activate mzsql_3_11
+pip install -e .
+pip install git+https://github.com/PNNL-m-q/mzapy.git@no_full_mz_array#egg=mzapy
+
+cd java_things
+curl -O https://dlcdn.apache.org/maven/maven-3/3.9.9/binaries/apache-maven-3.9.9-bin.zip
+tar -xf apache-maven-3.9.9-bin.zip
+git clone https://github.com/optimusmoose/MZTree.git
+cd MZTree
+cd msDataServer
+C:\Users\willi\Documents\mzsql\java_things\apache-maven-3.9.9\bin\mvn package
+java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar target\msDataServer-1.0.jar
+(manually) open mztree file in port 4568
+
+git clone https://github.com/yrm9837/mzMD-java.git
+cd mzMD-java
+C:\Users\willi\Documents\mzsql\java_things\apache-maven-3.9.9\bin\mvn package
+java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar target\msDataServer-1.0.jar
+(manually) open mzmd file in port 4567
+cd ..
 
@@
 ## Python-based file parsers
-
-```{r}
-git clone https://github.com/wkumler/mzsql
-```
 
@@
 
---
+$sample_metadata
 
@@
 
-Which file types we were able to find/write parsers for
+## Building parsers
+
+Which file types we were able to find/write parsers for and our notes on their methodology
+
+Good documentation for pyteomics, pyopenms, pymzml, mzmlb (at least via pyteomics)
+Middling documentation for mza/mzapy, mzDB
+Terrible documentation for Aird, MzTree, mzMD, mz5
+Nonexistent documentation for YAFMS, Shaduf
+
+pyopenms requires an old version of Python (3.11) and numpy (1.26.4) because of AVX support - discovered via GitHub issue with no requirements.txt file or environment.yml.
+Therefore requires using old version of the package.
+
+Aird has zero documentation or application support (Python and C# versions are functionally deprecated)
+
+Toffee only supports TOF data
+
+MzTree and mzMD have no MS2+ support and don't support scan extraction
+Neither of them handle multiple files
+
+mz5 doesn't make it clear how to access the precursor masses for MS2 things so those parsers are missing, ref mzMLb paper here
+
+mzDB handles metadata as XML strings and doesn't support bounding boxes for MS2 things afaics
+Doesn't support .d files (only .raw and .wiff)
+Doesn't handle polarity switching data
+
+MZA can't perform centroiding simultaneously with .raw conversion, requires switch from .mzML
+MZA doesn't support .wiff (only .raw or .d)
+mzapy officially only supports .d (.raw requires installing from a separate development branch)
+mzapy doesn't support direct scan extraction - baffling given how well the file type exposes this
+mzapy doesn't index by MS2 things (only RT/DT)
+
+Several fundamental inefficiencies identified:
+1. Looping over each scan within a file
+2. Looping over each file within a dataset
+3. Decoding mz/int information each time if filtering on these fields
+4. RTs are not accessible outside of looping over each scan for RT subsetting
+5. Scan ids are not accessible outside of looping over each scan for scan extraction
+
+## Timings for a single file
+
+## Timings for a dataset with multiple files (and file sizes?)
+
+## Database optimization 
+1) Indices and ordering
+2) Peak data extraction via joins vs loops vs unified query
 
@@
 This is accomplished by extremely poor documentation that makes it difficult to do anything beyond the original designer's intent, typically with data structures that are not familiar to the MS expert.
+
+Not a single parser was without flaws - all were either slow, poorly documented, or had highly limited functionality
+
+Querying things by scan number is dumb.
+1) Different indices for pyopenms vs pymzml vs pyteomics
+2) idx = scan_num only works for consecutive scans (no multi-experiment, no polarity switching, no filtering)
+3) Only USEFUL when you're parsing the entire file! Nobody knows what scan number 976 corresponds to, making extraction by number useless
 

### 2025-01-31 manuscript/scripts/singlefile_timing.py
Switching to mzapy only for chrom and rtrange

@@ -60,7 +60,6 @@ function_list = [
     ("mzMLb", "mzmlb", ".mzMLb"),
     ("mzDB", "mzdb", ".raw.mzDB"),
     ("MZA", "mza", ".mza"),
-    ("mzapy", "mzapy", ".mza"),
     ("mz5", "mz5", ".mz5"),
     ("SQLite", "sqlite", ".sqlite"),
     ("DuckDB", "duckdb", ".duckdb"),

### 2025-01-31 manuscript/scripts/singlefile_timing.py
Adding file size measurement to singlefile script

@@ -116,3 +116,21 @@ for target_frag in top_fragments:
         time_df.to_csv('data/singlefile_times.csv', mode='a', header=False, index=False)
 
 print("Success!")
+
+file_sizes = {}
+for name, suffix, filetype in function_list:
+    if(name == "mzMD"):
+        file_sizes[suffix] = os.path.getsize("E:/mzsql/MTBLS10066/result.mzMD")
+    elif(name == "MZTree"):
+        file_sizes[suffix] = os.path.getsize("E:/mzsql/MTBLS10066/01-30-2025_10-35-06.mzTree")
+    elif(name == "Parquet"):
+        dirsize = sum(
+            os.path.getsize(os.path.join(root, f)) 
+            for root, _, files in os.walk("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds") 
+            for f in files
+        )
+        file_sizes[suffix] = dirsize
+    else:
+        file_sizes[suffix] = os.path.getsize(f"E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01{filetype}")
+
+pd.DataFrame(file_sizes.items(), columns=["method", "file_size"]).to_csv("data/file_sizes.csv", index=False)

### 2025-01-31 manuscript/scripts/singlefile_timing.py
Tidying up timing script

@@ -61,11 +61,11 @@ function_list = [
     ("mzDB", "mzdb", ".raw.mzDB"),
     ("MZA", "mza", ".mza"),
     ("mz5", "mz5", ".mz5"),
+    ("MZTree", "mztree", "http://127.0.0.1:4568"),
+    ("mzMD", "mzMD", "http://127.0.0.1:4567"),
     ("SQLite", "sqlite", ".sqlite"),
     ("DuckDB", "duckdb", ".duckdb"),
-    ("Parquet", "parquet", "_pqds"),
-    ("MZTree", "mztree", "http://127.0.0.1:4568"),
-    ("mzMD", "mzMD", "http://127.0.0.1:4567")
+    ("Parquet", "parquet", "_pqds")
 ]
 
 # Create a CSV to store the data in
@@ -75,7 +75,7 @@ init_df.to_csv('data/singlefile_times.csv', index=False)
 # And perform the timings!
 for target_spec in rand_MS1_scans:
     print(target_spec)
-    for name, suffix, filetype in [function_list[i] for i in [0, 1, 3, 4,5,6,8,9,10,11]]:
+    for name, suffix, filetype in [function_list[i] for i in [0, 1, 3, 4,5,6,7,10,11,12]]:
         time_taken = time_spec(suffix, filetype, target_spec, verbose=True)
         time_df = pd.DataFrame({"query":["ms1_spec"], "target":[target_spec], "method":[name], "time":time_taken})
         time_df.to_csv('data/singlefile_times.csv', mode='a', header=False, index=False)
@@ -96,41 +96,41 @@ for target_rt in top_rts:
 
 for target_spec in rand_MS2_scans:
     print(target_spec)
-    for name, suffix, filetype in [function_list[i] for i in [0, 1, 3, 4,5,6,8,9,10,11]]:
+    for name, suffix, filetype in [function_list[i] for i in [0, 1, 3, 4, 5, 6, 7, 10, 11, 12]]:
         time_taken = time_spec(suffix, filetype, target_spec, verbose=True)
         time_df = pd.DataFrame({"query":["ms2_spec"], "target":[target_spec], "method":[name], "time":time_taken})
         time_df.to_csv('data/singlefile_times.csv', mode='a', header=False, index=False)
 
 for target_pre in top_precursors:
     print(target_pre)
-    for name, suffix, filetype in [function_list[i] for i in [0, 1, 3, 4,5,6,9,10,11]]:
[…]

### 2025-01-31 manuscript/scripts/singlefile_timing.py
New data

@@ -120,7 +120,10 @@ for name, suffix, filetype in function_list:
     if(name == "mzMD"):
         file_sizes[name] = os.path.getsize("E:/mzsql/MTBLS10066/result.mzMD")
     elif(name == "MZTree"):
-        file_sizes[name] = os.path.getsize("E:/mzsql/MTBLS10066/01-30-2025_10-35-06.mzTree")
+        total_size = sum([
+            os.path.getsize("E:/mzsql/MTBLS10066/01-30-2025_10-35-06.mzTree"),
+            os.path.getsize("E:/mzsql/MTBLS10066/01-30-2025_10-35-06.mzTree-points")
+        ])
     elif(name == "Parquet"):
         dirsize = sum(
             os.path.getsize(os.path.join(root, f)) 

### 2025-01-31 manuscript/figscript.R
Fixing merge conflict

@@ -1,221 +1,78 @@
 
-
 library(tidyverse)
 
+method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
+                  "mzMLb", "mzDB", "MZA", "mz5", "MZTree", "mzMD", 
+                  "SQLite", "DuckDB", "Parquet")
 
-all_timings <- c("chrom", "spec", "rtrange") %>%
-  map(function(metric){
-  read_csv(paste0(metric, "_timing_data.csv")) %>%
-    pivot_longer(everything()) %>%
-    mutate(metric=metric)
-}) %>%
-  bind_rows() %>%
-  mutate(category=case_when(
-    str_detect(name, "pyteomics|pyopenms|pymzml")~"mzML",
-    str_detect(name, "SQLite|DuckDB")~"Database",
-    TRUE~name
-  )) %>%
-  mutate(name=ifelse(name=="MZA (package)", "mzapy", name)) %>%
-  mutate(name=fct_inorder(name)) %>%
-  mutate(category=fct_inorder(category)) %>%
-  mutate(metric=factor(metric, levels=c("spec", "chrom", "rtrange"),
-                       labels=c("Spectrum extraction",
-                                "Chromatogram extraction",
-                                "Retention time range")))
+all_timings <- read_csv("data/singlefile_times.csv") %>% 
+  mutate(method=factor(method, levels=method_names)) %>%
+  mutate(query=factor(query, levels=c("ms1_scan", "chrom", "rtrange",
+                                      "ms2_scan", "premz", "fragmz"),
+                      labels=c("MS1 scan", "Chromatogram", "RT Range",
+                               "MS2 scan", "Precursor search",
+                               "Fragment search")))
+all_sizes <- read_csv("data/file_sizes.csv") %>%
+  mutate(method=factor(method, levels=method_names))
 
 all_timings %>%
-  filter(category=="mzML") %>%
-  filter(!str_detect(name, "idx")) %>%
-  filter(metric=="Spectrum extraction") %>%
-  ggplot(aes(x=name, y=value)) +
-  geom_boxplot(color="black") +
-  geom_jitter(width = 0.1, alpha=0.5) +
-  geom_hline(yintercept = 0) +
[…]

### 2025-01-31 manuscript/scripts/figscript.R
Tidying figure code

@@ -7,8 +7,8 @@ method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
 
 all_timings <- read_csv("data/singlefile_times.csv") %>% 
   mutate(method=factor(method, levels=method_names)) %>%
-  mutate(query=factor(query, levels=c("ms1_scan", "chrom", "rtrange",
-                                      "ms2_scan", "premz", "fragmz"),
+  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
+                                      "ms2_spec", "premz", "fragmz"),
                       labels=c("MS1 scan", "Chromatogram", "RT Range",
                                "MS2 scan", "Precursor search",
                                "Fragment search")))
@@ -45,7 +45,7 @@ all_timings %>%
   scale_y_log10() +
   labs(x="File size (MB)", y="Time (seconds)", color="File type", fill="File type") +
   theme_bw()
-ggsave("figures/singlefile_fig.png")
+
 
 
 
@@ -75,4 +75,7 @@ size_barplot <- all_sizes %>%
   theme_bw() +
   theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), legend.position = "none")
 library(patchwork)
-time_boxplot + size_barplot + plot_layout(widths = c(3, 1))
+time_boxplot + size_barplot + plot_layout(widths = c(3.5, 1))
+
+ggsave("figures/singlefile_fig.png", width = 6.5, height=5)
+

### 2025-01-31 manuscript/scripts/singlefile_timing.py
Fixed sizing bug

@@ -124,6 +124,7 @@ for name, suffix, filetype in function_list:
             os.path.getsize("E:/mzsql/MTBLS10066/01-30-2025_10-35-06.mzTree"),
             os.path.getsize("E:/mzsql/MTBLS10066/01-30-2025_10-35-06.mzTree-points")
         ])
+        file_sizes[name] = total_size
     elif(name == "Parquet"):
         dirsize = sum(
             os.path.getsize(os.path.join(root, f)) 

### 2025-02-05 manuscript/scripts/figscript.R
Settling on patchwork Fig 1

@@ -15,42 +15,6 @@ all_timings <- read_csv("data/singlefile_times.csv") %>%
 all_sizes <- read_csv("data/file_sizes.csv") %>%
   mutate(method=factor(method, levels=method_names))
 
-all_timings %>%
-  ggplot() +
-  geom_boxplot(aes(x=method, y=time, color=method)) +
-  facet_wrap(~query) +
-  scale_y_log10() +
-  labs(x=NULL, y="Time (seconds)", color=NULL) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5))
-all_timings %>%
-  left_join(all_sizes) %>%
-  ggplot() +
-  geom_point(aes(x=file_size/1e6, y=time, color=method), size=4, alpha=0.5) +
-  facet_wrap(~query) +
-  scale_x_continuous(limits = c(0, NA)) +
-  scale_y_log10() +
-  labs(x="File size (MB)", y="Time (seconds)", color="File type") +
-  theme_bw()
-
-all_timings %>%
-  group_by(query, method) %>%
-  summarise(avg_time=mean(time), sd_time=sd(time)) %>%
-  left_join(all_sizes) %>%
-  ggplot() +
-  geom_segment(aes(x=file_size/1e6, xend=file_size/1e6, y=avg_time-sd_time*2, yend=avg_time+sd_time*2, color=method), linewidth = 1.2) +
-  geom_point(aes(x=file_size/1e6, y=avg_time, fill=method), size=4, pch=21, color="black") +
-  facet_wrap(~query) +
-  scale_x_continuous(limits = c(0, NA)) +
-  scale_y_log10() +
-  labs(x="File size (MB)", y="Time (seconds)", color="File type", fill="File type") +
-  theme_bw()
-
-
-
-
-
-
 time_boxplot <- all_timings %>%
   ggplot() +
   geom_boxplot(aes(x=method, y=time, color=method)) +

### 2025-02-05 manuscript/manuscript.Rmd
Adding methods and results

@@
 Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity tuples is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
-We compare the time and space required to extract a representative data subset under three conditions (single scan, ion chromatogram, and all scans within a retention time range) and perform these tests on multiple databases as well as mzML and other MS data formats across multiple MS experiments with varying magnitudes for comparison.
+We compare the time and space required to extract a representative data subset under six conditions and perform these tests on multiple databases as well as mzML and other MS data formats.
 
@@
 We performed a literature search for mass-spectrometry data formats that have been published in the last 15 years and attempted to find or construct parsers for each format in Python, a popular high-level interpreted language.
-For formats which had multiple parsers available (mzML), we performed intercomparisons between the different parsers and used the best one available for each query type.
+Each parser was written to perform three common exploratory data analysis operations on full-scan data and three common operations on MS/MS fragmentation data.
+Full scan queries consisted of 1) single scan extraction by scan number, 2) retention time range extraction of all scans within a specified retention time range, and 3) chromatogram extraction, which collects the ions within a specified parts-per-million (PPM) error of a known mass.
+These queries generally correspond to the methods used in @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser.
+Note that the chromatogram extraction does not extract a precompiled chromatogram of the sort commonly found at the end of mzML files or as a result of SIM/PRM analysis but instead refers to sifting through the raw data for data tuples with an *m/z* value between specified bounds.
+MS/MS queries involved extracting three relevant subsets, consisting of 1) a single scan extraction by scan number similar to that of the full scan, 2) extraction of all the fragments associated with a precursor *m/z* within a given PPM, and 3) extraction of all fragments with *m/z* values within a given PPM.
 
-Each parser was written to perform three common exploratory data analysis operations on full-scan data and four common operations on MS/MS fragmentation data.
-Full scan queries consisted of single scan extraction by scan number, retention time range extraction of all scans within a specified retention time range, and chromatogram extraction, which collects the ions within a specified parts-per-million (ppm) error of a known mass.
-These queries generally correspond to the methods used in @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser.
-MS/MS queries involved four relevant queries: a single scan extraction by scan number similar to that of the full scan; extraction of all the fragments associated with a precursor *m/z* within a given ppm; extraction of all fragments with *m/z* values within a given ppm; and finally a neutral loss query computed by subtracting the fragment mass from the precursor mass where the difference between the two fell within a specified ppm window around a given *m/z*.
+We explored the available documentation on PyPI and Github for each mass spectrometry data format and either identified existing functions and packages that would perform the above queries or wrote our own functions if necessary.
 
@@
 
-## Python-based file parsers
-
-Comparisons between different mzML methods
-
-Some details about the parsers I wrote, nuances of which can be looped and which can't
-
 ## Database schema
@@
 
-## Building parsers
+## All existing MS data formats demand a high level of domain knowledge
 
-Which file types we were able to find/write parsers for and our notes on their methodology
+We were able to obtain or write parsers for seven different existing mass spectrometry (MS) data formats: mzML, mzMLb, mz5, mzDB, MZA, MzTree, and mzMD.
+Multiple Python packages existed for the mzML data format so we used each of the three dominant packages (pyteomics, pyOpenMS, and pymzml) and compared their timing results as well.
+We failed to produce parsers for the YAFMS and Shaduf file types due to complete deprecation (links to these no longer exist), the toffee file type due to its application solely to time-of-flight (TOF) data-independent acquisition (DIA) data, the Aird file type due to its current deprecation in Python and C#, and the UIMF format due to a complete lack of interface documentation.
 
-Good documentation for pyteomics, pyopenms, pymzml, mzmlb (at least via pyteomics)
-Middling documentation for mza/mzapy, mzDB
-Terrible documentation for Aird, MzTree, mzMD, mz5
-Nonexistent documentation for YAFMS, Shaduf
+### File conversion support varied enormously
 
-pyopenms requires an old version of Python (3.11) and numpy (1.26.4) because of AVX support - discovered via GitHub issue with no requirements.txt file or environment.yml.
-Therefore requires using old version of the package.
+Conversion from the initial Thermo .raw file type to the open-source .mzML format was seamlessly performed by Proteowizard's `msconvert` library.
+Similarly, Proteowizard support for the .mz5 and .mzMLb file types made their conversion trivial.
 
-Aird has zero documentation or application support (Python and C# versions are functionally deprecated)
+mzDB and MZA both had reasonably good documentation, both formats notably providing self-contained extensions to `msconvert` for ease of conversion.
+However, both converters provide limited coverage, with mzDB notably missing support for Waters and Agilent .d files while MZA currently lacks support for AB Sciex .wiff and Bruker .baf files and is only available via binary executable (.exe), restricting its use to Windows platforms.
+Additionally, both parsers appear to be unable to separate scans from a polarity-switching experiment or support any of the other filters available natively in `msconvert`, as additional arguments passed to the executable throw errors instead of being passed along to the algorithm.
 
-Toffee only supports TOF data
+MZTree and its derivative, mzMD, provided significantly less documentation about the conversion process than the other file types.
+This documentation consisted solely of the README available in the associated Github repositories and their installation and deployment required rebuilding the Java applet, of which the bare-bones instructions make several assumptions about the user's PATH environmental variable.
+In the case of mzMD, no documentation for installation and build was provided and this instead needed to be deduced from MZTree.
+Additionally, we ran into issues with hardware acceleration once the GUI was launched that required extensive debugging.
+The GUI conversion, however, is straightforward once the app is correctly compiled and launched, albeit requiring a manual entry of a single file at a time with no apparent batch processing available.
 
-MzTree and mzMD have no MS2+ support and don't support scan extraction
-Neither of them handle multiple files
+### Universal lack of support for the six relevant queries
 
-mz5 doesn't make it clear how to access the precursor masses for MS2 things so those parsers are missing, ref mzMLb paper here
+Despite the relative simplicity and relevance of our queries, none of the available mass spectrometry (MS) formats had existing functions or documented examples of all six queries.
+Unsurprisingly, the mzML file type had the most extensive coverage but documentation and prebuilt functionality was still disappointingly sparse.
+The pyteomics package provides four "combined examples" that focus on the spectrum visualization and annotation common to proteomics research but provide minimal guidance about chromatogram or retention time range extraction.
+pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb.
+The pyOpenMS package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS1 data.
+Additionally, pyOpenMS required installing an old version of the package (3.0.0), Python itself (3.11) and the numpy package (<2.0) due to more recent builds requiring AVX support which was unavailable on our hardware.
+pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing.
 
-mzDB handles metadata as XML strings and doesn't support bounding boxes for MS2 things afaics
-Doesn't support .d files (only .raw and .wiff)
-Doesn't handle polarity switching data
+mz5's documentation was remarkably sparse, especially considering it was one of the earliest mzML formats and is supported by Proteowizard.
+Crucially, the original paper [@Wilhelm2012] contains links to a website (https://software.steenlab.org/mz5) which currently returns an HTTP error 500.
+A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2) and hasn't been updated in 12 years and is predominately a simple fork of h5py [@ref] with three mz5-specific commits on top.
+Most problematically, we were unable to determine how mz5 stores precursor *m/z* ratios, making the fragment and precursor searches impossible.
+This was largely due to the variable-length nested compound structures mz5 that are not supported in all APIs, e.g.
+Java (@Bhamber2020).
 
-MZA can't perform centroiding simultaneously with .raw conversion, requires switch from .mzML
-MZA doesn't support .wiff (only .raw or .d)
-mzapy officially only supports .d (.raw requires installing from a separate development branch)
-mzapy doesn't support direct scan extraction - baffling given how well the file type exposes this
-mzapy doesn't index by MS2 things (only RT/DT)
+mzDB access was hamstrung by several issues, primarily the outdated repository that implies Python and R support via a port from Rust but was unavailable at the time of development, though we are grateful for the responsive developer who notified us that this implementation was not feature-complete.
+This required that we deduce the SQLite BLOB type compression format from scratch when writing a parser and spend extensive time reading through the documentation to determine how best to link the various tables provided in the mzDB file.
+Scan metadata in this file type is stored as raw XML strings, producing the worst of both worlds in requiring both SQLite knowledge in their extraction and XML processing to obtain the relevant information.
+Additionally, its failure to implement the clever bounding box and run slice scheme for MS/MS data negated with our ability to avoid parsing every MS/MS spectrum when performing precursor and fragment searches.
 
-Several fundamental inefficiencies identified:
-1. Looping over each scan within a file
-2. Looping over each file within a dataset
-3. Decoding mz/int information each time if filtering on these fields
-4. RTs are not accessible outside of looping over each scan for RT subsetting
-5. Scan ids are not accessible outside of looping over each scan for scan extraction
+MZA provides a complementary Python package, `mzapy`, for access to MZA files.
+Here again we ran into several issues with its installation and use stemming largely from the deployed package requiring TOF bins for parsing, though a separate Github branch provides a workaround and the rapid developer response was appreciated.
+The `mzapy` package provides a clear example of chromatogram extraction as well as a method for retention time range extraction, though there exists no clear function for the extraction of a single spectrum by scan number despite the internal file structure being highly optimized for this purpose.
+`mzapy` also provides good support for ion mobility extraction but fails to index MS2 information or provide any clear way to extract fragments by *m/z* or precursor.
 
-## Timings for a single file
+MZTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server to be launched that could then be queried via an API.
+For users without prior knowledge of HTTP request methods or exposure to programming APIs, the README is entirely unhelpful because it simply documents the API's endpoints and provides no complete query strings as examples to guide the user.
+This combination of GUI server and command-line HTTP request inverts the typical paradigm of GUI for exploration and command line for construction to unnecessarily complicated results.
+More problematically for this analysis, the API provides no apparent way to access MS/MS data or query the files by scan number, with only RT and *m/z* bounds controlling the subset of data extracted.
+Finally, the GUI provides no way to open multiple files simultaneously or iterate through files programmatically and instead requiring point-and-click interaction with the GUI each time a file is opened or closed, preventing us from making reasonable comparisons in tests requiring multiple files.
 
-## Timings for a dataset with multiple files (and file sizes?)
+## SQL-based parsers were simple to write and use
+
+We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for MS1 and MS2 data.
+The MS1 table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
+The MS2 table consisted of the same fields except that the *m/z* column was separated into precursor and fragment *m/z*.
+Although we did not extend these databases to include the metadata associated with each file and scan, the logical framework could be easily extended in future work and the metadata typically represents a small fraction of the total space within the file, allowing us to make reasonable comparisons about file size between the databases and the metadata-rich other file types.
+We also converted each file's MS1 and MS2 table into CSV and Parquet representations for comparison using the same field/column schema.
+
+We found that the documentation for SQLite, DuckDB, and Parquet file formats in Python to far exceed the documentation available for any mzML parser.
+This is unsurprising given that these file formats are used widely outside of MS research and are developed and maintained by dedicated teams.
+Additionally, the use of a consistent SQL syntax for table creation and insertion meant that the same code could be used to write to both SQLite and DuckDB, as well as any other databases supported in Python.
+The use of packages such as SQLAlchemy could be used to additionally streamline this process to any additional database by simply swapping in a new database engine.
+
+Querying the MS1 and MS2 tables was also very straightforward.
+After establishing a connection to the database, the six queries could be asked using nearly human-readable SQL syntax.
+Requesting the thousandth MS1 scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the pandas.read_sql_query function.
+More complicated queries such as retention time range (`SELECT * FROM MS1 WHERE rt BETWEEN 6 AND 8`) and a precursor mass search (`SELECT * FROM MS2 WHERE premz BETWEEN 118.086 AND 118.087`) were similarly intuitive.
+
+## Time and space requirements for a single DDA file across formats
+
+![](figures/singlefile_fig.png)
+
+## Timings for a dataset with multiple files
 
 ## Database optimization
+
 1) Indices and ordering
@@
 
+Several fundamental inefficiencies identified:
+1. Looping over each scan within a file
+2. Looping over each file within a dataset
+3. Decoding mz/int information each time if filtering on these fields
+4. RTs are not accessible outside of looping over each scan for RT subsetting
+5. Scan ids are not accessible outside of looping over each scan for scan extraction
+
 Querying things by scan number is dumb.
@@
 3) Only USEFUL when you're parsing the entire file! Nobody knows what scan number 976 corresponds to, making extraction by number useless
+
+DuckDB's zonemaps operate similarly to mzDB's bounding boxes
+
+Unlikely that databases replace the original files so this is meant to be an auxiliary data structure alongside the .raw or .mzML files, not a substitute.
+
+Databases place a larger burden on the initial builder/designer and much less on the end user.
+
+Ultimately, the design decision for mass spectrometry data format will likely result from a variety of factors, most crucially 1) initial vendor type, 2) programming language of the developer, 3) types of MS data included (MS1-only vs MS2+, metadata), 4) whether the entire file will be processed or only a subset (subsetting doesn't matter if the whole file is going to be investigated anyway),
 

### 2025-02-05 src/mzsql/duckdb_functions.py
Adding multifile option to database functions

@@ -1,42 +1,63 @@
 
 import numpy as np
 import pandas as pd
+import os
 import duckdb
 import pyteomics.mzml
 from .helpers import pmppm
 
-def turn_mzml_duckdb(file, outfile, ordered=None):
+def turn_mzml_duckdb(files, outfile, ordered=None):
     conn = duckdb.connect(outfile)
     conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS2")
+    empty_ms1 = pd.DataFrame(columns=["filename", "id", "mz", "int", "rt"]).astype(
+        {"filename": "string", "id": "int64", "mz": "float64", "int": "float64", "rt": "float64"}
+    )
+    conn.execute("CREATE TABLE MS1 AS SELECT * FROM empty_ms1")
+    empty_ms2 = pd.DataFrame(columns=["filename", "id", "premz", "fragmz", "int", "rt"]).astype(
+        {"filename": "string", "id": "int64", "premz": "float64", "fragmz": "float64", "int": "float64", "rt": "float64"}
+    )
+    conn.execute("CREATE TABLE MS2 AS SELECT * FROM empty_ms2")
 
-    MS1_dfs = []
-    MS2_dfs = []
-    for spectrum in pyteomics.mzml.MzML(file):
-        if spectrum['ms level'] == 1:
-            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-            mz_vals = spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            df_scan = pd.DataFrame({'id': idx, 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            MS1_dfs.append(df_scan)
-        if spectrum["ms level"] == 2:
-            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-            premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-            mz_vals = spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            df_scan = pd.DataFrame({'id': idx, 'premz': premz_val, 'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            MS2_dfs.append(df_scan)
+    if isinstance(files, str):
[…]

### 2025-02-05 src/mzsql/mzml_idx_functions.py

@@ -1,76 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import pyteomics.mzml
-from .helpers import pmppm
-
-
-def get_chrom_mzml_idx(idx_file, mz, ppm):
-    """
-    Retrieves chromatogram data from an indexed mzML file for a specific m/z range.
-
-    Args:
-        idx_file (str): Path to the indexed mzML file.
-        mz (float): The target m/z value.
-        ppm (float): The mass tolerance in parts per million (ppm).
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time (rt), m/z, and intensity data 
-                      within the specified m/z range.
-    """
-    raise Exception("Indexed mzML files not currently supported")
-    mzmin, mzmax = pmppm(mz, ppm)
-    scan_dfs = []
-    # file_data = file_data=pyteomics.mzml.PreIndexedMzML(idx_file).build_byte_index()
-    # with pyteomics.mzml.read(idx_file, use_index=True) as reader
-    for spectrum in file_data:
-        rt_val = spectrum['scanList']['scan'][0]['scan start time']
-        mz_vals=spectrum['m/z array']
-        int_vals = spectrum['intensity array']
-        bet_idxs = (mzmin < spectrum["m/z array"]) & (spectrum["m/z array"] < mzmax)
-        if(sum(bet_idxs)>0):
-            df_scan = pd.DataFrame({'mz':mz_vals[bet_idxs], 'int':int_vals[bet_idxs], 'rt':[rt_val]*sum(bet_idxs)})
-            scan_dfs.append(df_scan)    
-    return(pd.concat(scan_dfs, ignore_index=True))
-
-
-def get_spec_mzml_idx(idx_file, scan_num):
-    """
-    Retrieves spectrum data for a specific scan number from an indexed mzML file.
-
-    Args:
-        idx_file (str): Path to the indexed mzML file.
-        scan_num (int): The scan number.
-
-    Returns:
-        pd.DataFrame: A DataFrame containing m/z and intensity data for the specified scan number.
-    """
-    raise Exception("Indexed mzML files not currently supported")
-    # file_data=pyteomics.mzml.PreIndexedMzML(idx_file).build_byte_index()
-    # with pyteomics.mzml.read(idx_file, use_index=True) as reader:
-    return(pd.DataFrame({"mz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
-    
-def get_rtrange_mzml_idx(idx_file, rtstart, rtend):
-    """
[…]

### 2025-02-05 src/mzsql/parquet_functions.py

@@ -35,7 +35,7 @@ def turn_mzml_parquet(file, outdir, ordered=None):
     table_MS2 = pa.Table.from_pandas(all_MS2)
 
     basename = os.path.splitext(os.path.basename(file))[0]
-    os.makedirs(outdir)
+    os.makedirs(outdir, exist_ok=True)
     os.makedirs(f"{outdir}/MS1", exist_ok=True)
     os.makedirs(f"{outdir}/MS2", exist_ok=True)
     pq.write_table(table_MS1, f"{outdir}/MS1/{basename}.parquet")

### 2025-02-05 src/mzsql/sqlite_functions.py

@@ -4,49 +4,60 @@ import pyteomics.mzml
 import os
 from .helpers import pmppm
 
-def turn_mzml_sqlite(file, outfile, ordered=None):
-    """
-    Converts an mzML file into an SQLite database.
-
-    This function reads an mzML file, extracts MS1 spectra data, and stores it in an SQLite database.
-    Optionally, it can create an index on a specified column ('mz', 'int', or 'rt') for faster querying.
-
-    Parameters:
-    - file (str): Path to the input mzML file.
-    - outfile (str): Path to the output SQLite database file.
-    - ordered (str, optional): Column name to create an index on. Must be one of 'mz', 'int', or 'rt'.
-                               If None, no index is created. Default is None.
-
-    Returns:
-    - str: The path to the created SQLite database file.
-    """
-    if not os.path.exists(file):
-        raise FileNotFoundError(f"Input mzML file '{file}' does not exist.")
-
+def turn_mzml_sqlite(files, outfile, ordered=None):
     conn = sqlite3.connect(outfile)
     conn.execute("DROP TABLE IF EXISTS MS1")
     conn.execute("DROP TABLE IF EXISTS MS2")
-
-    for spectrum in pyteomics.mzml.MzML(file):
-        if spectrum['ms level'] == 1:
-            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-            mz_vals = spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            df_scan = pd.DataFrame({'id': idx, 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            df_scan.to_sql("MS1", conn, if_exists="append", index=False)
-        if spectrum['ms level'] == 2:
-            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-            mz_vals = spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-            df_scan = pd.DataFrame({'id': idx, 'premz': premz_val, 'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            df_scan.to_sql("MS2", conn, if_exists="append", index=False)
+    if isinstance(files, str):
[…]

### 2025-02-12 manuscript/scripts/singlefile_supp.py
Adding mzMLb to supp timing info

@@ -3,6 +3,7 @@ import numpy as np
 import timeit
 import pyopenms
 import pyteomics.mzml
+import pyteomics.mzmlb
 import pymzml
 import h5py
 import seaborn as sns
@@ -23,6 +24,10 @@ def get_randscan_mzml_pymzml(scan_num):
     run = pymzml.run.Reader("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", build_index_from_scratch=True)
     spec1_data = run[scan_num].peaks("raw")
     return(pd.DataFrame({"mz":spec1_data[:,0], "int":spec1_data[:,1]}))
+
+def get_randscan_mzmlb(scan_num):
+    file_data = pyteomics.mzmlb.MzMLb("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzMLb")
+    return(pd.DataFrame({"mz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
 
 def get_randscan_mza(scan_num):
     mza = h5py.File("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mza", 'r')
@@ -54,6 +59,9 @@ def min_randscan_mzml_pymzml(scan_num):
     spec1_data = pymzml_run[scan_num].peaks("raw")
     return(pd.DataFrame({"mz":spec1_data[:,0], "int":spec1_data[:,1]}))
 
+def min_randscan_mzmlb(scan_num):
+    return(pd.DataFrame({"mz":mzmlb_data[scan_num]['m/z array'], "int":mzmlb_data[scan_num]['intensity array']}))
+
 def min_randscan_mza(scan_num):
     intensities = mza_file["Arrays_intensity/"+str(scan_num)][:]
     mz = mza_file["Arrays_mz/"+str(scan_num)][:]
@@ -74,10 +82,11 @@ for scan_num in range(3, 12, 2):
     pyteo=timeit.repeat(f"get_randscan_mzml_pyteomics({scan_num})", globals=globals(), number=1, repeat=3)
     pyopen=timeit.repeat(f'get_randscan_mzml_pyopenms({scan_num})', globals=globals(), number=1, repeat=3)
     pymzml_data=timeit.repeat(f'get_randscan_mzml_pymzml({scan_num})', globals=globals(), number=1, repeat=3)
+    mzmlb=timeit.repeat(f"get_randscan_mzmlb({scan_num})", globals=globals(), number=1, repeat=3)
[…]

### 2025-02-12 manuscript/manuscript.Rmd
Adding to results section, moving figure code into main text

@@
 library(tidyverse)
+library(ggtext)
+library(patchwork)
+
+method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
+                  "mzMLb", "mzDB", "MZA", "mz5", "MZTree", "mzMD", 
+                  "SQLite", "DuckDB", "Parquet")
+
 ```
@@
 
+Details about the size, number of scans (MS1 and MS2), RT range, etc. for the files
+
 Metabolights search for
@@
 
-for %f in (*.raw) do mza.exe -file %f
+for %f in (*.mzML) do mza.exe -file %f
 ```
@@
 
+[TODO: Note the method used to identify the m/zs, scans, and RT ranges extracted here]
+
 Via %timeit% (need to redo these with randomization)
@@
 ### Universal lack of support for the six relevant queries
+
+[TODO: note the dependencies required by each of the mzML parsers (and mzapy/mz5?)]
 
@@
 Although we did not extend these databases to include the metadata associated with each file and scan, the logical framework could be easily extended in future work and the metadata typically represents a small fraction of the total space within the file, allowing us to make reasonable comparisons about file size between the databases and the metadata-rich other file types.
-We also converted each file's MS1 and MS2 table into CSV and Parquet representations for comparison using the same field/column schema.
+We also converted each file's MS1 and MS2 table into Parquet representations for comparison using the same field/column schema.
 
@@
 
+```{r singlefile_fig}
+all_timings <- read_csv("data/singlefile_times.csv") %>% 
+  mutate(method=factor(method, levels=method_names)) %>%
+  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
+                                      "ms2_spec", "premz", "fragmz"),
+                      labels=c("MS1 scan", "Chromatogram", "RT Range",
+                               "MS2 scan", "Precursor search",
+                               "Fragment search")))
+all_sizes <- read_csv("data/file_sizes.csv") %>%
+  mutate(method=factor(method, levels=method_names))
+
+time_boxplot <- all_timings %>%
+  ggplot() +
+  geom_boxplot(aes(x=method, y=time)) +
+  facet_wrap(~query) +
+  scale_y_log10() +
+  labs(x=NULL, y="Time (seconds)", color=NULL) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
+        legend.position = "none")
+size_barplot <- all_sizes %>%
+  filter(!method%in%c("pyteomics", "pyopenms", "pyopenms_2DPeak")) %>%
+  mutate(method=as.character(method)) %>%
+  mutate(method=ifelse(method=="pymzml", "mzml", method)) %>%
+  mutate(method=factor(method, levels=c("mzml", method_names))) %>%
+  mutate(query="File size") %>%
+  ggplot() +
+  geom_hline(yintercept = 0) +
+  geom_col(aes(x=method, y=file_size/1e6), color="black") +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
+  facet_wrap(~query) +
+  labs(x=NULL, y="File size (MB)", fill="File type") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), legend.position = "none")
+
+time_boxplot + size_barplot + plot_layout(widths = c(3.5, 1))
+
+ggsave("figures/singlefile_fig.png", width = 6.5, height=5)
+
+```
+
 ![](figures/singlefile_fig.png)
 
+*Figure XX: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper.
+The left six panels show boxplots representing the time required in seconds to extract an MS1 scan, an ion chromatogram, all data within a retention time range, an MS2 scan, the fragments of a specified precursor, and all precursors with a specified fragment.
+The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z.
+The right panel shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*
+
+```{r singlefile_supp}
+fun_levels = c("<span style='color:#a41118'>Including initial load</span>", 
+               "<span style='color:#0b505c'>Extraction only</span>")
+init_outside_gp <- read_csv("data/singlefile_supp.csv") %>%
+  pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
+  mutate(method=factor(method, levels=method_names)) %>%
+  mutate(fun_type=factor(fun_type, levels=c("inclusive", "minimal"),
+                         labels=fun_levels)) %>%
+  ggplot() +
+  geom_boxplot(aes(x=method, y=time, color=fun_type)) +
+  scale_y_log10() +
+  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
+  labs(x=NULL, y="Time (seconds)", color=NULL) +
+  theme_bw() +
+  theme(legend.text=element_markdown(), legend.position = "inside",
+        legend.justification.inside = c(0.98, 0.95),
+        legend.background = element_rect(color='black'),
+        plot.title.position = "plot", 
+        plot.caption.position =  "plot") +
+  ggtitle("A) Scan extraction with and without pre-loading")
+
+pyopenms_precomp_gp <- read_csv("data/pyopenms_precomp.csv") %>%
+  pivot_longer(rep1:rep3) %>%
+  mutate(type=factor(type, levels=c("init", "precomp"),
+                       labels=fun_levels)) %>%
+  mutate(method=factor(method, levels=c("spec", "chrom", "chrom_2D", "rtrange",
+                                        "rtrange_2D", "premz", "fragmz"),
+                       labels = c("Spectrum", "EIC", "EIC\n(2D Peak method)",
+                                  "RT Range", "RT Range\n(2D Peak method)",
+                                  "Precursor\nsearch", "Fragment\nsearch"))) %>%
+  ggplot() +
+  geom_boxplot(aes(x=method, y=value, color=type)) +
+  scale_y_log10() +
+  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
+  theme_bw() +
+  theme(legend.position = "none", 
+        plot.title.position = "plot", 
+        plot.caption.position =  "plot") +
+  labs(x=NULL, y="Time (seconds)", color=NULL) +
+  ggtitle("B) pyopenms metric timings with and without pre-loading")
+
+init_outside_gp / pyopenms_precomp_gp
+ggsave("figures/supp_preload_fig.png", width = 6.5, height=5)
+```
+
+*Supplemental figure XX: Timing information with and without pre-loading the object into Python prior to extraction.
+Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types.
+Panel B) shows the number of seconds required for pyopenms to perform each of the *
+
+### Spectrum extraction
+
+The simplest and most abundantly documented query was the extraction of asingle spectrum.
+In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 1, leftmost panels).
+Here, we found that the mzML and mzMLb file types were consistently the slowest to parse the mzML file type and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
+`pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX).
+It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number as scans are not always consecutive (e.g.
+during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan.
+
+The `pymzml` package was able to extract both MS1 and MS2 spectra from the mzML file nearly two orders of magnitude faster than the other mzML parsers, largely due to its use of naming the scans by their number and thus avoiding the expensive scan number extraction step.
+mzDB had the only notable difference between MS1 and MS2 scans, performing slightly better than `pyteomics` and `pyopenms` methods for MS1 data and significantly better for MS2 data, placing it approximately on par with `pymzml` in taking about a tenth of a second.
+The simple database methods (SQLite, DuckDB, and Parquet) also fell in this ~0.1 second range, with SQLite performing most poorly and DuckDB ~10x faster.
+Finally, both mz5 and MZA were an additional order of magnitude faster than any other method, returning the data within the spectrum in thousandths of a second.
+This shows the power of the HDF5 file system for data access when its location within the file is known in advance.
+
+### Chromatogram extraction and subsetting by retention time range
+
+Ion chromatogram extraction and retention time range subsets were a key metric for us, corresponding to essential tasks in chromatographic peakpicking and adduct, isotope, and in-source fragment detection (Figure 1, top middle and top right panels).
+EIC query times here were universally slower than those for a single spectrum extraction, reflecting the way in which a scan-based file type is sub-optimal for chromatogram extraction because each scan must be parsed to find data within a given *m/z* range.
+MZA and mz5 particularly suffered, with this query type entirely negating the advantages of the HDF5 file structure.
+
+MzTree and mzMD are both file types optimized exclusively for chromatogram extraction and performed very well on the EIC metric and were two orders of magnitude faster than those parsing mzMLs, with mzMD surprisingly less performant than the older MzTree file type it was based on.
+However, we also note that both Java-based applications have a slow initial file load step that must be done through a GUI and therefore could not be counted in the timing comparison, the inclusion of which would likely mitigate any advantage for a single chromatogram extraction.
+The mzDB file type is also optimized for chromatogram extraction and was an order of magnitude faster than the other existing file types for which all queries could be run (MzTree and mzMD do not provide interfaces for spectrum extraction or MS2 data).
+
+The SQLite, DuckDB, and Parquet formats were just as speedy as mzMD and MzTree with sub-second SQLite taking half a second, Parquet requiring a tenth of a second, and DuckDB reaching query times of hundredths of a second, far outstripping the seconds or even minutes typically expected of this task and resulting in a functionally instantaneous interaction for the user.
+
+Retention time range extraction times were an average of the single-spectrum extraction and the chromatogram extraction times across the board, potentially hinting at a major predictive factor in timing estimation being the total amount of scan parsing required.
+
+### MS/MS precursor and fragment search
+
+### File sizes
+
 ## Timings for a dataset with multiple files
+
+The single-file, single-metric case discussed above and shown in Figure XX is largely a worst-case scenario for many MS data systems that have a slow initial setup step that makes downstream analysis faster.
+To compare these systems more fairly to our database schema, we also tested timings across multiple files and queries of more than a single metric.
 

### 2025-02-12 manuscript/scripts/figscript.R

@@ -1,45 +1,0 @@
-
-library(tidyverse)
-
-method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
-                  "mzMLb", "mzDB", "MZA", "mz5", "MZTree", "mzMD", 
-                  "SQLite", "DuckDB", "Parquet")
-
-all_timings <- read_csv("data/singlefile_times.csv") %>% 
-  mutate(method=factor(method, levels=method_names)) %>%
-  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
-                                      "ms2_spec", "premz", "fragmz"),
-                      labels=c("MS1 scan", "Chromatogram", "RT Range",
-                               "MS2 scan", "Precursor search",
-                               "Fragment search")))
-all_sizes <- read_csv("data/file_sizes.csv") %>%
-  mutate(method=factor(method, levels=method_names))
-
-time_boxplot <- all_timings %>%
-  ggplot() +
-  geom_boxplot(aes(x=method, y=time, color=method)) +
-  facet_wrap(~query) +
-  scale_y_log10() +
-  labs(x=NULL, y="Time (seconds)", color=NULL) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
-        legend.position = "none")
-size_barplot <- all_sizes %>%
-  filter(!method%in%c("pyteomics", "pyopenms", "pyopenms_2DPeak")) %>%
-  mutate(method=as.character(method)) %>%
-  mutate(method=ifelse(method=="pymzml", "mzml", method)) %>%
-  mutate(method=factor(method, levels=c("mzml", method_names))) %>%
-  mutate(query="File size") %>%
-  ggplot() +
-  geom_hline(yintercept = 0) +
-  geom_col(aes(x=method, y=file_size/1e6, fill=method), color="black") +
-  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
-  facet_wrap(~query) +
-  labs(x=NULL, y="File size (MB)", fill="File type") +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), legend.position = "none")
-library(patchwork)
-time_boxplot + size_barplot + plot_layout(widths = c(3.5, 1))
-
-ggsave("figures/singlefile_fig.png", width = 6.5, height=5)
-

### 2025-02-13 src/mzsql/mzml_functions.py
Switching pyopenms back to singlefile things

@@ -112,7 +112,7 @@ def get_MS2nloss_mzml_pyteomics(mzml_file, neutral_loss, ppm):
 
 
 # pyopenms things
-def get_chrom_mzml_pyopenms(files, mz, ppm):
+def get_chrom_mzml_pyopenms(file, mz, ppm):
     """
     Retrieves chromatogram data from a mzML file using pyopenms for a specific m/z range.
 
@@ -126,10 +126,7 @@ def get_chrom_mzml_pyopenms(files, mz, ppm):
                       within the specified m/z range.
     """
     exp = pyopenms.MSExperiment()
-    if isinstance(files, str):
-        files = [files]
-    for file in files:
-        pyopenms.MzMLFile().load(file, exp)
+    pyopenms.MzMLFile().load(file, exp)
     mzmin, mzmax = pmppm(mz, ppm)
     scan_dfs = []
     for spectrum in exp:
@@ -142,7 +139,7 @@ def get_chrom_mzml_pyopenms(files, mz, ppm):
                 scan_dfs.append(df_scan)
     return(pd.concat(scan_dfs, ignore_index=True))
     
-def get_chrom_mzml_pyopenms_2DPeak(files, mz, ppm):
+def get_chrom_mzml_pyopenms_2DPeak(file, mz, ppm):
     """
     Retrieves 2D chromatogram data from a mzML file using pyopenms for a specific m/z range.
 
@@ -156,22 +153,15 @@ def get_chrom_mzml_pyopenms_2DPeak(files, mz, ppm):
                       within the specified m/z range in a 2D peak format.
     """
     exp = pyopenms.MSExperiment()
-    if isinstance(files, str):
-        files = [files]
-    for file in files:
-        pyopenms.MzMLFile().load(file, exp)
+    pyopenms.MzMLFile().load(file, exp)
     exp.updateRanges()
     mzmin, mzmax = pmppm(mz, ppm)
     chrom_data=exp.get2DPeakDataLong(min_mz=mzmin, max_mz=mzmax, min_rt=exp.getMinRT(), max_rt=exp.getMaxRT())
     return(pd.DataFrame({"rt":chrom_data[0], "mz":chrom_data[1], "int":chrom_data[2]}))
     
-def get_spec_mzml_pyopenms(files, scan_num):
+def get_spec_mzml_pyopenms(file, scan_num):
     exp = pyopenms.MSExperiment()
-    if isinstance(files, str):
-        files = [files]
-    for file in files:
[…]

### 2025-02-17 manuscript/scripts/multichrom_timing.py
Adding multichrom timings things

@@ -8,12 +8,12 @@ import os
 import random
 
 # Create databases from mzML files if they don't already exist
-if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite"))):
-    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
-    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.duckdb")
-    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds")
+if(not(os.path.exists("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite"))):
+    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
+    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.duckdb")
+    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04_pqds")
 
-conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
+conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
 top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
 top_masses = []
 top_rts = []
@@ -195,18 +195,21 @@ def get_multichrom_parquet_loop(pqds_dir, mzs, ppm):
     return pd.concat(scan_data, ignore_index=True)
 
 function_list = [
-#    ("pyteomics", "mzml_pyteomics", ".mzML"),
[…]

### 2025-02-17 manuscript/scripts/dbloopjoinuni.py
Adding dbloopjoin things

@@ -1,0 +1,170 @@
+
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+from mzsql import *
+import sqlite3
+import os
+import random
+
+random.seed(123)
+files = ["E:/mzsql/MTBLS10066/"+f for f in os.listdir("E:/mzsql/MTBLS10066/") if f.endswith(".mzML")]
+msfiles = random.sample(files, 10)
+
+if(not(os.path.exists("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite"))):
+    turn_mzml_duckdb(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.duckdb")
+    turn_mzml_sqlite(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
+    turn_mzml_parquet(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.parquet")
+
+conn = sqlite3.connect("E:/mzsql/MTBLS10066/multifile.sqlite")
+
+top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 50000", conn)
+top_masses = []
+top_rts = []
+for i in range(10):
+    print(i)
+    top_rts.append(top_int_df["rt"].iloc[0])
+    top_mz = top_int_df["mz"].iloc[0]
+    top_masses.append(top_mz)
+    mzmin, mzmax = pmppm(top_mz, 10)
+    rtmin, rtmax = top_int_df["rt"].iloc[0]+(-0.5, 0.5)
+    top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax)) & ((top_int_df["rt"]<rtmin) | (top_int_df["rt"]>rtmax))]
+
+print(top_masses)
+
+
+
+def get_multichrom_duckdb_loop(file, mzs, ppm):
+    conn = duckdb.connect(file)
+    all_chroms = []
+    for mz_i in mzs:
+        query = "SELECT mz, int, rt FROM MS1 WHERE mz BETWEEN ? AND ?"
+        query_data = conn.execute(query, pmppm(mz_i, ppm)).fetchdf()
+        query_data["mz_i"] = mz_i
+        all_chroms.append(query_data)
+    conn.close()
[…]

### 2025-02-17 manuscript/scripts/dbloopjoinuni.py
Adding multi-chrom things to dbloopjoinuni

@@ -17,23 +17,16 @@ if(not(os.path.exists("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite"))):
     turn_mzml_sqlite(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
     turn_mzml_parquet(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.parquet")
 
-conn = sqlite3.connect("E:/mzsql/MTBLS10066/multifile.sqlite")
-
-top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 50000", conn)
+conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
+top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
 top_masses = []
 top_rts = []
-for i in range(10):
-    print(i)
+for i in range(100):
     top_rts.append(top_int_df["rt"].iloc[0])
     top_mz = top_int_df["mz"].iloc[0]
     top_masses.append(top_mz)
     mzmin, mzmax = pmppm(top_mz, 10)
-    rtmin, rtmax = top_int_df["rt"].iloc[0]+(-0.5, 0.5)
-    top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax)) & ((top_int_df["rt"]<rtmin) | (top_int_df["rt"]>rtmax))]
-
-print(top_masses)
-
-
+    top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax))]
 
 def get_multichrom_duckdb_loop(file, mzs, ppm):
     conn = duckdb.connect(file)
@@ -161,10 +154,13 @@ function_list = [
 init_df = pd.DataFrame(columns=["method", "time"])
 init_df.to_csv('data/dbloopjoinuni_times.csv', index=False)
 
-for fun_i, file_ending in function_list:
-    print(fun_i)
-    rep_function = f"{fun_i}('E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04{file_ending}', {top_masses}, 5)"
-    time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=3)
-    fun_name = fun_i.replace("get_multichrom_", "")
[…]

### 2025-02-18 src/mzsql/duckdb_functions.py
Bugfix ordering for duckdb and parquet

@@ -52,12 +52,20 @@ def turn_mzml_duckdb(files, outfile, ordered=None):
 
     if ordered is not None:
         if ordered == "rt":
-            conn.execute(f"CREATE TABLE MS1 AS SELECT * FROM MS1 ORDER BY {ordered}")
-            conn.execute(f"CREATE TABLE MS2 AS SELECT * FROM MS2 ORDER BY {ordered}")
+            conn.execute(f"CREATE TABLE MS1_ord AS SELECT * FROM MS1 ORDER BY {ordered}")
+            conn.execute(f"DROP TABLE MS1")
+            conn.execute(f"ALTER TABLE MS1_ord RENAME TO MS1")
+            conn.execute(f"CREATE TABLE MS2_ord AS SELECT * FROM MS2 ORDER BY {ordered}")
+            conn.execute(f"DROP TABLE MS2")
+            conn.execute(f"ALTER TABLE MS2_ord RENAME TO MS2")
         if ordered == "mz":
-            conn.execute(f"CREATE TABLE MS1 AS SELECT * FROM MS1 ORDER BY {ordered}")
+            conn.execute(f"CREATE TABLE MS1_ord AS SELECT * FROM MS1 ORDER BY {ordered}")
+            conn.execute(f"DROP TABLE MS1")
+            conn.execute(f"ALTER TABLE MS1_ord RENAME TO MS1")
         if ordered in ["fragmz", "premz"]:
-            conn.execute(f"CREATE TABLE MS2 AS SELECT * FROM MS2 ORDER BY {ordered}")
+            conn.execute(f"CREATE TABLE MS2_ord AS SELECT * FROM MS2 ORDER BY {ordered}")
+            conn.execute(f"DROP TABLE MS2")
+            conn.execute(f"ALTER TABLE MS2_ord RENAME TO MS2")
     conn.close()
 
     return outfile

### 2025-02-18 src/mzsql/parquet_functions.py

@@ -28,8 +28,15 @@ def turn_mzml_parquet(file, outdir, ordered=None):
     all_MS1 = pd.concat(MS1_dfs, ignore_index=True)
     all_MS2 = pd.concat(MS2_dfs, ignore_index=True)
     if ordered is not None:
-        all_MS1.sort_values(by=ordered, inplace=True)
-        all_MS2.sort_values(by=ordered, inplace=True)
+        if ordered == "rt":
+            all_MS1.sort_values(by=ordered, inplace=True)
+            all_MS2.sort_values(by=ordered, inplace=True)
+        if ordered == "mz":
+            all_MS1.sort_values(by=ordered, inplace=True)
+        if ordered == "fragmz":
+            all_MS2.sort_values(by=ordered, inplace=True)
+        if ordered == "premz":
+            all_MS2.sort_values(by=ordered, inplace=True)
 
     table_MS1 = pa.Table.from_pandas(all_MS1)
     table_MS2 = pa.Table.from_pandas(all_MS2)

### 2025-02-18 manuscript/scripts/idx_ord_timing.py
Adding indexed and ordered timing things

@@ -1,0 +1,45 @@
+
+import numpy as np
+import pandas as pd
+import sqlite3
+import os
+from mzsql import *
+
+# Create databases from mzML files if they don't already exist
+if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite"))):
+    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
+    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.duckdb")
+    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds")
+
+if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_ord.sqlite"))):
+    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_ord.sqlite", ordered="mz")
+    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_ord.duckdb", ordered="mz")
+    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds_ord", ordered="mz")
+
+conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
+top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
+top_masses = []
+top_rts = []
+for i in range(10):
+    top_rts.append(top_int_df["rt"].iloc[0])
+    top_mz = top_int_df["mz"].iloc[0]
+    top_masses.append(top_mz)
+    mzmin, mzmax = pmppm(top_mz, 10)
[…]

### 2025-02-18 manuscript/scripts/idx_ord_timing.py
Adding index size information

@@ -43,3 +43,17 @@ for target_mz in top_masses:
         time_taken = time_chrom(suffix, filetype, target_mz, 10, verbose=True)
         time_df = pd.DataFrame({"query":["chrom"], "target":[target_mz], "method":[name], "time":time_taken})
         time_df.to_csv('data/idx_ord_times.csv', mode='a', header=False, index=False)
+
+file_sizes = {}
+for name, suffix, filetype in function_list:
+    if(name == "Parquet"):
+        dirsize = sum(
+            os.path.getsize(os.path.join(root, f)) 
+            for root, _, files in os.walk("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds") 
+            for f in files
+        )
+        file_sizes[name] = dirsize
+    else:
+        file_sizes[name] = os.path.getsize(f"E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01{filetype}")
+
+pd.DataFrame(file_sizes.items(), columns=["method", "file_size"]).to_csv("data/idx_sizes.csv", index=False)

### 2025-02-18 manuscript/scripts/idx_ord_timing.py
Rerunning timings with correct pqds

@@ -46,7 +46,7 @@ for target_mz in top_masses:
 
 file_sizes = {}
 for name, suffix, filetype in function_list:
-    if(name == "Parquet"):
+    if(suffix == "parquet"):
         dirsize = sum(
             os.path.getsize(os.path.join(root, f)) 
             for root, _, files in os.walk("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds") 

### 2025-02-18 manuscript/scripts/idx_ord_timing.py
Combining file sizes w timing data

@@ -37,6 +37,9 @@ function_list = [
     ("Parquet (ordered)", "parquet", "_pqds_ord")
 ]
 
+init_df = pd.DataFrame(columns=["query", "target", "method","time"])
+init_df.to_csv('data/idx_ord_times.csv', index=False)
+
 for target_mz in top_masses:
     print(target_mz)
     for name, suffix, filetype in function_list:
@@ -56,4 +59,6 @@ for name, suffix, filetype in function_list:
     else:
         file_sizes[name] = os.path.getsize(f"E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01{filetype}")
 
-pd.DataFrame(file_sizes.items(), columns=["method", "file_size"]).to_csv("data/idx_sizes.csv", index=False)
+
+time_data = pd.DataFrame(file_sizes.items(), columns=["method", "file_size"])
+pd.read_csv('data/idx_ord_times.csv').merge(time_data, on="method", how="left").to_csv('data/idx_ord_times.csv')

### 2025-02-18 manuscript/manuscript.Rmd
Adding figures and text to manuscript

@@
 method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
-                  "mzMLb", "mzDB", "MZA", "mz5", "MZTree", "mzMD", 
+                  "mzMLb", "mzDB", "MZA", "mz5", 
+                  "MZTree", "mzMD", 
                   "SQLite", "DuckDB", "Parquet")
+method_colors <- c("#0B4F5B", "#147C8F", "#21A8C0", "#44C3DA", 
+                   "#028e34", "#f4bb23", "#e0670b", "#a41118", 
+                   "#E25C76", "#E891A2", 
+                   "#580770", "#8A1BAC", "#B04ACF")
 
@@
 
+## Hardware and software versioning
+
 # Results
@@
   ggplot() +
-  geom_boxplot(aes(x=method, y=time)) +
+  geom_boxplot(aes(x=method, y=time, color=method)) +
   facet_wrap(~query) +
   scale_y_log10() +
+  scale_color_manual(breaks = method_names, values=method_colors) +
   labs(x=NULL, y="Time (seconds)", color=NULL) +
@@
   mutate(method=as.character(method)) %>%
-  mutate(method=ifelse(method=="pymzml", "mzml", method)) %>%
-  mutate(method=factor(method, levels=c("mzml", method_names))) %>%
+  mutate(method=ifelse(method=="pymzml", "mzML", method)) %>%
+  mutate(method=factor(method, levels=c("mzML", method_names))) %>%
   mutate(query="File size") %>%
@@
   geom_hline(yintercept = 0) +
-  geom_col(aes(x=method, y=file_size/1e6), color="black") +
+  geom_col(aes(x=method, y=file_size/1e6, fill=method), color="black") +
   scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
+  scale_fill_manual(breaks = c("mzML", method_names), 
+                    values=c("#147C8F", method_colors)) +
   facet_wrap(~query) +
@@
 
+![](figures/supp_preload_fig.png)
+
 *Supplemental figure XX: Timing information with and without pre-loading the object into Python prior to extraction.
 Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types.
-Panel B) shows the number of seconds required for pyopenms to perform each of the *
+Panel B) shows the number of seconds required for pyopenms to perform each of the queries if the initial data load was removed from the timing count.
+Error bars show 3x replicate queries of 5x different targets.*
 
@@
 
-The SQLite, DuckDB, and Parquet formats were just as speedy as mzMD and MzTree with sub-second SQLite taking half a second, Parquet requiring a tenth of a second, and DuckDB reaching query times of hundredths of a second, far outstripping the seconds or even minutes typically expected of this task and resulting in a functionally instantaneous interaction for the user.
+The SQLite, DuckDB, and Parquet formats were just as speedy as mzMD and MzTree with SQLite taking half a second, Parquet requiring a tenth of a second, and DuckDB reaching query times of hundredths of a second, far outstripping the seconds or even minutes typically expected of this task and resulting in a functionally instantaneous interaction for the user.
 
-Retention time range extraction times were an average of the single-spectrum extraction and the chromatogram extraction times across the board, potentially hinting at a major predictive factor in timing estimation being the total amount of scan parsing required.
+Retention time range extraction times were an average of the single-spectrum extraction and the chromatogram extraction times across the board.
+potentially hinting at a major predictive factor in timing estimation being the total amount of scan parsing required.
 
@@
 
+We also investigated the efficacy of the various MS data formats for MS/MS data and found that support for fragmentation data searches was lacking or absent from the documentation and exposed functionality of each of these file types, requiring custom implementations every time.
+Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure XX, rightmost panels).
+
+All existing MS data types required multiple seconds to perform a single fragment search (Figure XX, bottom-right panel), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor.
+The SQL-based parsers, on the other hand, all took fractions of a second and consistently returned the relevant data hundreds of times more quickly than existing methods.
+The same was true for a precursor search across all methods aside from mzDB (Figure XX, bottom middle panel), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.
+
 ### File sizes
 
-## Timings for a dataset with multiple files
+File size is another important constraint on the efficacy of various MS formats.
+We measured the size on disk of each of the file types and found that they varied by approximately an order of magnitude, with HDF-based file types hovering around one-third the size of the mzML (83 megabytes, MB) while mzDB and mzMD were larger (108 MB and 96 MB, respectively).
+The SQLite object was the largest on disk of all the file types, nearly doubling the mzML's size at 145 MB, while DuckDB was on par with MZA at two-thirds of the mzML (54 MB) and Parquet improved slightly upon that again with its columnar-based storage format (Figure XX).
 
-The single-file, single-metric case discussed above and shown in Figure XX is largely a worst-case scenario for many MS data systems that have a slow initial setup step that makes downstream analysis faster.
-To compare these systems more fairly to our database schema, we also tested timings across multiple files and queries of more than a single metric.
+However, these comparisons are not perfect because not all files store exactly the same data.
+MZTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS<sup>1</sup> scans was 64 MB, a 23% size reduction.
+The SQLite, DuckDB, and Parquet formats also lack the scan and file metadata that's present in the other file types, though it is difficult to estimate the fraction of disk space allocated for this (and which will depend upon the precise definition of metadata).
 
-## Database optimization
+## Timings for multiple chromatograms
 
-1) Indices and ordering
-2) Peak data extraction via joins vs loops vs unified query
+The single-file, single-metric case discussed above and shown in Figure XX is largely a worst-case scenario for many MS data systems that have a slow initial setup step to make downstream analysis faster.
+To compare these systems more fairly to our database schema, we also tested timings across multiple chromatograms.
+In each case, this was implemented as a for loop iterating over an increasing number of chromatograms corresponding to the largest intensity ions in the file (Figure XX).
+
+```{r}
+multichrom_times <- read_csv("data/multichrom_times.csv")
+timelabs <- c("0.01 seconds", "One second", "One minute", "One hour")
+multichrom_times %>%
+  mutate(method=factor(method, levels=method_names)) %>%
+  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
+  geom_abline(slope=1, intercept=-3:3, color="black") +
+  geom_smooth(method="lm", formula="y~x", alpha=0.2, fill="grey50") +
+  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
+  scale_y_log10(breaks=c(0.01, 1, 60, 3600), labels=timelabs) +
+  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
+  scale_fill_manual(breaks = method_names, values=method_colors, 
+                    aesthetics = c("fill", "color")) +
+  labs(x="Number of chromatograms extracted", y="Time required",
+       color=NULL, fill=NULL) +
+  theme_bw() +
+  theme(panel.grid.minor = element_blank())
+ggsave("figures/multichrom_fig.png", width = 6.5, height=5)
+```
+
+![](figures/multichrom_fig.png)
+
+*Figure XX: Scatter plot of the time required to extract multiple chromatograms using various methods on logarithmic axes.
+Best-fit linear models have been added for each method are shown behind triplicate timing measurements.
+Transparent intervals around each best-fit line show a single standard error of the mean.
+Chromatograms correspond to the largest intensity ions in the file.
+1:1 lines have been added in black behind the data for comparison.*
+
+Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure XX).
+pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
+Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).
+
+```{r}
+multichrom_times <- read_csv("data/multichrom_times.csv")
+multichrom_baseline_df <- multichrom_times %>%
+  filter(n_chroms==1) %>%
+  summarise(time=mean(time), .by=method) %>%
+  mutate(method=factor(method, levels=method_names))
+multichrom_times %>%
+  mutate(method=factor(method, levels=method_names)) %>%
+  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
+  geom_abline(aes(slope=1, intercept=log10(time)), data=multichrom_baseline_df) +
+  geom_smooth(method="lm", formula="y~x", color="black") +
+  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
+  scale_y_log10(expand=expansion(0.1)) +
+  scale_x_log10(breaks=unique(multichrom_times$n_chroms), expand=expansion(0.1)) +
+  scale_fill_manual(breaks = method_names, values=method_colors, 
+                    aesthetics = c("fill", "color")) +
+  facet_wrap(~method, scales="free_y") +
+  labs(x="Number of chromatograms extracted", y="Time required (seconds)") +
+  theme_bw() +
+  theme(legend.position = "none")
+ggsave("figures/supp_multichrom_facets.png", width = 6.5, height=5)
+```
+
+![](figures/supp_multichrom_facets.png)
+
+*Supplemental figure XX: Data from main text Figure XX with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
+Best-fit linear models have been added for each method are shown behind triplicate timing measurements.
+Transparent intervals around each best-fit line show a single standard error of the mean.*
+
+We also explored whether database queries could be improved via the use of either a unified query (single SQL statement with multiple OR clauses for each ion's *m/z* range) or a non-equi join between a peak table with *m/z* minimum and maximum columns (Supplemental Figure XX).
+SQLite and Parquet performed ~3-5 times faster with the unified query than with the loop method despite the necessity of and additional processing step for the looped query to correctly assign each data point to its original peak information.
+The opposite was true for DuckDB likely due to its optimized reader, with the unified query consistently outperformed by the non-equi join when 100 chromatograms were extracted.
+
+```{r}
+type_labels = c("<span style='color:#a41118'>For loop</span>", 
+               "<span style='color:#f4bb23'>Unified query</span>",
+               "<span style='color:#147C8F'>Non-equi join</span>")
+read_csv("data/dbloopjoinuni_times.csv") %>%
+  separate(method, into = c("method", "type")) %>%
+  mutate(type=case_when(
+    type=="join"~"Non-equi join",
+    type=="loop"~"For loop",
+    type=="unified"~"Unified query"
+  )) %>%
+  mutate(type=factor(type, levels=c("For loop", "Unified query", "Non-equi join"),
+                     labels=type_labels)) %>%
+  mutate(method=factor(method, levels=c("sqlite", "duckdb", "parquet"),
+                       labels=c("SQLite", "DuckDB", "Parquet"))) %>%
+  ggplot(aes(x=n_chroms, y=time, fill=type)) +
+  geom_point(position=position_dodge(width = 0.1), color="black", size=3, pch=21) +
+  facet_wrap(~method, nrow=1) +
+  scale_y_log10() +
+  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
+  scale_fill_manual(breaks=type_labels, values = c("#a41118", "#f4bb23", "#147C8F")) +
+  labs(x="Number of chromatograms", y="Time required (seconds)", fill=NULL) +
+  theme_bw() +
+  theme(legend.position = "inside", legend.justification.inside = c(0.5, 0.95),
+        legend.background = element_rect(color='black'),
+        legend.text=element_markdown(),
+        plot.title.position = "plot", 
+        plot.caption.position =  "plot")
+ggsave("figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
+```
+
+![](figures/supp_dbloopjoinuni.png)
+
+*Supplemental figure XX: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color.
+Ten replicates are shown for each data point.*
+
+## Database optimization via indices/ordering and multi-file constructions
+
+Databases also provide multiple ways to optimize queries.
+
+```{r}
+read_csv("data/idx_ord_times.csv") %>%
+  ggplot() +
+  geom_boxplot(aes(x=file_size/1e6, y=time, color=method)) +
+  scale_y_log10()
+```
 
@@
 
-While the gap between data scientist and mass spectrometrist continues to narrow, current implementations of MS data storage systems almost seem designed to *increase* it.
-This is accomplished by extremely poor documentation that makes it difficult to do anything beyond the original designer's intent, typically with data structures that are not familiar to the MS expert.
+As the gap between data scientist and mass spectrometrist continues to narrow, mass spectrometry (MS) data formats should facilitate this convergence.
+
+Large gaps etc.
+accomplished by extremely poor documentation that makes it difficult to do anything beyond the original designer's intent, typically with data structures that are not familiar to the MS expert.
 
@@
 
-Unlikely that databases replace the original files so this is meant to be an auxiliary data structure alongside the .raw or .mzML files, not a substitute.
+Unlikely that databases replace the original files so this is meant to be an auxiliary data structure **alongside** the .raw or .mzML files, not a substitute.
 
-Databases place a larger burden on the initial builder/designer and much less on the end user.
+Databases place a larger burden on the initial builder/designer and much less on the end user
+
+Better compression algorithms and fancier data structures can easily improve upon the text-based XML in both size on disk and speed of access but also tend to accumulate significant technical debt.
+Alternatives must therefore compete not only in computational efficacy but also in simplicity to allow widespread adoption, a feat thus far unachieved.
 

### 2025-02-19 src/mzsql/duckdb_functions.py
Fixing size bug in duckdb, now sorting earlier

@@ -47,25 +47,16 @@ def turn_mzml_duckdb(files, outfile, ordered=None):
     
         all_MS1 = pd.concat(MS1_dfs, ignore_index=True)
         all_MS2 = pd.concat(MS2_dfs, ignore_index=True)
+        if ordered is not None:
+            if ordered == "rt":
+                all_MS1.sort_values("rt", inplace=True)
+                all_MS2.sort_values("rt", inplace=True)
+            if ordered == "mz":
+                all_MS1.sort_values("mz", inplace=True)
+            if ordered in ["fragmz", "premz"]:
+                all_MS1.sort_values(ordered, inplace=True)
         conn.execute("INSERT INTO MS1 SELECT * FROM all_MS1")
         conn.execute("INSERT INTO MS2 SELECT * FROM all_MS2")
-
-    if ordered is not None:
-        if ordered == "rt":
-            conn.execute(f"CREATE TABLE MS1_ord AS SELECT * FROM MS1 ORDER BY {ordered}")
-            conn.execute(f"DROP TABLE MS1")
-            conn.execute(f"ALTER TABLE MS1_ord RENAME TO MS1")
-            conn.execute(f"CREATE TABLE MS2_ord AS SELECT * FROM MS2 ORDER BY {ordered}")
-            conn.execute(f"DROP TABLE MS2")
-            conn.execute(f"ALTER TABLE MS2_ord RENAME TO MS2")
-        if ordered == "mz":
-            conn.execute(f"CREATE TABLE MS1_ord AS SELECT * FROM MS1 ORDER BY {ordered}")
-            conn.execute(f"DROP TABLE MS1")
-            conn.execute(f"ALTER TABLE MS1_ord RENAME TO MS1")
-        if ordered in ["fragmz", "premz"]:
-            conn.execute(f"CREATE TABLE MS2_ord AS SELECT * FROM MS2 ORDER BY {ordered}")
-            conn.execute(f"DROP TABLE MS2")
-            conn.execute(f"ALTER TABLE MS2_ord RENAME TO MS2")
     conn.close()
 
     return outfile

### 2025-02-19 manuscript/scripts/idx_ord_timing.py
Adding new idx_ord timings

@@ -61,4 +61,4 @@ for name, suffix, filetype in function_list:
 
 
 time_data = pd.DataFrame(file_sizes.items(), columns=["method", "file_size"])
-pd.read_csv('data/idx_ord_times.csv').merge(time_data, on="method", how="left").to_csv('data/idx_ord_times.csv')
+pd.read_csv('data/idx_ord_times.csv').merge(time_data, on="method", how="left").to_csv('data/idx_ord_times.csv', index=False)

### 2025-02-20 manuscript/scripts/multifile_timing.py
Writing multi-file timing things db comparison

@@ -1,0 +1,78 @@
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import sqlite3
+import os
+import glob
+import random
+from mzsql import *
+
+random.seed(123)
+multifile_subset = random.sample(glob.glob("E:/mzsql/MTBLS10066/202*.mzML"), 100)
+multifile_subset = [f.replace("\\", "/") for f in multifile_subset]
+
+# Create individual sqlite, duckdb, and parquet files
+# for file_i in multifile_subset:
+#     turn_mzml_sqlite(file_i, outfile=file_i.replace(".mzML", ".sqlite"), ordered="mz")
+#     turn_mzml_duckdb(file_i, outfile=file_i.replace(".mzML", ".duckdb"), ordered="mz")
+#     turn_mzml_parquet(file_i, outdir=file_i.replace(".mzML", "_pqds"), ordered="mz")
+
+# Create multifile dbs
+# turn_mzml_sqlite(multifile_subset[0], outfile="E:/mzsql/MTBLS10066/multifile_1.sqlite", ordered="mz")
+# turn_mzml_duckdb(multifile_subset[0], outfile="E:/mzsql/MTBLS10066/multifile_1.duckdb", ordered="mz")
+# turn_mzml_parquet(multifile_subset[0], outdir="E:/mzsql/MTBLS10066/multifile_1_pqds", ordered="mz")
+# turn_mzml_sqlite(multifile_subset[:10], outfile="E:/mzsql/MTBLS10066/multifile_10.sqlite", ordered="mz")
+# turn_mzml_duckdb(multifile_subset[:10], outfile="E:/mzsql/MTBLS10066/multifile_10.duckdb", ordered="mz")
+# turn_mzml_parquet(multifile_subset[:10], outdir="E:/mzsql/MTBLS10066/multifile_10_pqds", ordered="mz")
+# turn_mzml_sqlite(multifile_subset, outfile="E:/mzsql/MTBLS10066/multifile_100.sqlite", ordered="mz")
+# turn_mzml_duckdb(multifile_subset, outfile="E:/mzsql/MTBLS10066/multifile_100.duckdb", ordered="mz")
+# turn_mzml_parquet(multifile_subset, outdir="E:/mzsql/MTBLS10066/multifile_100_pqds", ordered="mz")
+
+# Get the largest masses to query
+conn = sqlite3.connect("E:/mzsql/MTBLS10066/multifile_1.sqlite")
[…]

### 2025-02-20 manuscript/scripts/singlefile_timing.py
Adding MS1 vs MS1+MS2 size details

@@ -138,3 +138,7 @@ for name, suffix, filetype in function_list:
 pd.DataFrame(file_sizes.items(), columns=["method", "file_size"]).to_csv("data/file_sizes.csv", index=False)
 
 print("Success!")
+
+# msconvert .\20220921_LEAP-POS_BL01.mzML --filter "msLevel 1" -o "ms1_only"
+# os.path.getsize("ms1_only/20220921_LEAP-POS_BL01.mzML")/os.path.getsize("20220921_LEAP-POS_BL01.mzML")
+

### 2025-02-20 src/mzsql/duckdb_functions.py
Adding filename to duckdb output

@@ -64,7 +64,7 @@ def turn_mzml_duckdb(files, outfile, ordered=None):
 def get_chrom_duckdb(file, mz, ppm):
     conn = duckdb.connect(file)
     mz_min, mz_max = pmppm(mz, ppm)
-    query = "SELECT mz, int, rt FROM MS1 WHERE mz BETWEEN ? AND ?"
+    query = "SELECT filename, mz, int, rt FROM MS1 WHERE mz BETWEEN ? AND ?"
     query_data = conn.execute(query, (mz_min, mz_max)).fetchdf()    
     conn.close()
     return query_data

### 2025-02-20 src/mzsql/parquet_functions.py
Adding multi-file functionality to pqds

@@ -9,44 +9,47 @@ import pyarrow.compute as pc
 import pyarrow.dataset as ds
 import os
 
-def turn_mzml_parquet(file, outdir, ordered=None):
+def turn_mzml_parquet(files, outdir, ordered=None):
     MS1_dfs = []
     MS2_dfs = []
-    for spectrum in pyteomics.mzml.MzML(file):
-        idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-        mz_vals = spectrum['m/z array']
-        int_vals = spectrum['intensity array']
-        rt_val = spectrum['scanList']['scan'][0]['scan start time']
-        if spectrum['ms level'] == 1:
-            df_scan = pd.DataFrame({'id': idx, 'mslevel':"MS1", 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            MS1_dfs.append(df_scan)
-        if spectrum["ms level"] == 2:
-            premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-            df_scan = pd.DataFrame({'id': idx, 'mslevel':"MS2", 'premz': premz_val, 'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-            MS2_dfs.append(df_scan)
-
-    all_MS1 = pd.concat(MS1_dfs, ignore_index=True)
-    all_MS2 = pd.concat(MS2_dfs, ignore_index=True)
-    if ordered is not None:
-        if ordered == "rt":
-            all_MS1.sort_values(by=ordered, inplace=True)
-            all_MS2.sort_values(by=ordered, inplace=True)
-        if ordered == "mz":
-            all_MS1.sort_values(by=ordered, inplace=True)
-        if ordered == "fragmz":
-            all_MS2.sort_values(by=ordered, inplace=True)
-        if ordered == "premz":
-            all_MS2.sort_values(by=ordered, inplace=True)
-
-    table_MS1 = pa.Table.from_pandas(all_MS1)
-    table_MS2 = pa.Table.from_pandas(all_MS2)
-
-    basename = os.path.splitext(os.path.basename(file))[0]
-    os.makedirs(outdir, exist_ok=True)
-    os.makedirs(f"{outdir}/MS1", exist_ok=True)
-    os.makedirs(f"{outdir}/MS2", exist_ok=True)
-    pq.write_table(table_MS1, f"{outdir}/MS1/{basename}.parquet")
[…]

### 2025-02-20 manuscript/scripts/multifile_timing.py
Patched csv write bug and got multifile data

@@ -75,4 +75,4 @@ for mz_i in top_masses:
 
 singlefile_df = pd.concat(singlefile_timings, ignore_index=True)
 multifile_df = pd.concat(multifile_timings, ignore_index=True)
-pd.concat(singlefile_df, multifile_df, ignore_index=True).to_csv('data/multifile_times.csv', mode='w', index=False)
+pd.concat([singlefile_df, multifile_df], ignore_index=True).to_csv('data/multifile_times.csv', mode='w', index=False)

### 2025-02-20 manuscript/scripts/multifile_timing.py
Adding new timing data that actually measures multiple files

@@ -48,15 +48,22 @@ for mz_i in top_masses:
     for n_files in [1, 10, 100]:
         print(n_files)
         this_multifile_subset = multifile_subset[:n_files]
+        total_time = {"duckdb":0, "sqlite":0, "parquet":0}
         for mzml_file in this_multifile_subset:
             for db_name, db_ending in [("duckdb", ".duckdb"), ("sqlite", ".sqlite"), ("parquet", "_pqds")]:
                 db_path = mzml_file.replace(".mzML", db_ending)
                 rep_function = f"get_chrom_{db_name}('{db_path}', {mz_i}, 5)"
-                time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=3)
-                time_data = pd.DataFrame({"method":db_name, "mz_target":mz_i, "n_files":n_files, "type":"singlefile",
-                                          "mzml_file":os.path.basename(mzml_file), 
-                                          "time":time_vals})
-                singlefile_timings.append(time_data)
+                time_val = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
+                total_time[db_name] += time_val[0]
+        for db_name, time_val in total_time.items():
+            time_data = pd.DataFrame([{
+                "method": db_name,
+                "mz_target": mz_i,
+                "n_files": n_files,
+                "type": "singlefile",
+                "time": time_val
+            }])
+            singlefile_timings.append(time_data)
 
 multifile_timings = []
 for mz_i in top_masses:
@@ -67,7 +74,7 @@ for mz_i in top_masses:
             db_path = f"E:/mzsql/MTBLS10066/multifile_{n_files}{db_ending}"
             print(db_path)
             rep_function = f"get_chrom_{db_name}('{db_path}', {mz_i}, 5)"
-            time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=3)
+            time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
             time_data = pd.DataFrame({"method":db_name, "mz_target":mz_i, "n_files":n_files, "type":"multifile",
                                       "mzml_file":os.path.basename(mzml_file), 
                                       "time":time_vals})

### 2025-02-21 manuscript/scripts/multifile_timing.py
Adding 3 and 30 to multifile systems

@@ -21,9 +21,15 @@ multifile_subset = [f.replace("\\", "/") for f in multifile_subset]
 # turn_mzml_sqlite(multifile_subset[0], outfile="E:/mzsql/MTBLS10066/multifile_1.sqlite", ordered="mz")
 # turn_mzml_duckdb(multifile_subset[0], outfile="E:/mzsql/MTBLS10066/multifile_1.duckdb", ordered="mz")
 # turn_mzml_parquet(multifile_subset[0], outdir="E:/mzsql/MTBLS10066/multifile_1_pqds", ordered="mz")
+# turn_mzml_sqlite(multifile_subset[:3], outfile="E:/mzsql/MTBLS10066/multifile_3.sqlite", ordered="mz")
+# turn_mzml_duckdb(multifile_subset[:3], outfile="E:/mzsql/MTBLS10066/multifile_3.duckdb", ordered="mz")
+# turn_mzml_parquet(multifile_subset[:3], outdir="E:/mzsql/MTBLS10066/multifile_3_pqds", ordered="mz")
 # turn_mzml_sqlite(multifile_subset[:10], outfile="E:/mzsql/MTBLS10066/multifile_10.sqlite", ordered="mz")
 # turn_mzml_duckdb(multifile_subset[:10], outfile="E:/mzsql/MTBLS10066/multifile_10.duckdb", ordered="mz")
 # turn_mzml_parquet(multifile_subset[:10], outdir="E:/mzsql/MTBLS10066/multifile_10_pqds", ordered="mz")
+# turn_mzml_sqlite(multifile_subset[:30], outfile="E:/mzsql/MTBLS10066/multifile_30.sqlite", ordered="mz")
+# turn_mzml_duckdb(multifile_subset[:30], outfile="E:/mzsql/MTBLS10066/multifile_30.duckdb", ordered="mz")
+# turn_mzml_parquet(multifile_subset[:30], outdir="E:/mzsql/MTBLS10066/multifile_30_pqds", ordered="mz")
 # turn_mzml_sqlite(multifile_subset, outfile="E:/mzsql/MTBLS10066/multifile_100.sqlite", ordered="mz")
 # turn_mzml_duckdb(multifile_subset, outfile="E:/mzsql/MTBLS10066/multifile_100.duckdb", ordered="mz")
 # turn_mzml_parquet(multifile_subset, outdir="E:/mzsql/MTBLS10066/multifile_100_pqds", ordered="mz")
[…]

### 2025-02-22 manuscript/manuscript.Rmd
Adding new edits to manuscript

@@
   - Anitra E. Ingalls
-output: word_document
+output: 
+  word_document:
+    reference_docx: template.docx
 bibliography:
@@
 ```{r setup, include=FALSE}
-options(timeout=3600)
-options(pillar.sigfig=7)
-options(digits=10)
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, include=FALSE, eval=FALSE)
 library(tidyverse)
@@
 Both of these systems are widely used for big data and can be applied to MS data in a plethora of ways, leading to the proliferation of implementations we see today.
-Both backends provide excellent universality, larger-than-memory support, and rapid access to data, but HDF5-based systems excel at self-description and hierarchical structures [@Askenazi2017] while the relational database model is optimized for multi-table queries using a consistent syntax [@Cobb1970].
+Both backends provide excellent universality, larger-than-memory support, and rapid access to data, but HDF5-based systems excel at self-description and hierarchical structures [@Askenazi2017] while the relational database model is optimized for multi-table queries using a consistent syntax [@Codd1970].
 Relational databases are increasingly seen in MS workflows for both raw and processed data, with SQLite backends now supported in the popular peakpicking software xcms [@Smith2006] via the Spectra package [@Rainer2022] (though in-memory and HDF5 options are also supported) and on MetabolomicsWorkbench [@Sud2016] while the development of MassQL [@Jarmusch2022] demonstrates the increasing comfort that MS analysts have with the adoption of SQL.
@@
 While HDF5 is a common scientific data format, databases are constantly under development by industry titans deeply invested in their maintenance and optimization.
-Online analytical processing (OLAP) methods are particularly well suited for MS data given their optimization for read speed under the assumption of infrequent transactions, making modern systems such as DuckDB [@Raasveldt2019] or Apache Arrow’s columnar formats highly appealing while preserving the familiar serverless approach.
+Online analytical processing (OLAP) methods are particularly well suited for MS data given their optimization for read speed under the assumption of infrequent transactions, making modern systems such as DuckDB [@Raasveldt2019] or Apache's Parquet formats highly appealing while preserving the familiar file-based serverless approach.
 
-Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage and access.
+Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage.
 Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity tuples is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
@@
 
-Details about the size, number of scans (MS1 and MS2), RT range, etc. for the files
+We browsed Metabolights [@Haug2019] for suitable LCMS datasets, looking for studies that included 100+ gigabytes of data from both full scan and MS/MS analysis.
+We were also restricted to the Thermo Scientific .raw file format, as it was the most widely supported by alternative MS storage methods.
+We also excluded polarity-switching data as it is unclear whether all converters would be able to separate scans based on polarity.
 
-Metabolights search for
-1. MS1 & MS2 data
-2. .raw or .wiff file types (mzDB doesn't support .d or .mzML or .mzXML)
-3. 100+ GB folder
-4. Positive mode only(?)
+Files were downloaded as .raw.
+mzML, mz5, and mzMLb were all natively supported by Proteowizard's `msconvert` software (version 3.0.25009) while MZA (v1.24.11.16) and mzDB (v0.9.10_build20170802) had separate extensions to this executable enabling their own conversion.
+MzTree and mzMD were converted via their GUI which did not have release or versioning information available but were downloaded from Github (https://github.com/optimusmoose/MZTree and https://github.com/yrm9837/mzMD-java, respectively) and built via Maven (v3.9.9) for Java (v21.0.6).
+SQL databases were built using Python 3.11.11 with SQLite (v3.48.0) via the Python sqlite3 package (v2.6.0), DuckDB via the duckdb package (v1.1.3), and Parquet files via the pyarrow package (v19.0.0).
+We were additionally able to download and install
 
-Converted via msconvert or custom implementation
+mzML access was done with Python's pyteomics package (v4.7.5), the pymzml package (v2.5.10), and the pyopenms package (3.0.0.dev20230306).
+MZA files were accessed via the mzapy library (v1.8.dev4 from the no_full_mz_array branch on Github) and via custom code built around the h5py package (3.12.1).
+Custom parsers were required for mzDB, mz5, MzTree, and mzMD.
 
-```{shell}
+```{shell, eval=FALSE}
 cd Documents\Will
@@
 msconvert *.raw --filter "peakPicking true 1-" --filter "polarity positive" 
-# Also convert neg mode to see what benefit to offer pos/neg combos
-msconvert *.raw --filter "peakPicking true 1-" --filter "polarity negative" -o "negative"
 msconvert *.raw --mzMLb --filter "peakPicking true 1-" --filter "polarity positive" 
 msconvert *.raw --mz5 --filter "peakPicking true 1-" --filter "polarity positive"
-
 for %f in (*.raw) do "C:\Program Files\ProteoWizard\raw2mzDB_0.9.10_build20170802\raw2mzDB.exe" -i %f -c 1-2 --safeMode
-
 for %f in (*.mzML) do mza.exe -file %f
@@
 
-## Hardware and software versioning
+# Results
 
-# Results
+We settled on a large dataset of gut microbiota LCMS files published in @Portlock2025 and available on Metabolights under accession number MTBLS10066.
 
@@
 mzDB and MZA both had reasonably good documentation, both formats notably providing self-contained extensions to `msconvert` for ease of conversion.
-However, both converters provide limited coverage, with mzDB notably missing support for Waters and Agilent .d files while MZA currently lacks support for AB Sciex .wiff and Bruker .baf files and is only available via binary executable (.exe), restricting its use to Windows platforms.
-Additionally, both parsers appear to be unable to separate scans from a polarity-switching experiment or support any of the other filters available natively in `msconvert`, as additional arguments passed to the executable throw errors instead of being passed along to the algorithm.
+However, both converters provide limited coverage, with mzDB notably missing support for Waters and Agilent .d files while MZA currently lacks support for AB Sciex .wiff and Bruker .baf files.
+Both converters are only available via binary executable (.exe), restricting their use to Windows platforms.
+Additionally, both parsers appear to be unable to separate scans from a polarity-switching experiment or support any of the other filters available natively in `msconvert`, as additional arguments passed to the executable throw errors instead of being passed along to the original software.
 
@@
 
+The Aird file type was straightforward to convert on Windows via the executable available on Github (v6.0.0) but does not seem to be available for other operating systems, much like MZA and mzDB.
+Unfortunately the Python package designed to allow an interface to the file type has been deprecated and we were unable to install or use it in a meaningful way and were unable to reverse-engineer the file type sufficiently to compare it here.
+The UIMF file type from the Pacific Northwest National Lab (PNNL) provided documentation exclusively in the form of C# commands and did not supply instructions for file conversion, making it unclear what input formats were supported.
+The toffee format also provided zero documentation for conversion from other formats and was restricted to time-of-flight (TOF) data independent acquisition (DIA) MS data.
+Thus, we were unable to directly compare any of these three file types to the others.
+
 ### Universal lack of support for the six relevant queries
-
-[TODO: note the dependencies required by each of the mzML parsers (and mzapy/mz5?)]
 
@@
 pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb.
-The pyOpenMS package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS1 data.
+The pyopenms package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS<sup>1</sup> data.
 Additionally, pyOpenMS required installing an old version of the package (3.0.0), Python itself (3.11) and the numpy package (<2.0) due to more recent builds requiring AVX support which was unavailable on our hardware.
-pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing.
+pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing (February 2025).
 
@@
 This was largely due to the variable-length nested compound structures mz5 that are not supported in all APIs, e.g.
-Java (@Bhamber2020).
+Java (@Bhamber2021).
 
@@
 The `mzapy` package provides a clear example of chromatogram extraction as well as a method for retention time range extraction, though there exists no clear function for the extraction of a single spectrum by scan number despite the internal file structure being highly optimized for this purpose.
-`mzapy` also provides good support for ion mobility extraction but fails to index MS2 information or provide any clear way to extract fragments by *m/z* or precursor.
+`mzapy` also provides good support for ion mobility extraction but fails to index MS/MS information or provide any clear way to extract fragments by *m/z* or precursor.
 
-MZTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server to be launched that could then be queried via an API.
+MZTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server that is then be queried via an HTTP API.
 For users without prior knowledge of HTTP request methods or exposure to programming APIs, the README is entirely unhelpful because it simply documents the API's endpoints and provides no complete query strings as examples to guide the user.
-This combination of GUI server and command-line HTTP request inverts the typical paradigm of GUI for exploration and command line for construction to unnecessarily complicated results.
+This combination of GUI server and command-line HTTP request inverts the typical paradigm of GUI for exploration and command line for construction to convoluted effect, though the structure of the data returned by the server is impressively simple.
 More problematically for this analysis, the API provides no apparent way to access MS/MS data or query the files by scan number, with only RT and *m/z* bounds controlling the subset of data extracted.
@@
 
-We found that the documentation for SQLite, DuckDB, and Parquet file formats in Python to far exceed the documentation available for any mzML parser.
+We found that the documentation for SQLite, DuckDB, and Parquet file formats in Python far exceeded the documentation available for any mzML parser.
 This is unsurprising given that these file formats are used widely outside of MS research and are developed and maintained by dedicated teams.
@@
 After establishing a connection to the database, the six queries could be asked using nearly human-readable SQL syntax.
-Requesting the thousandth MS1 scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the pandas.read_sql_query function.
+Requesting the thousandth MS1 scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the `pandas.read_sql_query` function.
 More complicated queries such as retention time range (`SELECT * FROM MS1 WHERE rt BETWEEN 6 AND 8`) and a precursor mass search (`SELECT * FROM MS2 WHERE premz BETWEEN 118.086 AND 118.087`) were similarly intuitive.
@@
 Databases also provide multiple ways to optimize queries.
+SQLite allows the construction of indices for a field within a table that then speeds up queries at the cost of additional disk space.
+Alternatively, DuckDB and Parquet files rely predominantly on the data order when it's written to disk and use their sophisticated row group methodology when subsetting.
+
+We found that SQLite queries benefitted significantly from the construction of an index on the *m/z* column when extracting chromatograms, improving lookup times by an order of magnitude (dropping from [0.3] seconds to [0.03 seconds], Figure XX).
+However, because the SQLite index is stored on disk alongside the data, this improvement also increased the file size by 33% from [150MB] to [210MB].
+DuckDB also improved significantly with ordered data but to a smaller degree and surprisingly required additional space to do so, likely due to the reordering resulting in a different compression strategy (Figure XX).
+Parquet files had the smallest improvement upon data ordering but remained the same small size they were when unordered.
+
+Crucially, these improvements also persisted when multiple files were stored in a single database.
+We built databases consisting of between 1 and 100 individual MS files and tested the time required to extract ion chromatograms from each after an index was constructed (Figure XX, left panel).
+DuckDB was consistently the fastest ion chromatogram extraction method, with query times around 0.03 seconds for a single file and 1 second for one hundred files.
+SQLite had much higher variance and slower extraction times with datasets consisting of more than one file, typically an order of magnitude slower than DuckDB, while Parquet fell between the two.
+Importantly, only DuckDB had a slope significantly less than one.
+This is what would be expected if the database was performing a simple binary search on the index, with an expected time efficiency of O(log(# of files)).
+However, DuckDB's performance degrades at larger database sizes and approaches a 1:1 slope, possibly due to the overhead of reading large amounts of data into memory after it's found.
+We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure XX, right panel).
+Interestingly, SQLite timings were slightly better with multiple database files (Figure XX, bottom panel) instead of one combined database.
 
 ```{r}
-read_csv("data/idx_ord_times.csv") %>%
+idx_gp <- read_csv("data/idx_ord_times.csv") %>%
+  mutate(file_size=ifelse(method=="Parquet", file_size-1, file_size)) %>%
   ggplot() +
-  geom_boxplot(aes(x=file_size/1e6, y=time, color=method)) +
-  scale_y_log10()
+  geom_boxplot(aes(x=file_size/1e6, y=time, color=method), position=position_identity(), width=10, fill="transparent") +
+  scale_y_log10() +
+  labs(x="Database size (MB)", y="Time (seconds)", color=NULL) +
+  theme_bw() +
+  theme(legend.position="inside", legend.justification.inside = c(0, 0),
+        legend.background = element_rect(color="black")) +
+  guides(color=guide_legend(nrow=2))
+
+multifile_data <- read_csv("data/multifile_times.csv") %>%
+  mutate(type=factor(type, levels=c("multifile", "singlefile"),
+                     labels=c("Single database", "One database per MS file"))) %>%
+  mutate(method=factor(method, levels=c("sqlite", "duckdb", "parquet"),
+                       labels=c("SQLite", "DuckDB", "Parquet"))) %>%
+  mutate(n_file_fact=factor(n_files, levels=c(1, 3, 10, 30, 100), 
+                        labels=c("One file", "3 files", "10 files", 
+                                "30 files", "100 files")))
+
+multifile_data %>%
+  select(method, n_files, time, type) %>%
+  mutate(n_files=log10(n_files), time=log10(time)) %>%
+  nest(data=c(n_files, time)) %>%
+  mutate(lmfit=map(data, ~broom::tidy(lm(.x$time~.x$n_files, offset = .x$n_files)))) %>%
+  unnest(lmfit) %>%
+  filter(term!="(Intercept)")
+
+xdb_gp <- multifile_data %>%
+  ggplot(aes(x=n_files, y=time, color=method, fill=method)) +
+  geom_abline(slope = 1, intercept = -5:3) +
+  geom_smooth(method="lm", formula=y~x, position=position_dodge(width=0.2), fill="grey50") +
+  geom_point(position=position_dodge(width=0.2), color="black", pch=21) +
+  facet_wrap(~type, nrow=1) +
+  scale_y_log10() +
+  scale_x_log10() +
+  # scale_color_manual(breaks = method_names, values=method_colors, aesthetics = c("color", "fill")) +
+  scale_color_viridis_d(aesthetics = c("color", "fill"), begin = 0.2, end = 0.8) +
+  labs(x="Number of files", y="Time (seconds)", color=NULL, fill=NULL) +
+  theme_bw() +
+  theme(legend.position="inside", legend.justification.inside = c(0.05, 0.95),
+        legend.background = element_rect(color="grey80"))
+
+xmeth_gp <- multifile_data %>%
+  ggplot() +
+  geom_boxplot(aes(x=n_file_fact, y=time, color=type)) +
+  facet_wrap(~method, nrow=1) +
+  scale_y_log10() +
+  scale_color_viridis_d(option = "A", begin = 0.4, end = 0.8) +
+  theme_bw() +
+  labs(x="Number of files", y="Time (seconds)", color="Database type") +
+  theme(legend.position="top")
+
+idx_gp + (xdb_gp/xmeth_gp + plot_layout(heights = c(2,1)))
+ggsave("figures/multifile_fig.png", width = 10.5, height=7)
 ```
+
+![](figures/multifile_fig.png)
 
@@
 DuckDB's zonemaps operate similarly to mzDB's bounding boxes
+DuckDB also has automatic compression detection (e.g. the delta encoding of mz5)
 

### 2025-02-27 manuscript/manuscript.Rmd
Cleaning up conversion code

@@
 
-msconvert *.raw --filter "peakPicking true 1-" --filter "polarity positive" 
-msconvert *.raw --mzMLb --filter "peakPicking true 1-" --filter "polarity positive" 
-msconvert *.raw --mz5 --filter "peakPicking true 1-" --filter "polarity positive"
-for %f in (*.raw) do "C:\Program Files\ProteoWizard\raw2mzDB_0.9.10_build20170802\raw2mzDB.exe" -i %f -c 1-2 --safeMode
-for %f in (*.mzML) do mza.exe -file %f
+# After adding msconvert, raw2mzdb, and mza executables to PATH:
+msconvert 20220921_LEAP-POS_BL01.raw --filter "peakPicking true 1-"
+msconvert 20220921_LEAP-POS_BL01.raw --mzMLb --filter "peakPicking true 1-"
+msconvert 20220921_LEAP-POS_BL01.raw --mz5 --filter "peakPicking true 1-"
+raw2mzDB -i 20220921_LEAP-POS_BL01.raw  -c 1-2 --safeMode
+mza -file 20220921_LEAP-POS_BL01.raw
 ```

### 2025-02-28 manuscript/manuscript.Rmd
Tidying shell commands

@@
 SQL databases were built using Python 3.11.11 with SQLite (v3.48.0) via the Python sqlite3 package (v2.6.0), DuckDB via the duckdb package (v1.1.3), and Parquet files via the pyarrow package (v19.0.0).
-We were additionally able to download and install
+We were additionally able to download and install []
 
@@
 C:\Users\willi\Documents\mzsql\java_things\apache-maven-3.9.9\bin\mvn package
-java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar target\msDataServer-1.0.jar
+java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar java_things\MZTree\msDataServer\target\msDataServer-1.0.jar
 (manually) open mztree file in port 4568
@@
 C:\Users\willi\Documents\mzsql\java_things\apache-maven-3.9.9\bin\mvn package
-java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar target\msDataServer-1.0.jar
+java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar java_things\mzMD-java\target\msDataServer-1.0.jar
 (manually) open mzmd file in port 4567
-cd ..
 
@@
 # After adding msconvert, raw2mzdb, and mza executables to PATH:
-msconvert 20220921_LEAP-POS_BL01.raw --filter "peakPicking true 1-"
-msconvert 20220921_LEAP-POS_BL01.raw --mzMLb --filter "peakPicking true 1-"
-msconvert 20220921_LEAP-POS_BL01.raw --mz5 --filter "peakPicking true 1-"
-raw2mzDB -i 20220921_LEAP-POS_BL01.raw  -c 1-2 --safeMode
-mza -file 20220921_LEAP-POS_BL01.raw
+msconvert 20220923_LEAP-POS_QC04.raw --filter "peakPicking true 1-"
+msconvert 20220923_LEAP-POS_QC04.raw --mzMLb --filter "peakPicking true 1-"
+msconvert 20220923_LEAP-POS_QC04.raw --mz5 --filter "peakPicking true 1-"
+# Needs full path or it crashes on startup?
+"C:\Program Files\ProteoWizard\raw2mzDB_0.9.10_build20170802\raw2mzDB.exe" -i 20220923_LEAP-POS_QC04.raw  -c 1-2 --safeMode
+mza -file 20220923_LEAP-POS_QC04.raw
 ```

### 2025-02-28 manuscript/scripts/dbloopjoinuni.py
Big script cleanup before rerun

@@ -1,23 +1,22 @@
 
 import numpy as np
 import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
 from mzsql import *
 import sqlite3
 import os
 import random
+import glob
 
 random.seed(123)
-files = ["E:/mzsql/MTBLS10066/"+f for f in os.listdir("E:/mzsql/MTBLS10066/") if f.endswith(".mzML")]
-msfiles = random.sample(files, 10)
+basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
+print(basename)
 
-if(not(os.path.exists("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite"))):
-    turn_mzml_duckdb(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.duckdb")
-    turn_mzml_sqlite(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
-    turn_mzml_parquet(msfiles[0], "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.parquet")
+if(not(os.path.exists(f"{basename}.sqlite"))):
+    turn_mzml_sqlite(f"{basename}.mzML", f"{basename}.sqlite")
+    turn_mzml_duckdb(f"{basename}.mzML", f"{basename}.duckdb")
+    turn_mzml_parquet(f"{basename}.mzML", f"{basename}_pqds")
 
-conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
+conn = sqlite3.connect(f"{basename}.sqlite")
 top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
 top_masses = []
 top_rts = []
@@ -27,6 +26,7 @@ for i in range(100):
     top_masses.append(top_mz)
     mzmin, mzmax = pmppm(top_mz, 10)
     top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax))]
+conn.close()
 
 def get_multichrom_duckdb_loop(file, mzs, ppm):
     conn = duckdb.connect(file)
@@ -156,7 +156,7 @@ for n_chroms in [1, 3, 10, 30, 100]:
     print(n_chroms)
[…]

### 2025-02-28 manuscript/scripts/idx_ord_timing.py

@@ -1,64 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import sqlite3
-import os
-from mzsql import *
-
-# Create databases from mzML files if they don't already exist
-if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite"))):
-    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
-    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.duckdb")
-    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds")
-
-if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_ord.sqlite"))):
-    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_ord.sqlite", ordered="mz")
-    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_ord.duckdb", ordered="mz")
-    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds_ord", ordered="mz")
-
-conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
-top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
-top_masses = []
-top_rts = []
-for i in range(10):
-    top_rts.append(top_int_df["rt"].iloc[0])
-    top_mz = top_int_df["mz"].iloc[0]
-    top_masses.append(top_mz)
-    mzmin, mzmax = pmppm(top_mz, 10)
[…]

### 2025-02-28 manuscript/scripts/multichrom_timing.py

@@ -1,19 +1,23 @@
+
 import numpy as np
 import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
 from mzsql import *
 import sqlite3
 import os
 import random
+import glob
+
+random.seed(123)
+basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
+print(basename)
 
 # Create databases from mzML files if they don't already exist
-if(not(os.path.exists("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite"))):
-    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
-    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.duckdb")
-    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", "E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04_pqds")
+if(not(os.path.exists(f"{basename}.sqlite"))):
+    turn_mzml_sqlite(f"{basename}.mzML", f"{basename}.sqlite")
+    turn_mzml_duckdb(f"{basename}.mzML", f"{basename}.duckdb")
+    turn_mzml_parquet(f"{basename}.mzML", f"{basename}_pqds")
 
-conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.sqlite")
+conn = sqlite3.connect(f"{basename}.sqlite")
 top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
 top_masses = []
 top_rts = []
@@ -23,6 +27,7 @@ for i in range(100):
     top_masses.append(top_mz)
     mzmin, mzmax = pmppm(top_mz, 10)
     top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax))]
+conn.close()
 
 def get_multichrom_mzml_pyteomics(file, mzs, ppm):
     file_data = pyteomics.mzml.MzML(file)
[…]

### 2025-02-28 manuscript/scripts/multifile_sizetimes.py

@@ -1,0 +1,109 @@
+import numpy as np
+import pandas as pd
+import sqlite3
+import os
+import glob
+import random
+import timeit
+from mzsql import *
+
+random.seed(123)
+multifile_subset = random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 100)
+multifile_subset = [f.replace("\\", "/") for f in multifile_subset]
+
+file_sequence = [1, 3, 10, 30, 100]
+
+for n_files in file_sequence:
+    for db_sort in ["ordered", "unordered"]:
+        ordering = "mz" if db_sort == "ordered" else None
+        for db_name in ["duckdb", "sqlite", "parquet"]:
+            db_ending = {"duckdb": ".duckdb", "sqlite": ".sqlite", "parquet": "_pqds"}
+            file_ending = "_ord" if db_sort == "ordered" else ""
+            file_ending += db_ending[db_name]
+            for db_type in ["consolidated", "multifile"]:
+                this_multifile_subset = multifile_subset[:n_files]
+                if db_type == "multifile":
+                    if n_files < max(file_sequence):
+                        continue
+                    for mzml_i in this_multifile_subset:
+                        output_db = mzml_i.replace(".mzML", file_ending)
+                        print(output_db)
+                        combined_cmd = f"turn_mzml_{db_name}(mzml_i, output_db, ordered=ordering)"
+                        eval(combined_cmd)
+                else:
+                    output_db = f"E:/mzsql/MTBLS10066/consolidated_{n_files}{file_ending}"
+                    print(output_db)
+                    combined_cmd = f"turn_mzml_{db_name}(this_multifile_subset, output_db, ordered=ordering)"
+                    eval(combined_cmd)
+
+conn = sqlite3.connect(f"E:/mzsql/MTBLS10066/consolidated_{max(file_sequence)}.sqlite")
+top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
+top_masses = []
+top_rts = []
+for i in range(10):
+    top_rts.append(top_int_df["rt"].iloc[0])
+    top_mz = top_int_df["mz"].iloc[0]
+    top_masses.append(top_mz)
+    mzmin, mzmax = pmppm(top_mz, 10)
+    rtmin, rtmax = top_int_df["rt"].iloc[0]+(-1, 1)
[…]

### 2025-02-28 manuscript/scripts/multifile_timing.py

@@ -1,91 +1,0 @@
-import numpy as np
-import pandas as pd
-import seaborn as sns
-import sqlite3
-import os
-import glob
-import random
-from mzsql import *
-
-random.seed(123)
-multifile_subset = random.sample(glob.glob("E:/mzsql/MTBLS10066/202*.mzML"), 100)
-multifile_subset = [f.replace("\\", "/") for f in multifile_subset]
-
-# Create individual sqlite, duckdb, and parquet files
-# for file_i in multifile_subset:
-#     turn_mzml_sqlite(file_i, outfile=file_i.replace(".mzML", ".sqlite"), ordered="mz")
-#     turn_mzml_duckdb(file_i, outfile=file_i.replace(".mzML", ".duckdb"), ordered="mz")
-#     turn_mzml_parquet(file_i, outdir=file_i.replace(".mzML", "_pqds"), ordered="mz")
-
-# Create multifile dbs
-# turn_mzml_sqlite(multifile_subset[0], outfile="E:/mzsql/MTBLS10066/multifile_1.sqlite", ordered="mz")
-# turn_mzml_duckdb(multifile_subset[0], outfile="E:/mzsql/MTBLS10066/multifile_1.duckdb", ordered="mz")
-# turn_mzml_parquet(multifile_subset[0], outdir="E:/mzsql/MTBLS10066/multifile_1_pqds", ordered="mz")
-# turn_mzml_sqlite(multifile_subset[:3], outfile="E:/mzsql/MTBLS10066/multifile_3.sqlite", ordered="mz")
-# turn_mzml_duckdb(multifile_subset[:3], outfile="E:/mzsql/MTBLS10066/multifile_3.duckdb", ordered="mz")
-# turn_mzml_parquet(multifile_subset[:3], outdir="E:/mzsql/MTBLS10066/multifile_3_pqds", ordered="mz")
-# turn_mzml_sqlite(multifile_subset[:10], outfile="E:/mzsql/MTBLS10066/multifile_10.sqlite", ordered="mz")
-# turn_mzml_duckdb(multifile_subset[:10], outfile="E:/mzsql/MTBLS10066/multifile_10.duckdb", ordered="mz")
-# turn_mzml_parquet(multifile_subset[:10], outdir="E:/mzsql/MTBLS10066/multifile_10_pqds", ordered="mz")
[…]

### 2025-02-28 manuscript/scripts/singlefile_supp.py

@@ -1,44 +1,50 @@
+
 import pandas as pd
 import numpy as np
 import timeit
+import glob
+import random
 import pyopenms
 import pyteomics.mzml
 import pyteomics.mzmlb
 import pymzml
 import h5py
-import seaborn as sns
 import matplotlib.pyplot as plt
 from mzsql import *
 
-def get_randscan_mzml_pyopenms(scan_num):
+random.seed(123)
+basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
+print(basename)
+
+def get_randscan_mzml_pyopenms(basename, scan_num):
     exp = pyopenms.MSExperiment()
-    pyopenms.MzMLFile().load("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", exp)
+    pyopenms.MzMLFile().load(f"{basename}.mzML", exp)
     spec_data = exp[scan_num].get_peaks()
     return(pd.DataFrame({"mz":spec_data[0], "int":spec_data[1]}))
 
-def get_randscan_mzml_pyteomics(scan_num):
-    file_data = pyteomics.mzml.MzML("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML")
+def get_randscan_mzml_pyteomics(basename, scan_num):
+    file_data = pyteomics.mzml.MzML(f"{basename}.mzML")
     return(pd.DataFrame({"mz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
 
-def get_randscan_mzml_pymzml(scan_num):
-    run = pymzml.run.Reader("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzML", build_index_from_scratch=True)
+def get_randscan_mzml_pymzml(basename, scan_num):
+    run = pymzml.run.Reader(f"{basename}.mzML", build_index_from_scratch=True)
     spec1_data = run[scan_num].peaks("raw")
     return(pd.DataFrame({"mz":spec1_data[:,0], "int":spec1_data[:,1]}))
 
-def get_randscan_mzmlb(scan_num):
-    file_data = pyteomics.mzmlb.MzMLb("E:/mzsql/MTBLS10066/20220923_LEAP-POS_QC04.mzMLb")
[…]

### 2025-02-28 manuscript/scripts/singlefile_timing.py

@@ -4,25 +4,39 @@ import numpy as np
 import pandas as pd
 import sqlite3
 import os
+import glob
+import random
+import timeit
 from mzsql import *
 
-# Create databases from mzML files if they don't already exist
-if(not(os.path.exists("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite"))):
-    turn_mzml_sqlite("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
-    turn_mzml_duckdb("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.duckdb")
-    turn_mzml_parquet("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.mzML", "E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01_pqds")
+random.seed(123)
+basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
+print(basename)
+# 20220923_LEAP-POS_QC04
 
-# Identify the bits of information to be extracted
-conn = sqlite3.connect("E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01.sqlite")
-# Get 10 random spectra
-rand_scan_query = """
-SELECT id 
-FROM (SELECT DISTINCT id FROM {0}) 
-ORDER BY RANDOM() 
-LIMIT 10
-"""
-rand_MS1_scans = pd.read_sql_query(rand_scan_query.format("MS1"), conn)["id"].tolist()
-rand_MS2_scans = pd.read_sql_query(rand_scan_query.format("MS2"), conn)["id"].tolist()
+# Create databases from mzML files if they don't already exist
+if(not(os.path.exists(f"{basename}.sqlite"))):
+    turn_mzml_sqlite(f"{basename}.mzML", f"{basename}.sqlite")
+    turn_mzml_duckdb(f"{basename}.mzML", f"{basename}.duckdb")
+    turn_mzml_parquet(f"{basename}.mzML", f"{basename}_pqds")
+
+# # Identify the bits of information to be extracted
+conn = sqlite3.connect(f"{basename}.sqlite")
[…]

### 2025-02-28 src/mzsql/__init__.py
Removing separate timing_functions script

@@ -10,4 +10,3 @@ from .mza_functions import *
 from .mzTree_functions import *
 from .mzMD_functions import *
 from .mzdb_functions import *
-from .timing_functions import *

### 2025-02-28 src/mzsql/timing_functions.py

@@ -1,64 +1,0 @@
-import time
-import timeit
-from mzsql import *
-
-def time_chrom(fun_suffix, file_ending, target_mz, ppm_acc, verbose=True):
-    if(fun_suffix in ["mztree", "mzMD"]):
-        rep_function = f"get_chrom_{fun_suffix}('{file_ending}', {target_mz}, {ppm_acc})"
-    else:
-        rep_function = f"get_chrom_{fun_suffix}('E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01{file_ending}', {target_mz}, {ppm_acc})"
-    if(verbose):
-        start_time = time.time()
-        print(f"Running {fun_suffix}", end="\r")
-    time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
-    if(verbose):
-        elapsed = time.time() - start_time
-        print(f"Running {fun_suffix}... ({elapsed:.2f}s)")
-    return(time_vals)
-
-def time_spec(fun_suffix, file_ending, spec_id, verbose=True):
-    rep_function = f"get_spec_{fun_suffix}('E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01{file_ending}', {spec_id})"
-    if(verbose):
-        print(f"Running {fun_suffix}", end="\r")
-        start_time = time.time()
-    time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
-    if(verbose):
-        elapsed = time.time() - start_time
-        print(f"Running {fun_suffix}... ({elapsed:.2f}s)")
-    return(time_vals)
-
-def time_rtrange(fun_suffix, file_ending, rt, verbose=True):
-    if(fun_suffix in ["mztree", "mzMD"]):
-        rep_function = f"get_rtrange_{fun_suffix}('{file_ending}', {rt-0.2}, {rt+0.2})"
-    else:
-        rep_function = f"get_rtrange_{fun_suffix}('E:/mzsql/MTBLS10066/20220921_LEAP-POS_BL01{file_ending}', {rt-0.2}, {rt+0.2})"
-    if(verbose):
-        print(f"Running {fun_suffix}", end="\r")
-        start_time = time.time()
-    time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
-    if(verbose):
-        elapsed = time.time() - start_time
-        print(f"Running {fun_suffix}... ({elapsed:.2f}s)")
[…]

### 2025-02-28 src/mzsql/sqlite_functions.py
Now writing filename out even if it's just a single file

@@ -9,13 +9,16 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
     conn.execute("DROP TABLE IF EXISTS MS1")
     conn.execute("DROP TABLE IF EXISTS MS2")
     if isinstance(files, str):
-        for spectrum in pyteomics.mzml.MzML(files):
+        files = [files]
+    for file in files:
+        for spectrum in pyteomics.mzml.MzML(file):
             if spectrum['ms level'] == 1:
                 idx = int(spectrum['id'].split("scan=")[-1].split()[0])
                 mz_vals = spectrum['m/z array']
                 int_vals = spectrum['intensity array']
                 rt_val = spectrum['scanList']['scan'][0]['scan start time']
-                df_scan = pd.DataFrame({'id': idx, 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'mz': mz_vals, 
+                                        'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
                 df_scan.to_sql("MS1", conn, if_exists="append", index=False)
             if spectrum['ms level'] == 2:
                 idx = int(spectrum['id'].split("scan=")[-1].split()[0])
@@ -23,31 +26,9 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
                 int_vals = spectrum['intensity array']
                 rt_val = spectrum['scanList']['scan'][0]['scan start time']
                 premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-                df_scan = pd.DataFrame({'id': idx, 'premz': premz_val, 'fragmz': mz_vals, 
-                                        'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'premz': premz_val, 
+                                        'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
                 df_scan.to_sql("MS2", conn, if_exists="append", index=False)
-    else:
-        for file in files:
-            if not os.path.exists(file):
-                raise FileNotFoundError(f"Input mzML file '{file}' does not exist.")
-            for spectrum in pyteomics.mzml.MzML(file):
-                if spectrum['ms level'] == 1:
-                    idx = int(spectrum['id'].split("scan=")[-1].split()[0])
[…]

### 2025-02-28 src/mzsql/mzml_functions.py
Fixed bug from overzealous cleanup

@@ -162,6 +162,7 @@ def get_chrom_mzml_pyopenms_2DPeak(file, mz, ppm):
 def get_spec_mzml_pyopenms(file, scan_num):
     exp = pyopenms.MSExperiment()
     pyopenms.MzMLFile().load(file, exp)
+    i = 0
     while i <= exp.size():
         if(int(exp[i].getNativeID().split("scan=")[-1].split()[0]) == scan_num):
             spec_data = exp[i].get_peaks()

### 2025-02-28 src/mzsql/parquet_functions.py
Fixed multi-file appending bug for parquet files

@@ -10,11 +10,11 @@ import pyarrow.dataset as ds
 import os
 
 def turn_mzml_parquet(files, outdir, ordered=None):
-    MS1_dfs = []
-    MS2_dfs = []
     if isinstance(files, str):
         files = [files]
     for file in files:
+        MS1_dfs = []
+        MS2_dfs = []
         for spectrum in pyteomics.mzml.MzML(file):
             idx = int(spectrum['id'].split("scan=")[-1].split()[0])
             mz_vals = spectrum['m/z array']

### 2025-02-28 src/mzsql/sqlite_functions.py
Now writing to disk after every file, not every scan

@@ -11,6 +11,8 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
     if isinstance(files, str):
         files = [files]
     for file in files:
+		MS1_dfs = []
+		MS2_dfs = []
         for spectrum in pyteomics.mzml.MzML(file):
             if spectrum['ms level'] == 1:
                 idx = int(spectrum['id'].split("scan=")[-1].split()[0])
@@ -19,7 +21,7 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
                 rt_val = spectrum['scanList']['scan'][0]['scan start time']
                 df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'mz': mz_vals, 
                                         'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                df_scan.to_sql("MS1", conn, if_exists="append", index=False)
+                MS1_dfs.append(df_scan)
             if spectrum['ms level'] == 2:
                 idx = int(spectrum['id'].split("scan=")[-1].split()[0])
                 mz_vals = spectrum['m/z array']
@@ -28,8 +30,9 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
                 premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
                 df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'premz': premz_val, 
                                         'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                df_scan.to_sql("MS2", conn, if_exists="append", index=False)
-    
+                MS2_dfs.append(df_scan)
+		pd.concat(MS1_dfs, ignore_index=True).to_sql("MS1", conn, if_exists="append", index=False)
+		pd.concat(MS2_dfs, ignore_index=True).to_sql("MS2", conn, if_exists="append", index=False)
     if ordered is not None:
         index_name = f"idx_{ordered}"
         if ordered == "rt":

### 2025-02-28 manuscript/scripts/multichrom_timing.py
Fixing bugs in multichrom_timing

@@ -6,6 +6,7 @@ import sqlite3
 import os
 import random
 import glob
+import timeit
 
 random.seed(123)
 basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
@@ -82,7 +83,7 @@ def get_multichrom_mzml_pyopenms(file, mzs, ppm):
                     df_scan = pd.DataFrame({'mz_i':mz_i, 'mz':mz_vals[bet_idxs], 'int':int_vals[bet_idxs], 'rt':[rt_val]*sum(bet_idxs)})
                     scan_dfs.append(df_scan)
     return(pd.concat(scan_dfs, ignore_index=True))
-def get_multichrom_mzml_pyopenms_2Dpeak(file, mzs, ppm):
+def get_multichrom_mzml_pyopenms_2DPeak(file, mzs, ppm):
     exp = pyopenms.MSExperiment()
     pyopenms.MzMLFile().load(file, exp)
     exp.updateRanges()
@@ -170,7 +171,7 @@ def get_multichrom_mzdb(file, mzs, ppm):
 def get_multichrom_mztree(url_port, mzs, ppm):
     all_chroms = []
     for mz_i in mzs:
-        mz_min, mz_max = pmppm(mz, ppm)
+        mz_min, mz_max = pmppm(mz_i, ppm)
         request_string = f"{url_port}/api/v2/getpoints?mzmin={mz_min}&mzmax={mz_max}&rtmin=0&rtmax=1000000&numpoints=0"
         response = requests.get(request_string)
         chrom_data = pd.DataFrame(response.json(), columns=["pointId", "traceId", "mz", "rt", "intensity"])
@@ -178,10 +179,10 @@ def get_multichrom_mztree(url_port, mzs, ppm):
         chrom_data.columns = ["rt", "mz", "int"]
         all_chroms.append(chrom_data)
     return(pd.concat(all_chroms, ignore_index=True))
-def get_multichrom_mzmd(url_port, mzs, ppm):
+def get_multichrom_mzMD(url_port, mzs, ppm):
     all_chroms = []
     for mz_i in mzs:
-        mz_min, mz_max = pmppm(mz, ppm)
+        mz_min, mz_max = pmppm(mz_i, ppm)
[…]

### 2025-02-28 manuscript/scripts/dbloopjoinuni.py
Adding missing dependency

@@ -6,6 +6,7 @@ import sqlite3
 import os
 import random
 import glob
+import timeit
 
 random.seed(123)
 basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")

### 2025-02-28 src/mzsql/sqlite_functions.py
Switching from tabs to spaces

@@ -11,8 +11,8 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
     if isinstance(files, str):
         files = [files]
     for file in files:
-		MS1_dfs = []
-		MS2_dfs = []
+        MS1_dfs = []
+        MS2_dfs = []
         for spectrum in pyteomics.mzml.MzML(file):
             if spectrum['ms level'] == 1:
                 idx = int(spectrum['id'].split("scan=")[-1].split()[0])
@@ -31,8 +31,8 @@ def turn_mzml_sqlite(files, outfile, ordered=None):
                 df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'premz': premz_val, 
                                         'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
                 MS2_dfs.append(df_scan)
-		pd.concat(MS1_dfs, ignore_index=True).to_sql("MS1", conn, if_exists="append", index=False)
-		pd.concat(MS2_dfs, ignore_index=True).to_sql("MS2", conn, if_exists="append", index=False)
+        pd.concat(MS1_dfs, ignore_index=True).to_sql("MS1", conn, if_exists="append", index=False)
+        pd.concat(MS2_dfs, ignore_index=True).to_sql("MS2", conn, if_exists="append", index=False)
     if ordered is not None:
         index_name = f"idx_{ordered}"
         if ordered == "rt":

### 2025-02-28 manuscript/manuscript.Rmd
Adding filename explicitly

@@
 java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar java_things\MZTree\msDataServer\target\msDataServer-1.0.jar
-(manually) open mztree file in port 4568
+(manually) open mztree file in port 4568 (20220923_LEAP-POS_QC04.mzTree)
 
@@
 java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar java_things\mzMD-java\target\msDataServer-1.0.jar
-(manually) open mzmd file in port 4567
+(manually) open mzmd file in port 4567 (20220923_LEAP-POS_QC04.mzMD)
 

### 2025-03-03 manuscript/scripts/multifile_sizetimes.py
Finished multifile rerun

@@ -29,12 +29,12 @@ for n_files in file_sequence:
                         output_db = mzml_i.replace(".mzML", file_ending)
                         print(output_db)
                         combined_cmd = f"turn_mzml_{db_name}(mzml_i, output_db, ordered=ordering)"
-                        eval(combined_cmd)
+                        #eval(combined_cmd)
                 else:
                     output_db = f"E:/mzsql/MTBLS10066/consolidated_{n_files}{file_ending}"
                     print(output_db)
                     combined_cmd = f"turn_mzml_{db_name}(this_multifile_subset, output_db, ordered=ordering)"
-                    eval(combined_cmd)
+                    #eval(combined_cmd)
 
 conn = sqlite3.connect(f"E:/mzsql/MTBLS10066/consolidated_{max(file_sequence)}.sqlite")
 top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
@@ -105,5 +105,5 @@ for n_files in file_sequence:
 
 all_timings = pd.concat(multifile_timings, ignore_index=True)
 all_sizes = pd.concat(file_sizes, ignore_index=True)
-all_sizes.merge(all_timings).to_csv("../data/multifile_timings.csv", index=False)
+all_sizes.merge(all_timings).to_csv("data/multifile_timings.csv", index=False)
 

### 2025-03-04 manuscript/scripts/multifile_sizetimes.py
Fixed bug in multifile things

@@ -61,13 +61,15 @@ for mz_i in top_masses:
                 for db_type in ["consolidated", "multifile"]:
                     this_multifile_subset = multifile_subset[:n_files]
                     if db_type == "multifile":
+                        multifile_time = 0
                         for mzml_i in this_multifile_subset:
                             output_db = mzml_i.replace(".mzML", file_ending)
                             rep_function = f"get_chrom_{db_name}('{output_db}', {mz_i}, 5)"
                             time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
-                            time_data = pd.DataFrame({"method":db_name, "mz_target":mz_i, "n_files":n_files, 
-                                                      "type":db_type, "db_sort":db_sort, "time":time_vals})
-                            multifile_timings.append(time_data)
+                            multifile_time+=time_vals[0]
+                        time_data = pd.DataFrame({"method":db_name, "mz_target":mz_i, "n_files":n_files,
+                                                  "type":db_type, "db_sort":db_sort, "time":time_vals})
+                        multifile_timings.append(time_data)
                     else:
                         output_db = f"E:/mzsql/MTBLS10066/consolidated_{n_files}{file_ending}"
                         rep_function = f"get_chrom_{db_name}('{output_db}', {mz_i}, 5)"

### 2025-03-04 manuscript/scripts/multifile_sizetimes.py
Fixing more bugs

@@ -68,7 +68,7 @@ for mz_i in top_masses:
                             time_vals = timeit.repeat(rep_function, globals=globals(), number=1, repeat=1)
                             multifile_time+=time_vals[0]
                         time_data = pd.DataFrame({"method":db_name, "mz_target":mz_i, "n_files":n_files,
-                                                  "type":db_type, "db_sort":db_sort, "time":time_vals})
+                                                  "type":db_type, "db_sort":db_sort, "time":[multifile_time]})
                         multifile_timings.append(time_data)
                     else:
                         output_db = f"E:/mzsql/MTBLS10066/consolidated_{n_files}{file_ending}"

### 2025-03-04 manuscript/manuscript.Rmd
New manuscript draft with updated figures

@@
 
-This “vanilla” database deprioritizes the metadata associated with a given file and focuses on performance and simplicity.
-Thus, this method does not replace the existing vendor-specific or mzML files but instead constructs a parallel data structure and represents our intuition that metadata is rarely the main target of MS analysis and that labs typically preserve the original files anyway.
+The “vanilla” database style proposed here abandons a 1:1 representation of the original vendor-specific file.
+This decision was made after discussion with a wide variety of experts, all of whom preserved the original MS files even after conversion to another file type, indicating that a highly-performant addition is more important than direct replacement.
+Here, we map MS concepts (retention time, *m/z*, intensity, etc.) directly to database fields to make downstream processing as intuitive as possible.
+Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure XX).
+We do not force compression of any of these fields because we believe that decoding compressed data is both a slow and unintuitive step, though automatic compression is supplied by the DuckDB and Parquet file types.
 
-Discuss how metadata fits into these parsers
+```{r gtables for figure 1}
+library(gt)
+data.frame(
+  filename=c("Smp_A", "Smp_B", "Smp_Y", "Smp_Z"),
+  n_scans=c(1445, 1489, 1388, 1390),
+  timestamp=c("2021-08-23 04:30", "2021-08-23 05:25", "2021-08-23 22:15", "2021-08-23 23:10"),
+  `etc.`=c("...", "...", "...", "...")
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             n_scans=html("n_scans</br><small>Scan count</small>"),
+             timestamp=html("timestamp</br><small>Timestamp</small>"),
+             `etc.`=html("Etc.</br><small>...</small>")) %>%
+  tab_row_group(label = "...additional samples...", rows = !c(TRUE, TRUE, FALSE, FALSE)) %>%
+  tab_row_group(label = "", rows = c(TRUE, TRUE, FALSE, FALSE)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: file_info") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/fileinfo_table.png", vwidth=1000)
 
-Discuss links between various tables via scan_idx column?
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1, 2, 1389, 1390),
+  ms_level=c(1, 2, 2, 2),
+  rt=c(0.10, 0.12, 22.43, 22.45),
+  # min_mz=c(59.99975, 49.99992, 50.00013, 49.99978),
+  TIC=c(2331809, 820427, 9555703, 1892338),
+  `etc.`=c("...", "...", "...", "...")
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             ms_level=html("ms_level</br><small>MS level</small>"),
+             rt=html("rt</br><small>Retention time</small>"), 
+             # min_mz=html("min_mz</br><small>Lowest <emph>m/z</emph></small>"),
+             TIC=html("TIC</br><small>Ion current</small>"),
+             `etc.`=html("Etc.</br><small>...</small>")) %>%
+  tab_row_group(label = "...many additional scans...", rows = !c(TRUE, TRUE, FALSE, FALSE)) %>%
+  tab_row_group(label = "", rows = c(TRUE, TRUE, FALSE, FALSE)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: scan_info") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/scaninfo_table.png", vwidth=1000)
 
-$MS1
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,1385,1385),
+  rt=c(0.10, 0.10, 22.35, 22.35),
+  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
+  int=c(6618, 2657, 158084, 4673)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             rt=html("rt</br><small>Retention time</small>"), 
+             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
+             int=html("int</br><small>Intensity</small>")) %>%
+  tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: MS1") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/ms1_table.png", vwidth=1000)
 
-$MS2
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(2,2,1390,1390),
+  prescan=c(1,1,1385,1385),
+  ms2_rt=c(0.12, 0.12, 22.45, 22.45),
+  fragmz=c(51.0238, 53.0394, 52.0186, 57.0923),
+  premz=c(241.0894, 241.0894, 185.1932, 185.1932),
+  ms2_int=c(36104, 243165, 28371, 129604)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             prescan=html("prescan</br><small>Precursor scan</small>"),
+             ms2_rt=html("rt</br><small>Retention time</small>"), 
+             fragmz=html("fragmz</br><small>Fragment <emph>m/z</emph></small>"), 
+             premz=html("premz</br><small>Precursor <emph>m/z</emph></small>"), 
+             ms2_int=html("int</br><small>Intensity</small>")) %>%
+  tab_row_group(label = "...millions of additional entries...", rows = ms2_rt>=mean(ms2_rt)) %>%
+  tab_row_group(label = "", rows = ms2_rt<=mean(ms2_rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: MS2") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/ms2_table.png", vwidth=1000)
+```
 
-$scan_info
+![](figures/db_fig.png)
 
-$file_info
-
-$sample_metadata
-
-$rt_correction
-
-$picked_peaks
+*Figure XX: Database schema for an example MS/MS dataset showing the organization of mass-spectrometry data into tables.
+Fields of interest are easily queryable with simple SQL commands as shown in the table at bottom.*
 
@@
 
-We settled on a large dataset of gut microbiota LCMS files published in @Portlock2025 and available on Metabolights under accession number MTBLS10066.
+We settled on a large dataset of gut microbiota LC-MS files published in @Portlock2025 and available on Metabolights under accession number MTBLS10066.
 
@@
 
-The simplest and most abundantly documented query was the extraction of asingle spectrum.
+The simplest and most abundantly documented query was the extraction of a single spectrum.
 In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 1, leftmost panels).
@@
 Crucially, these improvements also persisted when multiple files were stored in a single database.
-We built databases consisting of between 1 and 100 individual MS files and tested the time required to extract ion chromatograms from each after an index was constructed (Figure XX, left panel).
+We built databases consisting of between 1 and 100 individual MS files and tested the time required to extract ion chromatograms from each after an index was constructed (Supplemental fig XX).
 DuckDB was consistently the fastest ion chromatogram extraction method, with query times around 0.03 seconds for a single file and 1 second for one hundred files.
@@
 However, DuckDB's performance degrades at larger database sizes and approaches a 1:1 slope, possibly due to the overhead of reading large amounts of data into memory after it's found.
-We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure XX, right panel).
-Interestingly, SQLite timings were slightly better with multiple database files (Figure XX, bottom panel) instead of one combined database.
+We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure XX).
+Interestingly, SQLite timings were slightly better with multiple database files instead of one combined database (Figure XX).
 
 ```{r}
-idx_gp <- read_csv("data/idx_ord_times.csv") %>%
-  mutate(file_size=ifelse(method=="Parquet", file_size-1, file_size)) %>%
-  ggplot() +
-  geom_boxplot(aes(x=file_size/1e6, y=time, color=method), position=position_identity(), width=10, fill="transparent") +
-  scale_y_log10() +
-  labs(x="Database size (MB)", y="Time (seconds)", color=NULL) +
-  theme_bw() +
-  theme(legend.position="inside", legend.justification.inside = c(0, 0),
-        legend.background = element_rect(color="black")) +
-  guides(color=guide_legend(nrow=2))
+timelabs <- c("0.01 seconds", "0.1 seconds", "One second", "Ten seconds", "One minute")
 
-multifile_data <- read_csv("data/multifile_times.csv") %>%
-  mutate(type=factor(type, levels=c("multifile", "singlefile"),
+read_csv("data/multifile_timings.csv") %>%
+  mutate(type=factor(type, levels=c("consolidated", "multifile"),
                      labels=c("Single database", "One database per MS file"))) %>%
@@
                        labels=c("SQLite", "DuckDB", "Parquet"))) %>%
-  mutate(n_file_fact=factor(n_files, levels=c(1, 3, 10, 30, 100), 
-                        labels=c("One file", "3 files", "10 files", 
-                                "30 files", "100 files")))
-
-multifile_data %>%
-  select(method, n_files, time, type) %>%
-  mutate(n_files=log10(n_files), time=log10(time)) %>%
-  nest(data=c(n_files, time)) %>%
-  mutate(lmfit=map(data, ~broom::tidy(lm(.x$time~.x$n_files, offset = .x$n_files)))) %>%
-  unnest(lmfit) %>%
-  filter(term!="(Intercept)")
-
-xdb_gp <- multifile_data %>%
-  ggplot(aes(x=n_files, y=time, color=method, fill=method)) +
-  geom_abline(slope = 1, intercept = -5:3) +
-  geom_smooth(method="lm", formula=y~x, position=position_dodge(width=0.2), fill="grey50") +
-  geom_point(position=position_dodge(width=0.2), color="black", pch=21) +
-  facet_wrap(~type, nrow=1) +
-  scale_y_log10() +
-  scale_x_log10() +
-  # scale_color_manual(breaks = method_names, values=method_colors, aesthetics = c("color", "fill")) +
-  scale_color_viridis_d(aesthetics = c("color", "fill"), begin = 0.2, end = 0.8) +
-  labs(x="Number of files", y="Time (seconds)", color=NULL, fill=NULL) +
+  mutate(db_sort=factor(db_sort, levels = c("ordered", "unordered"),
+                        labels = c("With index/ordering", "No index/ordering"))) %>%
+  ggplot(aes(x=size, y=time, color=type, fill=type, shape=db_sort)) +
+  geom_point(alpha=0.5) +
+  # stat_summary(fun.data = mean_se, geom = "errorbar", linewidth=1.2) +
+  stat_summary(fun = mean, geom = "line", linewidth=1.2) +
+  stat_summary(fun = mean, geom = "point", size=3, color="black") +
+  facet_wrap(~method) +
+  scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 60), labels=timelabs) +
+  scale_x_log10(labels = scales::label_bytes()) +
+  scale_color_viridis_d(option = "A", begin = 0.4, end = 0.8, aesthetics = c("fill", "color")) +
+  scale_shape_manual(breaks=c("With index/ordering", "No index/ordering"), values=c(21, 22)) +
+  labs(x="Database size", y="Time taken", color=NULL, fill=NULL, shape=NULL) +
   theme_bw() +
-  theme(legend.position="inside", legend.justification.inside = c(0.05, 0.95),
-        legend.background = element_rect(color="grey80"))
-
-xmeth_gp <- multifile_data %>%
-  ggplot() +
-  geom_boxplot(aes(x=n_file_fact, y=time, color=type)) +
-  facet_wrap(~method, nrow=1) +
-  scale_y_log10() +
-  scale_color_viridis_d(option = "A", begin = 0.4, end = 0.8) +
-  theme_bw() +
-  labs(x="Number of files", y="Time (seconds)", color="Database type") +
-  theme(legend.position="top")
-
-idx_gp + (xdb_gp/xmeth_gp + plot_layout(heights = c(2,1)))
-ggsave("figures/multifile_fig.png", width = 10.5, height=7)
+  theme(legend.position = "top", legend.direction = "vertical")
+ggsave("figures/multifile_fig.png", width = 6.5, height=3.5)
 ```

### 2025-03-07 manuscript/scripts/dbloopjoinuni.py
Moving manuscript things to branch

@@ -1,164 +1,0 @@
-
-import numpy as np
-import pandas as pd
-from mzsql import *
-import sqlite3
-import os
-import random
-import glob
-import timeit
-
-random.seed(123)
-basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
-print(basename)
-
-if(not(os.path.exists(f"{basename}.sqlite"))):
-    turn_mzml_sqlite(f"{basename}.mzML", f"{basename}.sqlite")
-    turn_mzml_duckdb(f"{basename}.mzML", f"{basename}.duckdb")
-    turn_mzml_parquet(f"{basename}.mzML", f"{basename}_pqds")
-
-conn = sqlite3.connect(f"{basename}.sqlite")
-top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
-top_masses = []
-top_rts = []
-for i in range(100):
-    top_rts.append(top_int_df["rt"].iloc[0])
-    top_mz = top_int_df["mz"].iloc[0]
-    top_masses.append(top_mz)
-    mzmin, mzmax = pmppm(top_mz, 10)
-    top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax))]
-conn.close()
-
-def get_multichrom_duckdb_loop(file, mzs, ppm):
-    conn = duckdb.connect(file)
-    all_chroms = []
-    for mz_i in mzs:
-        query = "SELECT mz, int, rt FROM MS1 WHERE mz BETWEEN ? AND ?"
-        query_data = conn.execute(query, pmppm(mz_i, ppm)).fetchdf()
-        query_data["mz_i"] = mz_i
-        all_chroms.append(query_data)
-    conn.close()
-    return(pd.concat(all_chroms, ignore_index=True))
-
-def get_multichrom_duckdb_join(file, mzs, ppm):
-    mz_df = pd.DataFrame(mzs, columns=["mz_i"])
-    mz_df["mzmin"], mz_df["mzmax"] = zip(*mz_df["mz_i"].apply(lambda mz_i: pmppm(mz_i, 5)))
-    conn = duckdb.connect(file)
-    query = """
-        SELECT mz_df.mz_i, MS1.*
-        FROM mz_df
-        JOIN MS1 ON MS1.mz BETWEEN mz_df.mzmin AND mz_df.mzmax
-    """
[…]

### 2025-03-07 manuscript/scripts/multichrom_timing.py

@@ -1,259 +1,0 @@
-
-import numpy as np
-import pandas as pd
-from mzsql import *
-import sqlite3
-import os
-import random
-import glob
-import timeit
-
-random.seed(123)
-basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
-print(basename)
-
-# Create databases from mzML files if they don't already exist
-if(not(os.path.exists(f"{basename}.sqlite"))):
-    turn_mzml_sqlite(f"{basename}.mzML", f"{basename}.sqlite")
-    turn_mzml_duckdb(f"{basename}.mzML", f"{basename}.duckdb")
-    turn_mzml_parquet(f"{basename}.mzML", f"{basename}_pqds")
-
-conn = sqlite3.connect(f"{basename}.sqlite")
-top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
-top_masses = []
-top_rts = []
-for i in range(100):
-    top_rts.append(top_int_df["rt"].iloc[0])
-    top_mz = top_int_df["mz"].iloc[0]
-    top_masses.append(top_mz)
-    mzmin, mzmax = pmppm(top_mz, 10)
-    top_int_df = top_int_df[((top_int_df["mz"]<mzmin) | (top_int_df["mz"]>mzmax))]
-conn.close()
-
-def get_multichrom_mzml_pyteomics(file, mzs, ppm):
-    file_data = pyteomics.mzml.MzML(file)
-    scan_dfs = []
-    for spectrum in file_data:
-        if spectrum['ms level'] == 1:
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            mz_vals = np.array(spectrum['m/z array'])
-            int_vals = np.array(spectrum['intensity array'])
-            for mz_i in mzs:
-                mzmin, mzmax = pmppm(mz_i, ppm)
-                bet_idxs = (mz_vals > mzmin) & (mz_vals < mzmax)
-                if np.any(bet_idxs):
-                    scan_dfs.append(pd.DataFrame({
-                        'mz_i': mz_i,
-                        'mz': mz_vals[bet_idxs],
-                        'int': int_vals[bet_idxs],
-                        'rt': rt_val
-                    }))
[…]

### 2025-03-07 manuscript/scripts/multifile_sizetimes.py

@@ -1,111 +1,0 @@
-import numpy as np
-import pandas as pd
-import sqlite3
-import os
-import glob
-import random
-import timeit
-from mzsql import *
-
-random.seed(123)
-multifile_subset = random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 100)
-multifile_subset = [f.replace("\\", "/") for f in multifile_subset]
-
-file_sequence = [1, 3, 10, 30, 100]
-
-for n_files in file_sequence:
-    for db_sort in ["ordered", "unordered"]:
-        ordering = "mz" if db_sort == "ordered" else None
-        for db_name in ["duckdb", "sqlite", "parquet"]:
-            db_ending = {"duckdb": ".duckdb", "sqlite": ".sqlite", "parquet": "_pqds"}
-            file_ending = "_ord" if db_sort == "ordered" else ""
-            file_ending += db_ending[db_name]
-            for db_type in ["consolidated", "multifile"]:
-                this_multifile_subset = multifile_subset[:n_files]
-                if db_type == "multifile":
-                    if n_files < max(file_sequence):
-                        continue
-                    for mzml_i in this_multifile_subset:
-                        output_db = mzml_i.replace(".mzML", file_ending)
-                        print(output_db)
-                        combined_cmd = f"turn_mzml_{db_name}(mzml_i, output_db, ordered=ordering)"
-                        #eval(combined_cmd)
-                else:
-                    output_db = f"E:/mzsql/MTBLS10066/consolidated_{n_files}{file_ending}"
-                    print(output_db)
-                    combined_cmd = f"turn_mzml_{db_name}(this_multifile_subset, output_db, ordered=ordering)"
-                    #eval(combined_cmd)
-
-conn = sqlite3.connect(f"E:/mzsql/MTBLS10066/consolidated_{max(file_sequence)}.sqlite")
-top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
-top_masses = []
-top_rts = []
-for i in range(10):
-    top_rts.append(top_int_df["rt"].iloc[0])
-    top_mz = top_int_df["mz"].iloc[0]
-    top_masses.append(top_mz)
-    mzmin, mzmax = pmppm(top_mz, 10)
-    rtmin, rtmax = top_int_df["rt"].iloc[0]+(-1, 1)
[…]

### 2025-03-07 manuscript/scripts/singlefile_supp.py

@@ -1,234 +1,0 @@
-
-import pandas as pd
-import numpy as np
-import timeit
-import glob
-import random
-import pyopenms
-import pyteomics.mzml
-import pyteomics.mzmlb
-import pymzml
-import h5py
-import matplotlib.pyplot as plt
-from mzsql import *
-
-random.seed(123)
-basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
-print(basename)
-
-def get_randscan_mzml_pyopenms(basename, scan_num):
-    exp = pyopenms.MSExperiment()
-    pyopenms.MzMLFile().load(f"{basename}.mzML", exp)
-    spec_data = exp[scan_num].get_peaks()
-    return(pd.DataFrame({"mz":spec_data[0], "int":spec_data[1]}))
-
-def get_randscan_mzml_pyteomics(basename, scan_num):
-    file_data = pyteomics.mzml.MzML(f"{basename}.mzML")
-    return(pd.DataFrame({"mz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
-
-def get_randscan_mzml_pymzml(basename, scan_num):
-    run = pymzml.run.Reader(f"{basename}.mzML", build_index_from_scratch=True)
-    spec1_data = run[scan_num].peaks("raw")
-    return(pd.DataFrame({"mz":spec1_data[:,0], "int":spec1_data[:,1]}))
-
-def get_randscan_mzmlb(basename, scan_num):
-    file_data = pyteomics.mzmlb.MzMLb(f"{basename}.mzMLb")
-    return(pd.DataFrame({"mz":file_data[scan_num]['m/z array'], "int":file_data[scan_num]['intensity array']}))
-
-def get_randscan_mza(basename, scan_num):
-    mza = h5py.File(f"{basename}.mza", 'r')
-    intensities = mza[f"Arrays_intensity/{scan_num}"][:]
-    mz = mza[f"Arrays_mz/{scan_num}"][:]
-    mza.close()
-    spec_df = pd.DataFrame({"mz":mz, "int":intensities})
-    return(spec_df)
-
-def get_randscan_mz5(basename, scan_num):
-    mz5_file = h5py.File(f"{basename}.mz5", 'r')
-    scan_idxs = np.concatenate(([0], mz5_file["SpectrumIndex"][...]))
[…]

### 2025-03-07 manuscript/scripts/singlefile_timing.py

@@ -1,180 +1,0 @@
-
-# Setup things
-import numpy as np
-import pandas as pd
-import sqlite3
-import os
-import glob
-import random
-import timeit
-from mzsql import *
-
-random.seed(123)
-basename=random.sample(glob.glob("E:/mzsql/MTBLS10066/*.mzML"), 1)[0].replace(".mzML", "").replace("\\", "/")
-print(basename)
-# 20220923_LEAP-POS_QC04
-
-# Create databases from mzML files if they don't already exist
-if(not(os.path.exists(f"{basename}.sqlite"))):
-    turn_mzml_sqlite(f"{basename}.mzML", f"{basename}.sqlite")
-    turn_mzml_duckdb(f"{basename}.mzML", f"{basename}.duckdb")
-    turn_mzml_parquet(f"{basename}.mzML", f"{basename}_pqds")
-
-# # Identify the bits of information to be extracted
-conn = sqlite3.connect(f"{basename}.sqlite")
-# # Get 10 random spectra
-# rand_scan_query = """
-# SELECT id 
-# FROM (SELECT DISTINCT id FROM {0}) 
-# ORDER BY RANDOM() 
-# LIMIT 10
-# """
-# rand_MS1_scans = pd.read_sql_query(rand_scan_query.format("MS1"), conn)["id"].tolist()
-# rand_MS2_scans = pd.read_sql_query(rand_scan_query.format("MS2"), conn)["id"].tolist()
-# print(rand_MS1_scans)
-# print(rand_MS2_scans)
-# Manually specify scans here since SQLite doesn't handle seed setting very well:
-rand_MS1_scans = (9044, 187, 7532, 2396, 10409, 87, 5259, 69, 5157, 7988)
-rand_MS2_scans = (3778, 3815, 5685, 6859, 4223, 3372, 6948, 6354, 3140, 912)
-
-# Get the top 10ish masses by intensity and their retention times
-top_int_df = pd.read_sql_query("SELECT * FROM MS1 ORDER BY int DESC LIMIT 30000", conn)
-top_masses = []
-top_rts = []
-for i in range(10):
-    top_rts.append(top_int_df["rt"].iloc[0])
-    top_mz = top_int_df["mz"].iloc[0]
-    top_masses.append(top_mz)
-    mzmin, mzmax = pmppm(top_mz, 10)
[…]

### 2025-03-07 src/mzsql/__init__.py
Removing package things from repo

@@ -1,12 +1,0 @@
-
-from .helpers import pmppm
-from .mzml_functions import *
-from .mzmlb_functions import *
-from .sqlite_functions import *
-from .parquet_functions import *
-from .duckdb_functions import *
-from .mz5_functions import *
-from .mza_functions import *
-from .mzTree_functions import *
-from .mzMD_functions import *
-from .mzdb_functions import *

### 2025-03-07 src/mzsql/duckdb_functions.py

@@ -1,105 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import os
-import duckdb
-import pyteomics.mzml
-from .helpers import pmppm
-
-def turn_mzml_duckdb(files, outfile, ordered=None):
-    conn = duckdb.connect(outfile)
-    conn.execute("DROP TABLE IF EXISTS MS1")
-    conn.execute("DROP TABLE IF EXISTS MS2")
-    empty_ms1 = pd.DataFrame(columns=["filename", "id", "mz", "int", "rt"]).astype(
-        {"filename": "string", "id": "int64", "mz": "float64", "int": "float64", "rt": "float64"}
-    )
-    conn.execute("CREATE TABLE MS1 AS SELECT * FROM empty_ms1")
-    empty_ms2 = pd.DataFrame(columns=["filename", "id", "premz", "fragmz", "int", "rt"]).astype(
-        {"filename": "string", "id": "int64", "premz": "float64", "fragmz": "float64", "int": "float64", "rt": "float64"}
-    )
-    conn.execute("CREATE TABLE MS2 AS SELECT * FROM empty_ms2")
-
-    if isinstance(files, str):
-        files = [files]
-    for file in files:
-        if not os.path.exists(file):
-            raise FileNotFoundError(f"Input mzML file '{file}' does not exist.")
-        MS1_dfs = []
-        MS2_dfs = []
-        for spectrum in pyteomics.mzml.MzML(file):
-            if spectrum['ms level'] == 1:
-                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-                mz_vals = spectrum['m/z array']
-                int_vals = spectrum['intensity array']
-                rt_val = spectrum['scanList']['scan'][0]['scan start time']
-                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 
-                                        'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                MS1_dfs.append(df_scan)
-            if spectrum["ms level"] == 2:
-                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-                premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-                mz_vals = spectrum['m/z array']
[…]

### 2025-03-07 src/mzsql/helpers.py

@@ -1,3 +1,0 @@
-
-def pmppm(mass, ppm=4):
-    return(mass*(1-ppm/1000000), mass*(1+ppm/1000000))

### 2025-03-07 src/mzsql/mz5_functions.py

@@ -1,57 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import h5py
-from .helpers import pmppm
-
-
-def get_chrom_mz5(file, mz, ppm):
-    mz5_file = h5py.File(file, 'r')
-    bounds_df = pd.DataFrame({"lower":np.concatenate(([0], mz5_file["SpectrumIndex"][...][:-1])),
-                              "upper":mz5_file["SpectrumIndex"][...]})
-    
-    mzmin, mzmax = pmppm(mz, ppm)
-    has_precursor = [len(item)>0 for item in mz5_file["SpectrumMetaData"]["precursors"]]
-    scan_dfs = []
-    for index, row in bounds_df.iterrows():
-        if(not(has_precursor[index])):
-            scan_df = pd.DataFrame({
-                "rt": mz5_file["ChomatogramTime"][index],
-                "mz": np.cumsum(mz5_file["SpectrumMZ"][row["lower"]:row["upper"]]),
-                "int": mz5_file["SpectrumIntensity"][row["lower"]:row["upper"]]
-            })
-            bet_df = scan_df[(scan_df["mz"]>mzmin) & (scan_df["mz"]<mzmax)]
-            scan_dfs.append(bet_df)
-    file_df = pd.concat(scan_dfs, ignore_index=True)
-    mz5_file.close()
-    return(file_df)
-
-def get_spec_mz5(file, scan_num):
-    mz5_file = h5py.File(file, 'r')
-    mz5_meta = mz5_file["SpectrumMetaData"]["id", "index"]
-    idx = np.where(np.char.find(mz5_meta['id'].astype(str), f'scan={scan_num}') != -1)[0][0]
-    scan_idxs = np.concatenate(([0], mz5_file["SpectrumIndex"][...]))
-    lower_bound = scan_idxs[idx-1]
-    upper_bound = scan_idxs[idx]
-    mz_vals = np.cumsum(mz5_file["SpectrumMZ"][lower_bound:upper_bound])
-    int_vals = mz5_file["SpectrumIntensity"][lower_bound:upper_bound]
-    return(pd.DataFrame({"mz":mz_vals, "int":int_vals}))
-
-def get_rtrange_mz5(file, rtstart, rtend):
-    mz5_file = h5py.File(file, 'r')
-    scan_idxs = np.concatenate(([0], mz5_file["SpectrumIndex"][...]))
-    scan_dfs = []
[…]

### 2025-03-07 src/mzsql/mzMD_functions.py

@@ -1,67 +1,0 @@
-import requests
-from requests.exceptions import ConnectionError
-import pandas as pd
-from .helpers import pmppm
-
-def get_chrom_mzMD(url_port, mz, ppm):
-    """
-    Retrieves chromatogram data from a server for a specific m/z range.
-
-    Args:
-        url_port (str): The URL and port of the server, e.g., 'http://localhost:5000'.
-        mz (float): The target m/z value.
-        ppm (float): The mass tolerance in parts per million (ppm).
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time (rt), m/z, and intensity data 
-                      within the specified m/z range.
-    """
-    mz_min, mz_max = pmppm(mz, ppm)
-    request_string = f"{url_port}/api/v2/getpoints?mzmin={mz_min}&mzmax={mz_max}&rtmin=0&rtmax=1000000&n=0&m=0"
-    
-    try:
-        response = requests.get(request_string)
-        response.raise_for_status()
-        if response.status_code == 204:
-            print("No content returned. Have you loaded a file?")
-    except ConnectionError as e:
-        if "Connection refused" in str(e):
-            print("Connection refused. Have you started the server?")
-        else:
-            print(f"Connection error: {e}")
-
-    chrom_data = pd.DataFrame(response.json(), columns=["pointId", "traceId", "mz", "rt", "intensity"])
-    chrom_data = chrom_data[["rt", "mz", "intensity"]].sort_values(by="rt")
-    chrom_data.columns = ["rt", "mz", "int"]
-    return chrom_data
-
-def get_rtrange_mzMD(url_port, rtstart, rtend):
-    """
-    Retrieves chromatogram data from a server for a specific retention time range.
-
-    Args:
-        url_port (str): The URL and port of the server, e.g., 'http://localhost:5000'.
-        rtstart (float): The start retention time value.
-        rtend (float): The end retention time value.
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time (rt), m/z, and intensity data 
-                      within the specified retention time range.
-    """
-    request_string = f"{url_port}/api/v2/getpoints?mzmin=0&mzmax=1000000&rtmin={rtstart}&rtmax={rtend}&n=0&m=0"
-    
-    try:
-        response = requests.get(request_string)
-        response.raise_for_status()
-        if response.status_code == 204:
-            print("No content returned. Have you loaded a file?")
[…]

### 2025-03-07 src/mzsql/mzTree_functions.py

@@ -1,56 +1,0 @@
-import requests
-from requests.exceptions import ConnectionError
-import pandas as pd
-from .helpers import pmppm
-
-def get_chrom_mztree(url_port, mz, ppm):
-    """
-    Fetches chromatogram data from a specified server endpoint within a given m/z range 
-    and processes it into a pandas DataFrame.
-
-    Args:
-        url_port (str): The URL and port where the server is running.
-        mz (float): The target m/z value to extract chromatogram data around.
-        ppm (float): The parts per million (ppm) tolerance to define the m/z range.
-
-    Returns:
-        pd.DataFrame: A DataFrame containing the chromatogram data with columns 
-                      ["rt", "mz", "intensity"], sorted by retention time (rt).
-    """
-    mz_min, mz_max = pmppm(mz, ppm)
-    request_string = f"{url_port}/api/v2/getpoints?mzmin={mz_min}&mzmax={mz_max}&rtmin=0&rtmax=1000000&numpoints=0"
-    
-    try:
-        response = requests.get(request_string)
-        response.raise_for_status()
-        if response.status_code == 204:
-            print("No content returned. Have you loaded a file?")
-    except ConnectionError as e:
-        if "Connection refused" in str(e):
-            print("Connection refused. Have you started the server?")
-        else:
-            print(f"Connection error: {e}")
-
-    chrom_data = pd.DataFrame(response.json(), columns=["pointId", "traceId", "mz", "rt", "intensity"])
-    chrom_data = chrom_data[["rt", "mz", "intensity"]].sort_values(by="rt")
-    chrom_data.columns = ["rt", "mz", "int"]
-    return chrom_data
-
-def get_rtrange_mztree(url_port, rtstart, rtend):
-    request_string = f"{url_port}/api/v2/getpoints?mzmin=0&mzmax=1000000&rtmin={rtstart}&rtmax={rtend}&numpoints=0"
-    
-    try:
-        response = requests.get(request_string)
-        response.raise_for_status()
-        if response.status_code == 204:
-            print("No content returned. Have you loaded a file?")
-    except ConnectionError as e:
-        if "Connection refused" in str(e):
-            print("Connection refused. Have you started the server?")
-        else:
-            print(f"Connection error: {e}")
-
-    chrom_data = pd.DataFrame(response.json(), columns=["pointId", "traceId", "mz", "rt", "intensity"])
[…]

### 2025-03-07 src/mzsql/mza_functions.py

@@ -1,127 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import h5py
-import mzapy
-from .helpers import pmppm
-
-def get_chrom_mza(file, mz, ppm):
-    """
-    Retrieves chromatogram data from an mzapy MZA file for a specified m/z range.
-
-    Args:
-        file (str): Path to the mzapy file to be processed.
-        mz (float): The target m/z value.
-        ppm (float): The allowed mass tolerance in parts per million (ppm).
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time and intensity data 
-                      for the specified m/z range.
-    """
-    mzmin, mzmax = pmppm(mz, ppm)
-    mza=mzapy.MZA(file)
-    xic_rt, xic_int = mza.collect_xic_arrays_by_mz(mzmin, mzmax)
-    mza.close()
-    chrom_data = pd.DataFrame({"rt":xic_rt, "int":xic_int})
-    return(chrom_data)
-    
-def get_rtrange_mza(file, rtstart, rtend):
-    """
-    Retrieves spectrum data within a specified retention time range from an mzapy MZA file.
-
-    Args:
-        file (str): Path to the mzapy file.
-        rtstart (float): The start retention time value.
-        rtend (float): The end retention time value.
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time, m/z, and intensity values for 
-                      spectra within the specified retention time range.
-    """
-    mza=mzapy.MZA(file)
-    rtrange_data = mza.collect_ms1_df_by_rt(rtstart, rtend)
-    rtrange_data = rtrange_data[["rt", "mz", "intensity"]]
-    rtrange_data.columns = ["rt", "mz", "int"]
-    mza.close()
-    return(rtrange_data)
-    
-def get_spec_mza(file, spectrum_idx):
-    """
-    Retrieves spectrum data for a specific scan index from an MZA file.
-
-    Args:
-        file (str): Path to the MZA file.
-        spectrum_idx (int): The scan index of the spectrum to retrieve.
-
-    Returns:
-        pd.DataFrame: A DataFrame containing m/z and intensity values for the specified spectrum.
-    """
-    mza = h5py.File(file, 'r')
[…]

### 2025-03-07 src/mzsql/mzdb_functions.py

@@ -1,250 +1,0 @@
-
-import sqlite3
-import pandas as pd
-import numpy as np
-import struct
-from .helpers import pmppm
-
-
-def unpack_raw_bb(raw_string):
-    """
-    Unpacks raw bounding box (BB) data from a binary string and converts it into a DataFrame.
-
-    Args:
-        raw_string (bytes): A raw binary string containing peak data for multiple scans.
-
-    Returns:
-        pd.DataFrame: A DataFrame with columns ['mz', 'int', 'scan_id'] representing 
-                      m/z, intensity, and scan ID for each peak.
-    """
-    offset = 0
-    scanids = []
-    offsets = []
-    nb_peaks_all = []
-    while offset < len(raw_string):
-        # Unpack two integers starting at a byte offset of zero
-        scan_id, nb_peaks = struct.Struct('<2i').unpack_from(raw_string, offset)
-        scanids.append(scan_id)
-        nb_peaks_all.append(nb_peaks)
-        offsets.append(offset)
-        # New offset is 8 bytes (metadata) + 12 * number of peaks (8 for mz and 4 for int)
-        offset+=8+12*nb_peaks
-    
-    scan_data = []
-    for i in range(len(scanids)):
-        # Encoded as little endian, then double (mz) and float (int) for each peak
-        decomp_string = '<'+'df'*nb_peaks_all[i]
-        # Unpack the data, remembering to skip the 8 bytes of metadata at the front!
-        data_tuple = struct.Struct(decomp_string).unpack_from(raw_string, offsets[i]+8)
-        mz_int_array = np.array(data_tuple).reshape(-1, 2)
-        mz_int_scanid_array = np.column_stack((mz_int_array, np.full(mz_int_array.shape[0], scanids[i])))
-        # Getting duplicate entries for some reason... using unique() to remove them for now
-        scan_data.append(np.unique(mz_int_scanid_array, axis=0))
-    
-    return(pd.DataFrame(np.vstack(scan_data), columns=['mz', 'int', 'scan_id']))
-
-
-def get_chrom_mzdb(file, mz, ppm):
-    """
-    Retrieves chromatogram data from an mzDB file for a specified m/z range.
-
-    Args:
[…]

### 2025-03-07 src/mzsql/mzml_functions.py

@@ -1,379 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import pyteomics.mzml
-import pyopenms
-import pymzml
-from .helpers import pmppm
-
-# pyteomics things
-def get_chrom_mzml_pyteomics(file, mz, ppm):
-    """
-    Retrieves chromatogram data from a mzML file using pyteomics for a specific m/z range.
-
-    Args:
-        file (str): Path to the mzML file.
-        mz (float): The target m/z value.
-        ppm (float): The mass tolerance in parts per million (ppm).
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time (rt), m/z, and intensity data 
-                      within the specified m/z range.
-    """
-    mzmin, mzmax = pmppm(mz, ppm)
-    scan_dfs = []
-    for spectrum in pyteomics.mzml.MzML(file):
-        if spectrum['ms level'] == 1:
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            mz_vals=spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-            bet_idxs = (mzmin < spectrum["m/z array"]) & (spectrum["m/z array"] < mzmax)
-            if(sum(bet_idxs)>0):
-                df_scan = pd.DataFrame({'mz':mz_vals[bet_idxs], 'int':int_vals[bet_idxs], 'rt':[rt_val]*sum(bet_idxs)})
-                scan_dfs.append(df_scan)    
-    return(pd.concat(scan_dfs, ignore_index=True))
-    
-def get_spec_mzml_pyteomics(file, scan_num):
-    file_data = pyteomics.mzml.MzML(file)
-    i = 0
-    while i <= len(file_data):
-        if(int(file_data[i]['id'].split("scan=")[-1].split()[0]) == scan_num):
-            return(pd.DataFrame({"mz":file_data[i]['m/z array'], "int":file_data[i]['intensity array']}))
-        else:
-            i += 1
-    raise Exception(f"No scan number {scan_num} found")
-
-
-def get_rtrange_mzml_pyteomics(file, rtstart, rtend):
-    """
-    Retrieves chromatogram data for a specific retention time range from a mzML file using pyteomics.
-
-    Args:
-        file (str): Path to the mzML file.
-        rtstart (float): The start retention time (in minutes).
[…]

### 2025-03-07 src/mzsql/mzmlb_functions.py

@@ -1,109 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import pyteomics.mzmlb
-from .helpers import pmppm
-
-def get_chrom_mzmlb(file, mz, ppm):
-    """
-    Retrieves chromatogram data from an mzMLb file for a specific m/z range.
-
-    Args:
-        file (str): Path to the mzMLb file.
-        mz (float): The target m/z value.
-        ppm (float): The mass tolerance in parts per million (ppm).
-
-    Returns:
-        pd.DataFrame: A DataFrame containing retention time (rt), m/z, and intensity data 
-                      within the specified m/z range.
-    """
-    scan_dfs = []
-    for spectrum in pyteomics.mzmlb.MzMLb(file):
-        if spectrum['ms level'] == 1:
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            mz_vals=spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-
-            mzmin, mzmax = pmppm(mz, ppm)
-            bet_idxs = (mzmin < spectrum["m/z array"]) & (spectrum["m/z array"] < mzmax)
-            if(sum(bet_idxs)>0):
-                df_scan = pd.DataFrame({'mz':mz_vals[bet_idxs], 'int':int_vals[bet_idxs], 'rt':[rt_val]*sum(bet_idxs)})
-                scan_dfs.append(df_scan)    
-    return(pd.concat(scan_dfs, ignore_index=True))
-
-def get_spec_mzmlb(file, scan_num):
-    file_data = pyteomics.mzmlb.MzMLb(file)
-    i = 0
-    while i <= len(file_data):
-        if(int(file_data[i]['id'].split("scan=")[-1].split()[0]) == scan_num):
-            return(pd.DataFrame({"mz":file_data[i]['m/z array'], "int":file_data[i]['intensity array']}))
-        else:
-            i += 1
-    raise Exception(f"No scan number {scan_num} found")
-
-
-def get_rtrange_mzmlb(file, rtstart, rtend):
-    """
-    Retrieves chromatogram data from an mzMLb file for a specific retention time range.
-
-    Args:
-        file (str): Path to the mzMLb file.
-        rtstart (float): The start retention time (in minutes).
-        rtend (float): The end retention time (in minutes).
-
-    Returns:
[…]

### 2025-03-07 src/mzsql/parquet_functions.py

@@ -1,101 +1,0 @@
-
-import numpy as np
-import pandas as pd
-import pyteomics.mzml
-from .helpers import pmppm
-import pyarrow as pa
-import pyarrow.parquet as pq
-import pyarrow.compute as pc
-import pyarrow.dataset as ds
-import os
-
-def turn_mzml_parquet(files, outdir, ordered=None):
-    if isinstance(files, str):
-        files = [files]
-    for file in files:
-        MS1_dfs = []
-        MS2_dfs = []
-        for spectrum in pyteomics.mzml.MzML(file):
-            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-            mz_vals = spectrum['m/z array']
-            int_vals = spectrum['intensity array']
-            rt_val = spectrum['scanList']['scan'][0]['scan start time']
-            if spectrum['ms level'] == 1:
-                df_scan = pd.DataFrame({'id': idx, 'mslevel':"MS1", 'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                MS1_dfs.append(df_scan)
-            if spectrum["ms level"] == 2:
-                premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-                df_scan = pd.DataFrame({'id': idx, 'mslevel':"MS2", 'premz': premz_val, 'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                MS2_dfs.append(df_scan)
-    
-        all_MS1 = pd.concat(MS1_dfs, ignore_index=True)
-        all_MS2 = pd.concat(MS2_dfs, ignore_index=True)
-        if ordered is not None:
-            if ordered == "rt":
-                all_MS1.sort_values(by=ordered, inplace=True)
-                all_MS2.sort_values(by=ordered, inplace=True)
-            if ordered == "mz":
-                all_MS1.sort_values(by=ordered, inplace=True)
-            if ordered == "fragmz":
-                all_MS2.sort_values(by=ordered, inplace=True)
-            if ordered == "premz":
-                all_MS2.sort_values(by=ordered, inplace=True)
-    
[…]

### 2025-03-07 src/mzsql/sqlite_functions.py

@@ -1,159 +1,0 @@
-import sqlite3
-import pandas as pd
-import pyteomics.mzml
-import os
-from .helpers import pmppm
-
-def turn_mzml_sqlite(files, outfile, ordered=None):
-    conn = sqlite3.connect(outfile)
-    conn.execute("DROP TABLE IF EXISTS MS1")
-    conn.execute("DROP TABLE IF EXISTS MS2")
-    if isinstance(files, str):
-        files = [files]
-    for file in files:
-        MS1_dfs = []
-        MS2_dfs = []
-        for spectrum in pyteomics.mzml.MzML(file):
-            if spectrum['ms level'] == 1:
-                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-                mz_vals = spectrum['m/z array']
-                int_vals = spectrum['intensity array']
-                rt_val = spectrum['scanList']['scan'][0]['scan start time']
-                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'mz': mz_vals, 
-                                        'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                MS1_dfs.append(df_scan)
-            if spectrum['ms level'] == 2:
-                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
-                mz_vals = spectrum['m/z array']
-                int_vals = spectrum['intensity array']
-                rt_val = spectrum['scanList']['scan'][0]['scan start time']
-                premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
-                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'premz': premz_val, 
-                                        'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
-                MS2_dfs.append(df_scan)
-        pd.concat(MS1_dfs, ignore_index=True).to_sql("MS1", conn, if_exists="append", index=False)
-        pd.concat(MS2_dfs, ignore_index=True).to_sql("MS2", conn, if_exists="append", index=False)
-    if ordered is not None:
-        index_name = f"idx_{ordered}"
-        if ordered == "rt":
-            conn.execute(f"CREATE INDEX IF NOT EXISTS {index_name} ON MS1 ({ordered})")
[…]

### 2025-03-07 tests/test_duckdb_funs.py
Removing package tests

@@ -1,45 +1,0 @@
-import numpy as np
-import pandas as pd
-import pytest
-from mzsql import *
-
-
-# Smoke test DuckDB creation
-def test_turn_mzml_duckdb():
-    turn_mzml_duckdb("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", "demo_data/180205_Poo_TruePoo_Full1.duckdb")
-
-
-# Tests for DuckDB file type
-def test_get_chrom_duckdb():
-    ref_data = get_chrom_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 118.0865, 10)
-    test_data = get_chrom_duckdb("demo_data/180205_Poo_TruePoo_Full1.duckdb", 118.0865, 10)
-    
-    # Align indices before comparison
-    ref_data, test_data = ref_data.align(test_data, join='inner')
-    
-    assert (ref_data["rt"] == test_data["rt"]).all()
-    assert (ref_data["mz"] == test_data["mz"]).all()
-    assert (ref_data["int"] == test_data["int"]).all()
-
-def test_get_spec_duckdb():
-    spec_data_mzml = get_spec_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 1)
-    spec_data_duckdb = get_spec_duckdb("demo_data/180205_Poo_TruePoo_Full1.duckdb", 1)
-    
-    # Align indices before comparison
-    spec_data_mzml, spec_data_duckdb = spec_data_mzml.align(spec_data_duckdb, join='inner')
-    
-    assert (spec_data_mzml["mz"] == spec_data_duckdb["mz"]).all()
-    assert (spec_data_mzml["int"] == spec_data_duckdb["int"]).all()
-
-def test_get_rtrange_duckdb():
-    rtrange_data_mzml = get_rtrange_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 6.5, 8)
-    rtrange_data_duckdb = get_rtrange_duckdb("demo_data/180205_Poo_TruePoo_Full1.duckdb", 6.5, 8)
-    
-    # Align indices before comparison
-    rtrange_data_mzml, rtrange_data_duckdb = rtrange_data_mzml.align(rtrange_data_duckdb, join='inner')
-
-    assert (rtrange_data_mzml["rt"] == rtrange_data_duckdb["rt"]).all()
[…]

### 2025-03-07 tests/test_mz5_funs.py

@@ -1,28 +1,0 @@
-import numpy as np
-import pandas as pd
-import pytest
-from mzsql import *
-
-# Tests for mz5 file type
-def test_get_chrom_mz5():
-    ref_data = get_chrom_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 118.0865, 10)
-    test_data = get_chrom_mz5("demo_data/180205_Poo_TruePoo_Full1.mz5", 118.0865, 10)
-    ref_data, test_data = ref_data.align(test_data, join='inner')
-    assert (ref_data["rt"] == test_data["rt"]).all()
-    assert (ref_data["mz"] == test_data["mz"]).all()
-    assert (ref_data["int"] == test_data["int"]).all()
-
-def test_get_spec_mz5():
-    spec_data_mzml = get_spec_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 1)
-    spec_data_mz5 = get_spec_mz5("demo_data/180205_Poo_TruePoo_Full1.mz5", 1)
-    spec_data_mzml, spec_data_mz5 = spec_data_mzml.align(spec_data_mz5, join='inner')
-    assert (spec_data_mzml["mz"] == spec_data_mz5["mz"]).all()
-    assert (spec_data_mzml["int"] == spec_data_mz5["int"]).all()
-
-def test_get_rtrange_mz5():
-    rtrange_data_mzml = get_rtrange_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 6.5, 8)
-    rtrange_data_mz5 = get_rtrange_mz5("demo_data/180205_Poo_TruePoo_Full1.mz5", 6.5, 8)
-    rtrange_data_mzml, rtrange_data_mz5 = rtrange_data_mzml.align(rtrange_data_mz5, join='inner')
-    assert (rtrange_data_mzml["rt"] == rtrange_data_mz5["rt"]).all()
-    assert (rtrange_data_mzml["mz"] == rtrange_data_mz5["mz"]).all()
-    assert (rtrange_data_mzml["int"] == rtrange_data_mz5["int"]).all()

### 2025-03-07 tests/test_mzML_funs.py

@@ -1,159 +1,0 @@
-import numpy as np
-import pandas as pd
-from pyteomics import mzml, mzmlb
-import pyopenms
-import pytest
-import pymzml
-from mzsql import *
-
-def test_gcm_pyteomics():
-    chrom_data = get_chrom_mzml_pyteomics("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    assert chrom_data.shape == (1359, 3)
-    assert min(chrom_data["mz"]) >= pmppm(118.0865, 10)[0]
-    assert max(chrom_data["mz"]) <= pmppm(118.0865, 10)[1]
-    assert max(chrom_data["int"]) < 4e8
-    assert min(chrom_data["rt"] > 0)
-    assert max(chrom_data["rt"] < 25)
-
-def test_gcm_pymzml():
-    chrom_data = get_chrom_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    assert chrom_data.shape == (1359, 3)
-    assert min(chrom_data["mz"]) >= pmppm(118.0865, 10)[0]
-    assert max(chrom_data["mz"]) <= pmppm(118.0865, 10)[1]
-    assert max(chrom_data["int"]) < 4e8
-    assert min(chrom_data["rt"] > 0)
-    assert max(chrom_data["rt"] < 25)
-
-def test_gcm_pyopenms():
-    chrom_data = get_chrom_mzml_pyopenms("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    assert chrom_data.shape == (1359, 3)
-    assert min(chrom_data["mz"]) >= pmppm(118.0865, 10)[0]
-    assert max(chrom_data["mz"]) <= pmppm(118.0865, 10)[1]
-    assert max(chrom_data["int"]) < 4e8
-    assert min(chrom_data["rt"] > 0)
-    assert max(chrom_data["rt"] < 25)
-
-def test_gcm_pyopenms_2d():
-    chrom_data = get_chrom_mzml_pyopenms_2DPeak("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    assert chrom_data.shape == (1359, 3)
-    assert min(chrom_data["mz"]) >= pmppm(118.0865, 10)[0]
-    assert max(chrom_data["mz"]) <= pmppm(118.0865, 10)[1]
[…]

### 2025-03-07 tests/test_mza_funs.py

@@ -1,38 +1,0 @@
-import numpy as np
-import pandas as pd
-import pytest
-from mzsql import *
-
-# Tests for MZA file type
-def test_get_chrom_mza():
-    ref_data = get_chrom_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 118.0865, 10)
-    test_data = get_chrom_mza("demo_data/180205_Poo_TruePoo_Full1.mza", 118.0865, 10)
-
-    # Merge data on 'rt' column
-    merged_data = pd.merge(ref_data, test_data, on="rt", suffixes=("_ref", "_test"))
-
-    assert (merged_data["mz_ref"] == merged_data["mz_test"]).all()
-    assert (merged_data["int_ref"] == merged_data["int_test"]).all()
-
-
-def test_get_spec_mza():
-    spec_data_mzml = get_spec_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 1)
-    spec_data_mza = get_spec_mza("demo_data/180205_Poo_TruePoo_Full1.mza", 1)
-
-    # Align the data for comparison
-    spec_data_mza, spec_data_mzml = spec_data_mza.align(spec_data_mzml, join='inner')
-
-    assert (spec_data_mzml["mz"] == spec_data_mza["mz"]).all()
-    assert (spec_data_mzml["int"] == spec_data_mza["int"]).all()
-
-def test_get_rtrange_mza():
-    rtrange_data_mzml = get_rtrange_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 6.5, 8)
-    rtrange_data_mza = get_rtrange_mza("demo_data/180205_Poo_TruePoo_Full1.mza", 6.5, 8)
-
-    # Align the data for comparison
-    rtrange_data_mza, rtrange_data_mzml = rtrange_data_mza.align(rtrange_data_mzml, join='inner')
-
-    assert (rtrange_data_mzml["rt"] == rtrange_data_mza["rt"]).all()
-    assert (rtrange_data_mzml["mz"] == rtrange_data_mza["mz"]).all()
-    assert (rtrange_data_mzml["int"] == rtrange_data_mza["int"]).all()
-

### 2025-03-07 tests/test_mzdb_functions.py

@@ -1,24 +1,0 @@
-import numpy as np
-import pandas as pd
-import pytest
-from mzsql import *
-
-def test_get_chrom_mzdb():
-    ref_data = get_chrom_mzml_pyteomics("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    test_data = get_chrom_mzdb("demo_data/180205_Poo_TruePoo_Full1.raw.mzDB", 118.0865, 10)
-    assert np.allclose(ref_data["rt"], test_data["rt"])
-    assert np.allclose(ref_data["mz"], test_data["mz"])
-    assert np.allclose(ref_data["int"], test_data["int"])
-
-def test_get_spec_mzdb():
-    spec_data_mzml = get_spec_mzml_pyteomics("demo_data/180205_Poo_TruePoo_Full1.mzML", 1)
-    spec_data_mzdb = get_spec_mzdb("demo_data/180205_Poo_TruePoo_Full1.raw.mzDB", 1)
-    assert np.allclose(spec_data_mzml["mz"], spec_data_mzdb["mz"])
-    assert np.allclose(spec_data_mzml["int"], spec_data_mzdb["int"])
-
-def test_get_rtrange_mzdb():
-    rtrange_data_mzml = get_rtrange_mzml_pyteomics("demo_data/180205_Poo_TruePoo_Full1.mzML", 6.5, 8)
-    rtrange_data_mzdb = get_rtrange_mzdb("demo_data/180205_Poo_TruePoo_Full1.raw.mzDB", 6.5, 8)
-    assert np.allclose(rtrange_data_mzml["rt"], rtrange_data_mzdb["rt"])
-    assert np.allclose(rtrange_data_mzml["mz"], rtrange_data_mzdb["mz"])
-    assert np.allclose(rtrange_data_mzml["int"], rtrange_data_mzdb["int"])

### 2025-03-07 tests/test_mzmlb_funs.py

@@ -1,36 +1,0 @@
-import numpy as np
-import pandas as pd
-import pytest
-from mzsql import *
-
-# Tests for mzMLB file type
-def test_get_chrom_mzmlb():
-    ref_data = get_chrom_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    test_data = get_chrom_mzmlb("demo_data/180205_Poo_TruePoo_Full1.mzMLb", 118.0865, 10)
-
-    # Merge data on 'rt' column
-    merged_data = pd.merge(ref_data, test_data, on="rt", suffixes=("_ref", "_test"))
-
-    assert (merged_data["mz_ref"] == merged_data["mz_test"]).all()
-    assert (merged_data["int_ref"] == merged_data["int_test"]).all()
-
-#expected failure
-def test_get_spec_mzmlb():
-    spec_data_mzml = get_spec_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1.mzML", 1)
-    spec_data_mzmlb = get_spec_mzmlb("demo_data/180205_Poo_TruePoo_Full1.mzMLb", 1)
-
-    # Merge data on 'rt' column
-    spec_data_mzml, spec_data_mzmlb = spec_data_mzml.align(spec_data_mzmlb, join='inner')
-
-    assert (spec_data_mzml["mz"] == spec_data_mzmlb["mz"]).all()
-    assert (spec_data_mzml["int"] == spec_data_mzmlb["int"]).all()
-
-def test_get_rtrange_mzmlb():
-    rtrange_data_mzml = get_rtrange_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1.mzML", 6.5, 8)
-    rtrange_data_mzmlb = get_rtrange_mzmlb("demo_data/180205_Poo_TruePoo_Full1.mzMLb", 6.5, 8)
-
-    rtrange_data_mzml, rtrange_data_mzmlb = rtrange_data_mzml.align(rtrange_data_mzmlb, join='inner')
-
-    assert (rtrange_data_mzml["rt"] == rtrange_data_mzmlb["rt"]).all()
-    assert (rtrange_data_mzml["mz"] == rtrange_data_mzmlb["mz"]).all()
-    assert (rtrange_data_mzml["int"] == rtrange_data_mzmlb["int"]).all()

### 2025-03-07 tests/test_sqlite_funs.py

@@ -1,35 +1,0 @@
-import numpy as np
-import pandas as pd
-import pytest
-from mzsql import *
-import os
-
-# Tests for SQLite file type
-def test_turn_mzml_sqlite():
-    turn_mzml_sqlite("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", "demo_data/180205_Poo_TruePoo_Full1.sqlite")
-
-def test_get_chrom_sqlite():
-    ref_data = get_chrom_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1.mzML", 118.0865, 10)
-    test_data = get_chrom_sqlite("demo_data/180205_Poo_TruePoo_Full1.sqlite", 118.0865, 10)
-    assert (ref_data["rt"] == test_data["rt"]).all()
-    assert (ref_data["mz"] == test_data["mz"]).all()
-    assert (ref_data["int"] == test_data["int"]).all()
-
-
-def test_get_spec_sqlite():
-    spec_data_mzml = get_spec_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1_idx.mzML", 1)
-    spec_data_sqlite = get_spec_sqlite("demo_data/180205_Poo_TruePoo_Full1.sqlite", 1)
-
-    # Check if the data matches
-    assert (spec_data_mzml["mz"] == spec_data_sqlite["mz"]).all()
-    assert (spec_data_mzml["int"] == spec_data_sqlite["int"]).all()
-
-
-def test_get_rtrange_sqlite():
-    rtrange_data_mzml = get_rtrange_mzml_pymzml("demo_data/180205_Poo_TruePoo_Full1.mzML", 6.5, 8)
-    rtrange_data_sqlite = get_rtrange_sqlite("demo_data/180205_Poo_TruePoo_Full1.sqlite", 6.5, 8)
-    assert (rtrange_data_mzml["rt"] == rtrange_data_sqlite["rt"]).all()
-    assert (rtrange_data_mzml["mz"] == rtrange_data_sqlite["mz"]).all()
-    assert (rtrange_data_mzml["int"] == rtrange_data_sqlite["int"]).all()
-
-os.remove("demo_data/180205_Poo_TruePoo_Full1.sqlite")

### 2025-03-08 README.qmd
Fully pivoting to the new repo structure

---
title: "The mzsql project"
format: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This repository implements "vanilla" SQL databases for the efficient storage and access of raw mass-spectrometry (MS) data. Here, "raw" MS data refers to the data encoded in vendor or .mzML files and contains coordinates of retention time, m/z ratio, and intensity.

Existing data formats fail to provide intuitive, rapid, and programmatic search of raw MS data and require learning the quirks and conceits of idiosyncratic file formats. Databases have consistent and transferrable syntax for queries, indexing and binary search for rapid data extraction, benefit from multi-file aggregations of data, can store processed data alongside, and are established as an industry standard.

## Convert mzML files into a database via the converters above

Converters from mzML to SQLite and DuckDB are currently available in the [converters folder](converters) for R and Python with Julia and C++ hopefully soon. Code from the associated script can be directly copied and pasted into your workflow, though you're responsible for managing the dependency installations.

The database format is shown in the graphic below:

### Convert from mzML to DuckDB using R:

Ensure `RaMS`, `DBI`, and `duckdb`/`RSQLite` are all installed from CRAN.

```{r make database, eval=FALSE}
library(RaMS)
msdata <- grabMSdata(list.files("demo_data", pattern = "mzML", full.names = TRUE))

library(DBI)
conn <- dbConnect(RSQLite::SQLite(), "demo_data/msdata.sqlite")
# conn <- dbConnect(duckdb::duckdb(), "demo_data/msdata.duckdb")
dbWriteTable(conn, "MS1", msdata$MS1)
dbWriteTable(conn, "MS2", msdata$MS2)
dbDisconnect(conn)
```

## Access the database using SQL

SQL is increasingly popular among mass spectrometrists and is consistent between programming languages.

### Extract a chromatogram

General query: `SELECT * FROM MS1 WHERE mz BETWEEN mzmin AND mzmax`

<details open>
<summary>R code</summary>

Ensure `DBI`, and `RSQLite`/`duckdb` are all installed from CRAN.

```{r rchrom}
library(DBI)
pmppm <- function(mass, ppm)c(mass * (1 - ppm/1e+06), mass * (1 + ppm/1e+06))

conn <- dbConnect(RSQLite::SQLite(), "demo_data/msdata.sqlite")
# conn <- dbConnect(duckdb::duckdb(), "demo_data/msdata.duckdb")
chr_statement <- "SELECT * FROM MS1 WHERE mz BETWEEN ? AND ?"
chrom <- dbGetQuery(conn, chr_statement, params=pmppm(118.0865, 10))
dbDisconnect(conn)
plot(chrom$rt, chrom$int, type = "l")
```

</details>

<details>
<summary>Python</summary>

Make sure the duckdb/sqlite and matplotlib packages are installed.

```{python pychrom}
import duckdb
def pmppm(mz, ppm):
  return(mz*(1-ppm/1000000), mz*(1+ppm/1000000))

conn = duckdb.connect("demo_data/msdata.duckdb")
chr_statement = "SELECT * FROM MS1 WHERE mz BETWEEN ? AND ?"
chrom = conn.execute(chr_statement, pmppm(118.0865, 10)).fetchdf()
conn.close()

import matplotlib.pyplot as plt
plt.plot(chrom["rt"], chrom["int"])
plt.show()
```

</details>

<details>
<summary>Julia</summary>

Make sure the SQLite/DuckDB, DataFrames, DBInterface, and Plots packages are installed.

```{julia jchrom}
using DuckDB, DataFrames, Plots;
pmppm(mass, ppm) = [mass * (1 - ppm / 1e6), mass * (1 + ppm / 1e6)];

conn = DBInterface.connect(DuckDB.DB, "demo_data/msdata.duckdb");
chr_statement = "SELECT * FROM MS1 WHERE mz BETWEEN ? AND ?";
chrom = DataFrame(DBInterface.execute(conn, chr_statement, pmppm(118.0865, 10)));
DBInterface.close!(conn)

plot(chrom.rt, chrom.int, lw=2)
```

</details>

### Search for fragments of a specific precursor

General query: `SELECT * FROM MS2 WHERE premz BETWEEN mzmin AND mzmax`

<details open>
<summary>R code</summary>

Ensure `DBI`, `ggplot2`, and `RSQLite`/`duckdb` are all installed from CRAN.

```{r rpre}
library(DBI)
pmppm <- function(mass, ppm)c(mass * (1 - ppm/1e+06), mass * (1 + ppm/1e+06))

conn <- dbConnect(RSQLite::SQLite(), "demo_data/msdata.sqlite")
# conn <- dbConnect(duckdb::duckdb(), "demo_data/msdata.duckdb")
frag_statement <- "SELECT * FROM MS2 WHERE premz BETWEEN ? AND ?"
frag_data <- dbGetQuery(conn, frag_statement, params=pmppm(118.0865, 10))
dbDisconnect(conn)

library(ggplot2)
ggplot(frag_data, aes(x=fragmz, y=int)) +
  geom_point() +
  geom_segment(aes(xend=fragmz, yend=0)) +
  facet_wrap(~filename, ncol=1) +
  scale_y_log10()
```

</details>

<details>
<summary>Python</summary>

Make sure the duckdb and matplotlib packages are installed.

```{python pypre}
import duckdb
import matplotlib.pyplot as plt
def pmppm(mz, ppm):
  return(mz*(1-ppm/1000000), mz*(1+ppm/1000000))

conn = duckdb.connect("demo_data/msdata.duckdb")
frag_statement = "SELECT * FROM MS2 WHERE premz BETWEEN ? AND ?"
frag_data = conn.execute(frag_statement, pmppm(118.0865, 10)).fetchdf()
fig, axes = plt.subplots(3, 1, sharex=True)
for ax, filename in zip(axes, frag_data["filename"].unique()):
    subset = frag_data[frag_data["filename"] == filename]
    ax.vlines(subset["fragmz"], 0, subset["int"], color='k', zorder=1, linewidths=0.5)
    ax.scatter(subset["fragmz"], subset["int"], color='k', zorder=2, s=10)
    ax.set_yscale("log")
    ax.set_title(filename)
conn.close()

plt.tight_layout()
plt.show()
```

</details>

<details>
<summary>Julia</summary>

Make sure the SQLite/DuckDB, DataFrames, DBInterface, and StatsPlots packages are installed.

```{julia jpre}
using DuckDB, DataFrames, StatsPlots;
pmppm(mass, ppm) = [mass * (1 - ppm / 1e6), mass * (1 + ppm / 1e6)];

conn = DBInterface.connect(DuckDB.DB, "demo_data/msdata.duckdb");
frag_statement = "SELECT * FROM MS2 WHERE premz BETWEEN ? AND ?";
frag_data = DataFrame(DBInterface.execute(conn, frag_statement, pmppm(118.0865, 10)));
DBInterface.close!(conn)

@df frag_data scatter(:fragmz, :int, markersize=3, group=:filename, yscale=:log10)
# @df frag_data plot!([:fragmz :fragmz]', [zeros(size(frag_data, 1)) frag_data.int]', lw=1, group=:filename)
```

</details>

### Search for precursors of a specific fragment

General query: `SELECT * FROM MS2 WHERE fragmz BETWEEN mzmin AND mzmax`

<details open>
<summary>R code</summary>

Ensure `DBI`, `ggplot2`, and `duckdb` are all installed from CRAN.

```{r rfrag}
library(DBI)
pmppm <- function(mass, ppm)c(mass * (1 - ppm/1e+06), mass * (1 + ppm/1e+06))

conn <- dbConnect(RSQLite::SQLite(), "demo_data/msdata.sqlite")
# conn <- dbConnect(duckdb::duckdb(), "demo_data/msdata.duckdb")
pre_statement <- "SELECT * FROM MS2 WHERE fragmz BETWEEN ? AND ?"
pre_data <- dbGetQuery(conn, pre_statement, params=pmppm(59.07336, 10))
dbDisconnect(conn)

library(ggplot2)
ggplot(pre_data, aes(x=premz, y=int)) +
  geom_point() +
  geom_segment(aes(xend=premz, yend=0)) +
  facet_wrap(~filename, ncol=1)
```

</details>

<details>
<summary>Python</summary>

Make sure the duckdb and matplotlib packages are installed.

```{python pyfrag}
import duckdb
import matplotlib.pyplot as plt
def pmppm(mz, ppm):
  return(mz*(1-ppm/1000000), mz*(1+ppm/1000000))

conn = duckdb.connect("demo_data/msdata.duckdb")
pre_statement = "SELECT * FROM MS2 WHERE fragmz BETWEEN ? AND ?"
pre_data = conn.execute(pre_statement, pmppm(59.07336, 10)).fetchdf()
fig, axes = plt.subplots(3, 1, sharex=True)
for ax, filename in zip(axes, pre_data["filename"].unique()):
    subset = pre_data[pre_data["filename"] == filename]
    ax.vlines(subset["premz"], 0, subset["int"], color='k', zorder=1, linewidths=0.5)
    ax.scatter(subset["premz"], subset["int"], color='k', zorder=2, s=10)
    ax.set_yscale("log")
    ax.set_title(filename)
conn.close()
plt.tight_layout()
plt.show()
```

</details>

<details>
<summary>Julia</summary>

Make sure the SQLite/DuckDB, DataFrames, DBInterface, and StatsPlots packages are installed.

```{julia jfrag}
using DuckDB, DataFrames, StatsPlots;
pmppm(mass, ppm) = [mass * (1 - ppm / 1e6), mass * (1 + ppm / 1e6)];

conn = DBInterface.connect(DuckDB.DB, "demo_data/msdata.duckdb");
pre_statement = "SELECT * FROM MS2 WHERE fragmz BETWEEN ? AND ?";
pre_data = DataFrame(DBInterface.execute(conn, pre_statement, pmppm(59.07336, 10)));
DBInterface.close!(conn)

@df pre_data scatter(:premz, :int, markersize=3, group=:filename, yscale=:log10)
# @df pre_data plot!([:premz :premz]', [zeros(size(pre_data, 1)) pre_data.int]', lw=1, group=:filename)
```

</details>

## Other README things

### 2025-03-08 converters/duckdb_functions.py

@@ -1,0 +1,105 @@
+
+import numpy as np
+import pandas as pd
+import os
+import duckdb
+import pyteomics.mzml
+from .helpers import pmppm
+
+def turn_mzml_duckdb(files, outfile, ordered=None):
+    conn = duckdb.connect(outfile)
+    conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS2")
+    empty_ms1 = pd.DataFrame(columns=["filename", "id", "mz", "int", "rt"]).astype(
+        {"filename": "string", "id": "int64", "mz": "float64", "int": "float64", "rt": "float64"}
+    )
+    conn.execute("CREATE TABLE MS1 AS SELECT * FROM empty_ms1")
+    empty_ms2 = pd.DataFrame(columns=["filename", "id", "premz", "fragmz", "int", "rt"]).astype(
+        {"filename": "string", "id": "int64", "premz": "float64", "fragmz": "float64", "int": "float64", "rt": "float64"}
+    )
+    conn.execute("CREATE TABLE MS2 AS SELECT * FROM empty_ms2")
+
+    if isinstance(files, str):
+        files = [files]
+    for file in files:
+        if not os.path.exists(file):
+            raise FileNotFoundError(f"Input mzML file '{file}' does not exist.")
+        MS1_dfs = []
+        MS2_dfs = []
+        for spectrum in pyteomics.mzml.MzML(file):
+            if spectrum['ms level'] == 1:
+                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
+                mz_vals = spectrum['m/z array']
+                int_vals = spectrum['intensity array']
+                rt_val = spectrum['scanList']['scan'][0]['scan start time']
+                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 
+                                        'mz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+                MS1_dfs.append(df_scan)
+            if spectrum["ms level"] == 2:
+                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
+                premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
+                mz_vals = spectrum['m/z array']
[…]

### 2025-03-08 converters/sqlite_functions.py

@@ -1,0 +1,159 @@
+import sqlite3
+import pandas as pd
+import pyteomics.mzml
+import os
+from .helpers import pmppm
+
+def turn_mzml_sqlite(files, outfile, ordered=None):
+    conn = sqlite3.connect(outfile)
+    conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS2")
+    if isinstance(files, str):
+        files = [files]
+    for file in files:
+        MS1_dfs = []
+        MS2_dfs = []
+        for spectrum in pyteomics.mzml.MzML(file):
+            if spectrum['ms level'] == 1:
+                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
+                mz_vals = spectrum['m/z array']
+                int_vals = spectrum['intensity array']
+                rt_val = spectrum['scanList']['scan'][0]['scan start time']
+                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'mz': mz_vals, 
+                                        'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+                MS1_dfs.append(df_scan)
+            if spectrum['ms level'] == 2:
+                idx = int(spectrum['id'].split("scan=")[-1].split()[0])
+                mz_vals = spectrum['m/z array']
+                int_vals = spectrum['intensity array']
+                rt_val = spectrum['scanList']['scan'][0]['scan start time']
+                premz_val = spectrum['precursorList']['precursor'][0]['isolationWindow']['isolation window target m/z']
+                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'premz': premz_val, 
+                                        'fragmz': mz_vals, 'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+                MS2_dfs.append(df_scan)
+        pd.concat(MS1_dfs, ignore_index=True).to_sql("MS1", conn, if_exists="append", index=False)
+        pd.concat(MS2_dfs, ignore_index=True).to_sql("MS2", conn, if_exists="append", index=False)
+    if ordered is not None:
+        index_name = f"idx_{ordered}"
+        if ordered == "rt":
+            conn.execute(f"CREATE INDEX IF NOT EXISTS {index_name} ON MS1 ({ordered})")
[…]

### 2025-03-11 manuscript/manuscript.Rmd
Nearly completed first draft.

@@
 Crucially, the original paper [@Wilhelm2012] contains links to a website (https://software.steenlab.org/mz5) which currently returns an HTTP error 500.
-A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2) and hasn't been updated in 12 years and is predominately a simple fork of h5py [@ref] with three mz5-specific commits on top.
+A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2) and hasn't been updated in 12 years and is predominately a simple fork of h5py [@Collette2017] with three mz5-specific commits on top.
 Most problematically, we were unable to determine how mz5 stores precursor *m/z* ratios, making the fragment and precursor searches impossible.
@@
 
-We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for MS1 and MS2 data.
+We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS1) and MS/MS (MS2) data.
 The MS1 table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
@@
                                       "ms2_spec", "premz", "fragmz"),
-                      labels=c("MS1 scan", "Chromatogram", "RT Range",
-                               "MS2 scan", "Precursor search",
+                      labels=c("Full scan", "Chromatogram", "RT Range",
+                               "MS/MS scan", "Precursor search",
                                "Fragment search")))
@@
 *Figure XX: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper.
-The left six panels show boxplots representing the time required in seconds to extract an MS1 scan, an ion chromatogram, all data within a retention time range, an MS2 scan, the fragments of a specified precursor, and all precursors with a specified fragment.
+The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum, an ion chromatogram, all data within a retention time range, an MS/MS scan, the fragments of a specified precursor, and all precursors with a specified fragment.
 The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z.
@@
 
+*Figure XX: blah*
+
 # Discussion
@@
 As the gap between data scientist and mass spectrometrist continues to narrow, mass spectrometry (MS) data formats should facilitate this convergence.
+Instead, MS software remains relatively opaque.
+Documentation is sparse and data structures are complex, resulting in a landscape that is essentially restricted to the original developer's intent.
+A particular pain point is the way in which MS data is stored because current methods must make trade-offs between simplicity, size, and speed.
+When we explored the wide range of MS file readers, we found every method had flaws that interfered with widespread adoption.
+The mzML file type appears to represent low-hanging fruit, with its XML-based structure that sacrifices speed and size in favor of clarity, but no alternative has yet reached a large audience of active users.
+Fifteen years of active competition continue to favor the highly explicit format, likely because users are leery of incomprehensible alternatives with no guarantee of continued maintenance.
 
-Large gaps etc.
-accomplished by extremely poor documentation that makes it difficult to do anything beyond the original designer's intent, typically with data structures that are not familiar to the MS expert.
+Certain methods are clear winners for individual use cases but all of the existing formats failed to perform well at the full suite of exploratory data analysis tasks we attempted.
+Scan extraction is perhaps the most widely used query, especially for proteomics, and as a result has been extensively optimized.
+Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number.
+The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supp Fig XX).
 
-Not a single parser was without flaws - all were either slow, poorly documented, or had highly limited functionality
+However, performant scan-based methods struggled significantly with chromatogram extraction because these axes are inherently orthogonal to each other.
+Here, mzDB performed quite well and was competitive with the specifically optimized mzTree and mzMD formats and pyopenms performed incredibly well when extracting multiple chromatograms via its 2DPeak method.
+Chromatograms (and retention time range queries) are of course only relevant for chromatography-based workflows but this type of analysis has become increasingly popular and we expect it to continue to do so, making ion extraction increasingly important.
+None of the existing MS formats we tested performed very well on MS/MS data, despite the growing availability of fragmentation data, though our use-case is oriented more towards exploration instead of comprehensive analysis.
 
-Several fundamental inefficiencies identified:
-1. Looping over each scan within a file
-2. Looping over each file within a dataset
-3. Decoding mz/int information each time if filtering on these fields
-4. RTs are not accessible outside of looping over each scan for RT subsetting
-5. Scan ids are not accessible outside of looping over each scan for scan extraction
+Finally, a complexity penalty must be noted for formats and packages requiring complex installation procedures.
+pyopenms appears to be the worst offender here, with its bindings to the OpenMS C++ libraries requiring us to step back to Python 3.11 and numpy 1.26 to successfully access the data on our setup and required direct input from the maintainers.
+The Java applications for MzTree and, more egregiously mzMD, provided essentially zero documentation on installation consisting of a single README file without an intact HTTP request example and have not seen updates in years.
+Similarly, mzDB files were difficult to parse due to its opaque SQLite schema and use of the BLOB encoding type that again lacked examples or documentation outside of Java and had to be deduced iteratively.
+Of course, these methods eventually provided enough information that they could be parsed unlike Aird, toffee, and UIMF.
 
-Querying things by scan number is dumb.
-1) Different indices for pyopenms vs pymzml vs pyteomics
-2) idx = scan_num only works for consecutive scans (no multi-experiment, no polarity switching, no filtering)
-3) Only USEFUL when you're parsing the entire file! Nobody knows what scan number 976 corresponds to, making extraction by number useless
+### Timing comparison to existing literature
 
-DuckDB's zonemaps operate similarly to mzDB's bounding boxes
-DuckDB also has automatic compression detection (e.g. the delta encoding of mz5)
+Novel formats are typically proposed with timing and sizing information, but the inconsistency of what's being queried makes it difficult to directly compare across the literature.
+However, we mostly obtained results in line with those reported elsewhere where other intercomparisons have been performed and report here the widest set of comparisons between MS formats to our knowledge.
 
-Unlikely that databases replace the original files so this is meant to be an auxiliary data structure **alongside** the .raw or .mzML files, not a substitute.
+@Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers.
+Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of two times smaller (50%).
 
-Databases place a larger burden on the initial builder/designer and much less on the end user
+@Bouyssie2015 also found that mz5 was about 80% smaller than the mzML and on par with their mzDB format.
+They reported query times of around 30 seconds for a wide (5 Da) ion chromatogram extraction and found that mz5 was about 40 times slower and that mzML was 200 times slower, in contrast to our parser which was 10 times slower for mz5 - possibly due to the large EIC width.
+They also were able to report mzDB scan queries on par with mz5 which we were able to replicate if the pyopenms or pyteomics libraries were used for full scan queries and pymzml for MS/MS.
 
-Better compression algorithms and fancier data structures can easily improve upon the text-based XML in both size on disk and speed of access but also tend to accumulate significant technical debt.
-Alternatives must therefore compete not only in computational efficacy but also in simplicity to allow widespread adoption, a feat thus far unachieved.
+MzTree [@Handy2017] compared their SQLite-based system to mzML, mz5, and mzDB and reported high numbers for both EIC and RT range queries from mzML (4-1000 seconds) that were on par with the values we observed here.
+Their mzDB and mz5 results were unexpectedly comparable to each other at about 0.5 seconds per random EIC query, with the mzDB values equal to ours but the mz5 values much lower than our parser was able to obtain.
+Our disk size measurements also corresponded well with their size estimate of MzTree at approximately twice the size of the mzML, a surprising result given that our mzML file contained MS/MS information and theirs appearing to be full-scan only.
+Their mz5 files were much larger (80% mzML size instead of our 20%) and their mzDB much smaller (20% mzML size instead of our 110%).
 
-Ultimately, the design decision for mass spectrometry data format will likely result from a variety of factors, most crucially 1) initial vendor type, 2) programming language of the developer, 3) types of MS data included (MS1-only vs MS2+, metadata), 4) whether the entire file will be processed or only a subset (subsetting doesn't matter if the whole file is going to be investigated anyway),
+The mzMD format [@Yang2022] appears to be a thin wrapper around the MzTree format that applies a different philosophy for data subsetting and summarization.
+They report EIC queries in the 50 millisecond range, very similar to the values we obtained for the mzMD file type.
+They also estimate file size to be approximately 28 bytes per *m/z*/intensity tuple for a total size of 100MB in our test file which agrees reasonably well with our 72MB actual measurement.
+Their comparison to MzTree also agrees with our results as they report slightly larger disk usage and slightly better performance.
+
+The mzMLb group [@Bhamber2021] reports only info for spectrum access at approximately 15ms/scan which agrees with our 5-100ms/scan estimates only if the data is loaded ahead of time.
+They perform extensive comparisons to mzML at varying compression methods and levels but our default options of 1024 KB chunk sizes for the mzMLb file and zlib but not numpress compression for the mzML resulting in timing values very similar between mzML and mzMLb using the pyteomics library for access.
+They also compared to the mz5 file type and we are able to validate their results of the mzML+zlib occupying significantly more space, though our mz5 parser outperformed theirs for full scan and MS/MS data by two orders of magnitude in time.
+
+We were also unable to test several other recent and promising formats.
+Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF, @Beagley2009) from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
+
+### Fundamental inefficiencies in existing mass-spectrometry formats
+
+We identified several fundamental inefficiencies when writing the parsers.
+First, scan metadata that was encoded within a scan instead of in a separate unit required looping over each scan to see whether it contained the information requested.
+Scan number, MS level, and retention time were all useful bits of information that could be included in a file header or footer to relate the three to the data location within the file and allow index use instead of looping over every scan.
+Second, needing to decode or parse a compressed *m/z*/intensity array in each scan introduced an additional overhead that was especially punishing during ion chromatogram extraction and MS/MS search.
+While the *m/z* and intensity tuples are an obvious candidate for data compression, this penalty should be of significant concern to engineers.
+Third, looping over files is inherently slow and introduces additional complexity relative to a single unified database that encodes filename or sample ID as an additional column.
+A particular strength of the database system we propose is its inherent support for multi-file systems, while all other methods require looping over files.
+
+The problems above highlight an important distinction between data *access* and data *search* that has been largely overlooked in our opinion.
+While HDF5 files or scan indexes excel at improving data access, they assume that the location of the data is known in advance and can be skipped to via bitwise offsets.
+If a search is required, however, this advantage is fully negated because each bit of information must be queried anyway.
+Finally, we must note that scan number is not inherently a useful bit of information.
+While we included it in our extraction metrics, it is entirely unclear when the scan number itself would be known in isolation.
+Additionally, this method often confuses the scan number with the scan's indexes in the data structure.
+Scan number is not always consecutive (i.e.
+during polarity switching, multi-experiment samples, or if any filtering is performed during processing), so even if the first or second item in the structure can be queried speedily this is no guarantee that the item will contain the information of interest.
+
+### Leveraging robust, future-oriented software development with SQL
+
+The proliferation of MS data storage formats and access algorithms illustrates the general dissatisfaction with existing alternatives to the vendor file or mzML.
+Formats that are faster to query or smaller on disk tend to be significantly more opaque, and those optimized for a particular method often fail to perform well on other metrics.
+This complexity is generally expected as optimization tends to require more complex data structures and assumptions about its use but it is not required if the complexity is outsourced to a robust and growing framework such as structured query language (SQL).
+
+SQL is widely used for data processing outside of mass-spectrometry, though its adoption is increasing in recent years.
+Efforts like mzDB, the Pacific Marine Environment Laboratory's UIMF format [@Beagley2009], and the internals of MzTree hint at SQL's suitability for MS data storage.
+SQL backends for the next-generation R processing package Spectra now exist [@Rainer2022] and the development of MassQL [@Jarmusch2022] indicates a growing comfort with SQL syntax for downstream processing, though the language itself strives for human readability in simple queries.
+The searching and subsetting inherent to MS data exploration represent very simple queries in database space, agnostic to high-level programming language and rarely requiring more than a single line of code.
+Additionally, the extensive documentation that exists across the internet means that large language models such as ChatGPT are easily able to translate queries for those unfamiliar with SQL's syntax.
+
+Just as the original database paper from @Codd1970 argued that the same problems were being solved over and over again, mass spectrometry data scientists are re-solving problems that have been more elegantly ironed out by dedicated teams in computer science and industry with much more extensive support.
+By leveraging existing optimizations in SQLite and DuckDB, we were able to create a highly performant system for storage of MS data that does not come with significant trade-offs between data extraction methods.
+
+While SQLite is broadly used and its long history testifies to its continued utility, we can use even more modern database methods to improve further upon its analytical processing capacity.
+We tested both DuckDB the Apache Parquet data formats [@Raasveldt2019; @Vohra2016] and found that they both performed better than SQLite in disk usage and query speed.
+DuckDB in particular is nearly a drop-in replacement for SQLite in many cases that's been extensively optimized for MS-related queries given its online analytical processing (OLAP) structure.
+DuckDB provides automatic compression algorithms and uses zonemaps to create bounding boxes for each subset of data, bringing together existing optimizations from mz5 (delta encoding), mzDB (bounding boxes), and MzTree (axis-agnostic queries) at zero additional cost.
+Importantly, as with all databases, only the subset of interest needs to be written into memory, making the hardware requirements relatively lightweight.
+
+Of course, to claim that existing frameworks should be discarded in favor of a novel method is to ignore decades of discussion and conversation.
+We acknowledge that our use case, that of largely exploratory and quality-control steps, is not a universal need and our lack of perfect metadata preservation in particular indicates that databases should become an auxiliary data structure alongside the vendor files or mzMLs, not substitute for them directly.
+Ultimately, the design decision for mass spectrometry data format will likely continue to be a point of contention and will result from a variety of factors, most crucially 1) initial vendor type, 2) programming language of the developer, 3) types of MS data included (e.g.
+full scan only versus MS/MS or metadata requirements), 4) whether the entire file will be processed or only a subset, and 5) how well a file type interfaces with downstream software.
+We intend to show with this manuscript that there is significant overlap between the goals of organizations much larger than any individual lab and that MS as a whole can benefit significantly from co-opting their development.
 
 # Conclusion
+
+We propose that a simple relational database is an intuitive and performant mass spectrometry (MS) data storage format.
+Tables containing fields that map directly to known MS concepts means that adoption is straightforward and facilitated by the widely-understood structured query language (SQL), reducing the code required to extract subsets of interest to a single line.
+We show that this structure can also take advantage of regular advancements in computer science by leveraging modern data formats such as DuckDB and Parquet to reduce the disk space required while improving access times by 1-2 orders of magnitude.
+We hope that widespread adoption of this format alongside the metadata-heavy vendor and mzML files will reduce the barriers to data access for mass spectrometrists and provide a consistent framework that covers a majority of the exploratory use cases.
+
+[add R/Julia comparison?]
 

### 2025-03-11 manuscript/manuscript.Rmd
First draft done and circulated to the lab

@@
                   "SQLite", "DuckDB", "Parquet")
-method_colors <- c("#0B4F5B", "#147C8F", "#21A8C0", "#44C3DA", 
-                   "#028e34", "#f4bb23", "#e0670b", "#a41118", 
-                   "#E25C76", "#E891A2", 
-                   "#580770", "#8A1BAC", "#B04ACF")
+method_colors <- c("#393B79", "#5254A3", "#6B6ECF", "#9C9EDE", 
+                   "#0F8299", "#637939", "#E6550D", "#843C39", 
+                   "#7B4173", "#CE6DBD", 
+                   "#8C6D31", "#E7BA52", "#efddb9")
 
@@
 
-## Mass-spectrometry files
+## Mass-spectrometry files and software used
 
@@
 SQL databases were built using Python 3.11.11 with SQLite (v3.48.0) via the Python sqlite3 package (v2.6.0), DuckDB via the duckdb package (v1.1.3), and Parquet files via the pyarrow package (v19.0.0).
-We were additionally able to download and install []
+We were additionally able to download and install [blah]
 
@@
 
-```{shell, eval=FALSE}
+```{r shell, eval=FALSE}
 cd Documents\Will
@@
 Here, we map MS concepts (retention time, *m/z*, intensity, etc.) directly to database fields to make downstream processing as intuitive as possible.
-Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure XX).
+Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure 1).
 We do not force compression of any of these fields because we believe that decoding compressed data is both a slow and unintuitive step, though automatic compression is supplied by the DuckDB and Parquet file types.
 
-```{r gtables for figure 1}
+```{r gtables for figure 1, eval=FALSE}
 library(gt)
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/fileinfo_table.png", vwidth=1000)
+  gtsave(filename = "manuscript/figures/fileinfo_table.png", vwidth=1000)
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/scaninfo_table.png", vwidth=1000)
+  gtsave(filename = "manuscript/figures/scaninfo_table.png", vwidth=1000)
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/ms1_table.png", vwidth=1000)
+  gtsave(filename = "manuscript/figures/ms1_table.png", vwidth=1000)
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/ms2_table.png", vwidth=1000)
+  gtsave(filename = "manuscript/figures/ms2_table.png", vwidth=1000)
 ```
@@
 
-*Figure XX: Database schema for an example MS/MS dataset showing the organization of mass-spectrometry data into tables.
+*Figure 1: Database schema for an example MS/MS dataset showing the organization of mass-spectrometry data into tables.
 Fields of interest are easily queryable with simple SQL commands as shown in the table at bottom.*
+
+*Supplemental figure XX: (TBD) Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imagining.*
 
@@
 
-[TODO: Note the method used to identify the m/zs, scans, and RT ranges extracted here]
+We randomly sampled a single file out of the full dataset for comparison across metrics and file formats (20220923_LEAP-POS_QC04).
+We sampled 100 random scan numbers using SQLite's ORDER BY RANDOM() function and pulled out the largest ions for chromatogram extraction using a 10 ppm mass range window.
+Retention time ranges were the highest-intensity retention time of each ion chromatogram plus or minus one minute.
+Similarly, the largest fragments by intensity were used for the MS/MS metrics with a 10 ppm mass range exclusion window.
 
-Via %timeit% (need to redo these with randomization)
+Timing was performed via Python's timeit library and the timeit.repeat function, with the various file formats as the innermost loop to ensure bias over time was distributed equally among function calls.
+File sizes were estimated using Python's os library with os.path.getsize.
+We did not exhaustively monitor memory usage, though our heuristic exploration of the timing scripts did not ever indicate that memory was a constraint.
+
+Timing data was obtained on a Lenovo XX blah XX with [cores] and [memory].
 
@@
 
+### Spectrum extraction
+
+The simplest and most abundantly documented query was the extraction of a single spectrum.
+In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 1, leftmost panels).
+Here, we found that the mzML and mzMLb file types were consistently the slowest to parse the mzML file type and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
+`pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX).
+It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number as scans are not always consecutive (e.g.
+during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan.
+
 ```{r singlefile_fig}
-all_timings <- read_csv("data/singlefile_times.csv") %>% 
+all_timings <- read_csv("manuscript/data/singlefile_times.csv") %>% 
   mutate(method=factor(method, levels=method_names)) %>%
@@
                                "Fragment search")))
-all_sizes <- read_csv("data/file_sizes.csv") %>%
+all_sizes <- read_csv("manuscript/data/file_sizes.csv") %>%
   mutate(method=factor(method, levels=method_names))
@@
   scale_fill_manual(breaks = c("mzML", method_names), 
-                    values=c("#147C8F", method_colors)) +
+                    values=c("#393B79", method_colors)) +
   facet_wrap(~query) +
@@
 
-ggsave("figures/singlefile_fig.png", width = 6.5, height=5)
+ggsave("manuscript/figures/singlefile_fig.png", width = 6.5, height=5)
 
@@
 
-*Figure XX: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper.
+*Figure 2: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper.
 The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum, an ion chromatogram, all data within a retention time range, an MS/MS scan, the fragments of a specified precursor, and all precursors with a specified fragment.
@@
                "<span style='color:#0b505c'>Extraction only</span>")
-init_outside_gp <- read_csv("data/singlefile_supp.csv") %>%
+init_outside_gp <- read_csv("manuscript/data/singlefile_supp.csv") %>%
   pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
@@
 
-pyopenms_precomp_gp <- read_csv("data/pyopenms_precomp.csv") %>%
+pyopenms_precomp_gp <- read_csv("manuscript/data/pyopenms_precomp.csv") %>%
   pivot_longer(rep1:rep3) %>%
@@
 init_outside_gp / pyopenms_precomp_gp
-ggsave("figures/supp_preload_fig.png", width = 6.5, height=5)
+ggsave("manuscript/figures/supp_preload_fig.png", width = 6.5, height=5)
 ```
@@
 Error bars show 3x replicate queries of 5x different targets.*
-
-### Spectrum extraction
-
-The simplest and most abundantly documented query was the extraction of a single spectrum.
-In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 1, leftmost panels).
-Here, we found that the mzML and mzMLb file types were consistently the slowest to parse the mzML file type and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
-`pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX).
-It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number as scans are not always consecutive (e.g.
-during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan.
 
@@
 We also investigated the efficacy of the various MS data formats for MS/MS data and found that support for fragmentation data searches was lacking or absent from the documentation and exposed functionality of each of these file types, requiring custom implementations every time.
-Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure XX, rightmost panels).
+Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure 2, rightmost panels).
 
-All existing MS data types required multiple seconds to perform a single fragment search (Figure XX, bottom-right panel), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor.
+All existing MS data types required multiple seconds to perform a single fragment search (Figure 2, bottom-right panel), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor.
 The SQL-based parsers, on the other hand, all took fractions of a second and consistently returned the relevant data hundreds of times more quickly than existing methods.
-The same was true for a precursor search across all methods aside from mzDB (Figure XX, bottom middle panel), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.
+The same was true for a precursor search across all methods aside from mzDB (Figure 2, bottom middle panel), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.
 
@@
 We measured the size on disk of each of the file types and found that they varied by approximately an order of magnitude, with HDF-based file types hovering around one-third the size of the mzML (83 megabytes, MB) while mzDB and mzMD were larger (108 MB and 96 MB, respectively).
-The SQLite object was the largest on disk of all the file types, nearly doubling the mzML's size at 145 MB, while DuckDB was on par with MZA at two-thirds of the mzML (54 MB) and Parquet improved slightly upon that again with its columnar-based storage format (Figure XX).
+The SQLite object was the largest on disk of all the file types, nearly doubling the mzML's size at 145 MB, while DuckDB was on par with MZA at two-thirds of the mzML (54 MB) and Parquet improved slightly upon that again with its columnar-based storage format (Figure 2).
 
@@
 
-The single-file, single-metric case discussed above and shown in Figure XX is largely a worst-case scenario for many MS data systems that have a slow initial setup step to make downstream analysis faster.
+The single-file, single-metric case discussed above and shown in Figure 2 is largely a worst-case scenario for many MS data systems that have a slow initial setup step to make downstream analysis faster.
 To compare these systems more fairly to our database schema, we also tested timings across multiple chromatograms.
-In each case, this was implemented as a for loop iterating over an increasing number of chromatograms corresponding to the largest intensity ions in the file (Figure XX).
+In each case, this was implemented as a for loop iterating over an increasing number of chromatograms corresponding to the largest intensity ions in the file (Figure 3).
 
 ```{r}
-multichrom_times <- read_csv("data/multichrom_times.csv")
+multichrom_times <- read_csv("manuscript/data/multichrom_times.csv")
 timelabs <- c("0.01 seconds", "One second", "One minute", "One hour")
@@
   theme(panel.grid.minor = element_blank())
-ggsave("figures/multichrom_fig.png", width = 6.5, height=5)
+ggsave("manuscript/figures/multichrom_fig.png", width = 6.5, height=5)
 ```
@@
 
-*Figure XX: Scatter plot of the time required to extract multiple chromatograms using various methods on logarithmic axes.
+*Figure 3: Scatter plot of the time required to extract multiple chromatograms using various methods on logarithmic axes.
 Best-fit linear models have been added for each method are shown behind triplicate timing measurements.
@@
 
-Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure XX).
+Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3).
 pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
@@
 ```{r}
-multichrom_times <- read_csv("data/multichrom_times.csv")
+multichrom_times <- read_csv("manuscript/data/multichrom_times.csv")
 multichrom_baseline_df <- multichrom_times %>%
@@
   theme(legend.position = "none")
-ggsave("figures/supp_multichrom_facets.png", width = 6.5, height=5)
+ggsave("manuscript/figures/supp_multichrom_facets.png", width = 6.5, height=5)
 ```
@@
 
-*Supplemental figure XX: Data from main text Figure XX with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
+*Supplemental figure XX: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
 Best-fit linear models have been added for each method are shown behind triplicate timing measurements.
@@
                "<span style='color:#147C8F'>Non-equi join</span>")
-read_csv("data/dbloopjoinuni_times.csv") %>%
+read_csv("manuscript/data/dbloopjoinuni_times.csv") %>%
   separate(method, into = c("method", "type")) %>%
@@
         plot.caption.position =  "plot")
-ggsave("figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
+ggsave("manuscript/figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
 ```
@@
 
-We found that SQLite queries benefitted significantly from the construction of an index on the *m/z* column when extracting chromatograms, improving lookup times by an order of magnitude (dropping from [0.3] seconds to [0.03 seconds], Figure XX).
+We found that SQLite queries benefitted significantly from the construction of an index on the *m/z* column when extracting chromatograms, improving lookup times by an order of magnitude (dropping from [0.3] seconds to [0.03 seconds], Figure 4).
 However, because the SQLite index is stored on disk alongside the data, this improvement also increased the file size by 33% from [150MB] to [210MB].
-DuckDB also improved significantly with ordered data but to a smaller degree and surprisingly required additional space to do so, likely due to the reordering resulting in a different compression strategy (Figure XX).
+DuckDB also improved significantly with ordered data but to a smaller degree and surprisingly required additional space to do so, likely due to the reordering resulting in a different compression strategy (Figure 4).
 Parquet files had the smallest improvement upon data ordering but remained the same small size they were when unordered.
@@
 Crucially, these improvements also persisted when multiple files were stored in a single database.
-We built databases consisting of between 1 and 100 individual MS files and tested the time required to extract ion chromatograms from each after an index was constructed (Supplemental fig XX).
+We built databases consisting of between 1 and 100 individual MS files and tested the time required to extract ion chromatograms from each after an index was constructed (Figure 4).
 DuckDB was consistently the fastest ion chromatogram extraction method, with query times around 0.03 seconds for a single file and 1 second for one hundred files.
@@
 However, DuckDB's performance degrades at larger database sizes and approaches a 1:1 slope, possibly due to the overhead of reading large amounts of data into memory after it's found.
-We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure XX).
-Interestingly, SQLite timings were slightly better with multiple database files instead of one combined database (Figure XX).
+We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure 4).
+Interestingly, SQLite timings were slightly better with multiple database files instead of one combined database (Figure 4).
 
@@
 
-read_csv("data/multifile_timings.csv") %>%
+read_csv("manuscript/data/multifile_timings.csv") %>%
   mutate(type=factor(type, levels=c("consolidated", "multifile"),
@@
   theme(legend.position = "top", legend.direction = "vertical")
-ggsave("figures/multifile_fig.png", width = 6.5, height=3.5)
+ggsave("manuscript/figures/multifile_fig.png", width = 6.5, height=3.5)
 ```
@@
 
-*Figure XX: blah*
+*Figure 4: Time required to extract an ion chromatogram from multiple files plotted against the size of the data, broken down by the type of database used (SQLite, DuckDB, or Parquet).
+Points correspond to a random subset of 1, 3, 10, 30, and 100 files, respectively.
+Colors specify whether the data was stored as a single consolidated database (purple) or with a single database per file (orange) and the shape of the point denotes whether the database was unstructured (squares) or indexed/ordered by m/z (SQLite has indexes, DuckDB and Parquet benefit from ordering).
+Ten replicates of each query were performed and are shown transparently behind the mean values connected with lines.*
 

### 2025-03-18 manuscript/manuscript.Rmd
First draft internal edits

@@
   - William Kumler
+  - Sam LaRue
   - Anitra E. Ingalls
@@
 Mass spectrometry (MS) still lacks a performant data access format.
-The mzML file type [@Martens2011], a result of over a decade of interlaboratory collaboration and workshopping, struggles to provide rapid computational access to the m/z and intensity tuples.
+The mzML file type [@Martens2011], a result of over a decade of interlaboratory collaboration and workshopping, struggles to provide rapid computational access to the *m/z* and intensity pairs.
 This is the crucial component in nearly all mass spectrometry analysis, but mzML’s text-based XML format requires time-consuming decompression performed one scan at a time.
@@
 Relational databases also have several distinct advantages over hierarchical or text-based systems, particularly in performing searches for subsets of data via indices.
-Importantly, this indexing differs from the byte-offset indexes that already exist in the indexed mzML and HDF5 formats because the search for a particular subset cannot be done efficiently with a byte-offset index when the m/z data is encoded, though access to a particular scan can be incredibly rapid.
-Additionally, data from multiple samples can be stored together in a single table, differing from formats like mzDB, mzTree, and mzMD and allowing queries of all dataset samples to be performed without looping through each file in turn, thereby avoiding the associated computational overhead and query complexity.
+Importantly, this indexing differs from the byte-offset indexes that already exist in the indexed mzML and HDF5 formats because the search for a particular subset cannot be done efficiently with a byte-offset index when the *m/z* data is encoded, though access to a particular scan can be incredibly rapid.
+Additionally, data from multiple samples can be stored together in a single database table, differing from formats like mzDB, mzTree, and mzMD and allowing queries of all dataset samples to be performed without looping through each file in turn, thereby avoiding the associated computational overhead and query complexity.
 
@@
 Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage.
-Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity tuples is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
+Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity pairs is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
 We compare the time and space required to extract a representative data subset under six conditions and perform these tests on multiple databases as well as mzML and other MS data formats.
@@
 
-*Supplemental figure XX: (TBD) Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imagining.*
+```{r}
+
+```
+
+*Supplemental figure XX: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
 
@@
 
-Timing data was obtained on a Lenovo XX blah XX with [cores] and [memory].
+Timing data was obtained on an Intel Xeon CPU with two X5650 (\@2.67 GHz) processors and 24 total cores running Windows 10 Pro (64 bit version).
+96 gigabytes of RAM (DDR3 @ 1333 MHz) were available and a solid-state drive was used for disk storage.
 
@@
 This was largely due to the variable-length nested compound structures mz5 that are not supported in all APIs, e.g.
-Java (@Bhamber2021).
+Java.[@Bhamber2021]
 
@@
 
-We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS1) and MS/MS (MS2) data.
-The MS1 table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
+We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS<sup>1</sup>) and MS/MS (MS<sup>2</sup>) data.
+The MS<sup>1</sup> table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
 The MS2 table consisted of the same fields except that the *m/z* column was separated into precursor and fragment *m/z*.
 Although we did not extend these databases to include the metadata associated with each file and scan, the logical framework could be easily extended in future work and the metadata typically represents a small fraction of the total space within the file, allowing us to make reasonable comparisons about file size between the databases and the metadata-rich other file types.
-We also converted each file's MS1 and MS2 table into Parquet representations for comparison using the same field/column schema.
+We also converted each file's MS<sup>1</sup> and MS2 table into Parquet representations for comparison using the same field/column schema.
 
@@
 After establishing a connection to the database, the six queries could be asked using nearly human-readable SQL syntax.
-Requesting the thousandth MS1 scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the `pandas.read_sql_query` function.
+Requesting the thousandth MS<sup>1</sup> scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the `pandas.read_sql_query` function.
 More complicated queries such as retention time range (`SELECT * FROM MS1 WHERE rt BETWEEN 6 AND 8`) and a precursor mass search (`SELECT * FROM MS2 WHERE premz BETWEEN 118.086 AND 118.087`) were similarly intuitive.
@@
 The simplest and most abundantly documented query was the extraction of a single spectrum.
-In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 1, leftmost panels).
+In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D).
 Here, we found that the mzML and mzMLb file types were consistently the slowest to parse the mzML file type and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
@@
 ```{r singlefile_fig}
-all_timings <- read_csv("manuscript/data/singlefile_times.csv") %>% 
+all_timings <- read_csv("data/singlefile_times.csv") %>% 
   mutate(method=factor(method, levels=method_names)) %>%
@@
                                       "ms2_spec", "premz", "fragmz"),
-                      labels=c("Full scan", "Chromatogram", "RT Range",
-                               "MS/MS scan", "Precursor search",
-                               "Fragment search")))
-all_sizes <- read_csv("manuscript/data/file_sizes.csv") %>%
+                      labels=c("A. Full scan", "B. Chromatogram", "C. RT Range",
+                               "D. MS/MS scan", "E. Precursor search",
+                               "F. Fragment search")))
+all_sizes <- read_csv("data/file_sizes.csv") %>%
   mutate(method=factor(method, levels=method_names))
@@
   theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
-        legend.position = "none")
+        legend.position = "none",
+        strip.text.x = element_text(hjust = 0))
 size_barplot <- all_sizes %>%
@@
   mutate(method=factor(method, levels=c("mzML", method_names))) %>%
-  mutate(query="File size") %>%
+  mutate(query="G. File size") %>%
   ggplot() +
@@
   theme_bw() +
-  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), legend.position = "none")
+  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5), 
+        legend.position = "none",
+        strip.text.x = element_text(hjust = 0))
 
@@
 
-ggsave("manuscript/figures/singlefile_fig.png", width = 6.5, height=5)
+ggsave("figures/singlefile_fig.png", width = 6.5, height=5)
 
@@
 *Figure 2: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper.
-The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum, an ion chromatogram, all data within a retention time range, an MS/MS scan, the fragments of a specified precursor, and all precursors with a specified fragment.
+The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum (A), an ion chromatogram (B), all data within a retention time range (C), an MS/MS scan (D), the fragments of a specified precursor (E), and all precursors with a specified fragment (F).
 The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z.
-The right panel shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*
+The right panel (G) shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*
 
@@
 
-The `pymzml` package was able to extract both MS1 and MS2 spectra from the mzML file nearly two orders of magnitude faster than the other mzML parsers, largely due to its use of naming the scans by their number and thus avoiding the expensive scan number extraction step.
-mzDB had the only notable difference between MS1 and MS2 scans, performing slightly better than `pyteomics` and `pyopenms` methods for MS1 data and significantly better for MS2 data, placing it approximately on par with `pymzml` in taking about a tenth of a second.
+The `pymzml` package was able to extract both MS<sup>1</sup> and MS<sup>2</sup> spectra from the mzML file nearly two orders of magnitude faster than the other mzML parsers, largely due to its use of naming the scans by their number and thus avoiding the expensive scan number extraction step.
+mzDB had the only notable difference between MS<sup>1</sup> and MS<sup>2</sup> scans, performing slightly better than `pyteomics` and `pyopenms` methods for MS<sup>1</sup> data and significantly better for MS<sup>2</sup> data, placing it approximately on par with `pymzml` in taking about a tenth of a second.
 The simple database methods (SQLite, DuckDB, and Parquet) also fell in this ~0.1 second range, with SQLite performing most poorly and DuckDB ~10x faster.
@@
 
-Ion chromatogram extraction and retention time range subsets were a key metric for us, corresponding to essential tasks in chromatographic peakpicking and adduct, isotope, and in-source fragment detection (Figure 1, top middle and top right panels).
+Ion chromatogram extraction and retention time range subsets were a key metric for us, corresponding to essential tasks in chromatographic peakpicking and adduct, isotope, and in-source fragment detection (Figure 2B and 2C).
 EIC query times here were universally slower than those for a single spectrum extraction, reflecting the way in which a scan-based file type is sub-optimal for chromatogram extraction because each scan must be parsed to find data within a given *m/z* range.
@@
 However, we also note that both Java-based applications have a slow initial file load step that must be done through a GUI and therefore could not be counted in the timing comparison, the inclusion of which would likely mitigate any advantage for a single chromatogram extraction.
-The mzDB file type is also optimized for chromatogram extraction and was an order of magnitude faster than the other existing file types for which all queries could be run (MzTree and mzMD do not provide interfaces for spectrum extraction or MS2 data).
+The mzDB file type is also optimized for chromatogram extraction and was an order of magnitude faster than the other existing file types for which all queries could be run (MzTree and mzMD do not provide interfaces for spectrum extraction or MS/MS data).
 
@@
 
-Retention time range extraction times were an average of the single-spectrum extraction and the chromatogram extraction times across the board.
-potentially hinting at a major predictive factor in timing estimation being the total amount of scan parsing required.
+Retention time range extraction times were an average of the single-spectrum extraction and the chromatogram extraction times across the board, potentially hinting at a major predictive factor in timing estimation being the total amount of scan parsing required.
 
@@
 We also investigated the efficacy of the various MS data formats for MS/MS data and found that support for fragmentation data searches was lacking or absent from the documentation and exposed functionality of each of these file types, requiring custom implementations every time.
-Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure 2, rightmost panels).
+Despite both precursor searches (where all the precursors of a given fragment are found) and fragment searches (where all the fragments of a given precursor are found) representing intuitive and useful methods of MS/MS data processing, these timings were consistently among the slowest of the six query types for the non-database methods (Figure 2E and 2F).
 
-All existing MS data types required multiple seconds to perform a single fragment search (Figure 2, bottom-right panel), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor.
+All existing MS data types required multiple seconds to perform a single fragment search (Figure 2F), representing a significant bottleneck for any downstream analysis requiring the data associated with the fragments of a given precursor.
 The SQL-based parsers, on the other hand, all took fractions of a second and consistently returned the relevant data hundreds of times more quickly than existing methods.
-The same was true for a precursor search across all methods aside from mzDB (Figure 2, bottom middle panel), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.
+The same was true for a precursor search across all methods aside from mzDB (Figure 2E), which benefited significantly from constructing a single bounding box for all MS/MS information that requires a single decoding into computer memory, though this strategy will fail for any file with sufficiently large MS/MS data.
 
@@
 File size is another important constraint on the efficacy of various MS formats.
-We measured the size on disk of each of the file types and found that they varied by approximately an order of magnitude, with HDF-based file types hovering around one-third the size of the mzML (83 megabytes, MB) while mzDB and mzMD were larger (108 MB and 96 MB, respectively).
-The SQLite object was the largest on disk of all the file types, nearly doubling the mzML's size at 145 MB, while DuckDB was on par with MZA at two-thirds of the mzML (54 MB) and Parquet improved slightly upon that again with its columnar-based storage format (Figure 2).
+We measured the size on disk of each of the file types and found that they varied by approximately an order of magnitude, with HDF-based file types hovering around one-third the size of the mzML (mzML size = 75 megabytes (MB), mzMLb = 18 MB, mz5 = 23 MB) while mzDB and mzMD were larger (95 MB and 99 MB, respectively).
+The SQLite object was the largest on disk of all the file types, nearly tripling the mzML's size at 197 MB, while DuckDB improved on MZA at two-thirds of the mzML (42 MB and 55 MB, respectively) and Parquet improved slightly upon that again (30 MB total) with its columnar-based storage format (Figure 2G).
 
 However, these comparisons are not perfect because not all files store exactly the same data.
-MZTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS<sup>1</sup> scans was 64 MB, a 23% size reduction.
-The SQLite, DuckDB, and Parquet formats also lack the scan and file metadata that's present in the other file types, though it is difficult to estimate the fraction of disk space allocated for this (and which will depend upon the precise definition of metadata).
+MZTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS<sup>1</sup> scans was 58 MB, a 23% size reduction.
+The SQLite, DuckDB, and Parquet formats also lack the extensive scan and file metadata that's present in the other file types, though it is difficult to estimate the fraction of disk space allocated for this (and which will depend upon the precise definition of metadata).
 
@@
 *Supplemental figure XX: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
-Best-fit linear models have been added for each method are shown behind triplicate timing measurements.
+Best-fit linear models have been added for each method and are shown behind triplicate timing measurements, while the black line shows a 1:1 slope intercepting the first data point to show nonlinear behavior with additional chromatograms.
 Transparent intervals around each best-fit line show a single standard error of the mean.*
@@
 
-read_csv("manuscript/data/multifile_timings.csv") %>%
+read_csv("data/multifile_timings.csv") %>%
   mutate(type=factor(type, levels=c("consolidated", "multifile"),
@@
   scale_color_viridis_d(option = "A", begin = 0.4, end = 0.8, aesthetics = c("fill", "color")) +
-  scale_shape_manual(breaks=c("With index/ordering", "No index/ordering"), values=c(21, 22)) +
+  scale_shape_manual(breaks=c("No index/ordering", "With index/ordering"), values=c(22, 21)) +
   labs(x="Database size", y="Time taken", color=NULL, fill=NULL, shape=NULL) +
   theme_bw() +
-  theme(legend.position = "top", legend.direction = "vertical")
-ggsave("manuscript/figures/multifile_fig.png", width = 6.5, height=3.5)
+  theme(legend.position = "top", legend.direction = "vertical") +
+  guides(shape=guide_legend(override.aes = list(size=5, fill="white")),
+         color=guide_legend(override.aes = list(shape=21, size=4, lwd=2)))
+ggsave("figures/multifile_fig.png", width = 6.5, height=3.5)
 ```

### 2025-03-19 README.qmd
Adding manuscript figures to README

@@
 
+![](README_files/figure-manual-manuscript/db_fig.png)
+
 Existing data formats fail to provide intuitive, rapid, and programmatic search of raw MS data and require learning the quirks and conceits of idiosyncratic file formats.
@@
 
+![](README_files/figure-manual-manuscript/singlefile_fig.png)
+
+*Comparison of existing MS file types in query timing and size on disk.
+Six different queries were tried here: extracting an MS1 or MS2 scan (A and D), a chromatogram (B), all data points within a retention time range (C), or fragment/precursor data (E and F).
+Thirteen different parsers were implemented in Python for this comparison, visible in the manuscript_things branch of this repository.*
+
 ## Convert mzML files into a database via the converters above
 
-Converters from mzML to SQLite and DuckDB are currently available in the [converters folder](converters) for R and Python with Julia and C++ hopefully soon.
+Converters from mzML to SQLite and DuckDB are currently available in the [converters folder](converters) for R with Python, Julia, and C++ hopefully soon.
 Code from the associated script can be directly copied and pasted into your workflow, though you're responsible for managing the dependency installations.
-
-The database format is shown in the graphic below:
 
@@
   geom_segment(aes(xend=premz, yend=0)) +
-  facet_wrap(~filename, ncol=1)
+  facet_wrap(~filename, ncol=1) +
+  scale_y_log10()
 ```
@@
 ## Other README things
+
+-----
+
+README last built on `r Sys.Date()`

### 2025-03-19 converters/turn_mzml_duckdb.R
Adding expanded R converters

@@ -1,0 +1,70 @@
+
+library(RaMS)
+library(DBI)
+library(RSQLite)
+library(data.table)
+library(xml2)
+
+turn_mzml_sqlite <- function(ms_files, outfile, ordered=NULL){
+  msdata <- grabMSdata(ms_files)
+  
+  file_info <- msdata$metadata[,setdiff(names(msdata$metadata), "config_data"),with = FALSE]
+  
+  ms_levels <- lapply(ms_files, grabAccessionData, "MS:1000511")
+  ms_file_vec <- rep(basename(ms_files), sapply(ms_levels, nrow))
+  ms_levels <- do.call(what = "rbind", ms_levels)
+  colnames(ms_levels) <- c("name", "ms_level")
+  ms_levels$filename <- ms_file_vec
+  ms_levels$ms_level <- as.numeric(ms_levels$ms_level)
+  ms_levels <- as.data.table(ms_levels)
+  
+  getScanNum <- function(filename){
+    xml_data <- read_xml(filename)
+    cont_nums <- xml_attr(xml_find_all(xml_data, "//d1:spectrum"), "id")
+    scan_nums <- as.numeric(gsub(".*scan=", "", cont_nums))
+    data.table(filename=basename(filename), scan_num=scan_nums)
+  }
+  scan_nums <- lapply(ms_files, getScanNum)
+  scan_nums <- do.call(what = "rbind", scan_nums)
+  scan_levels <- cbind(ms_levels, scan_nums[,filename:=NULL])
+  scan_levels[,name:=NULL]
+  ms1_nums <- scan_levels[ms_level==1,"scan_num"]
+  tic <- cbind(ms1_nums, msdata$TIC)
+  setnames(tic, c("scan_num", "rt", "TIC", "filename"))
+  bpc <- cbind(ms1_nums, msdata$BPC)
+  setnames(bpc, c("scan_num", "rt", "BPC", "filename"))
+  scan_info <- merge(scan_levels, merge(tic, bpc), all.x = TRUE)
+  
+  if(!is.null(ordered)){
+    if(ordered=="mz"){
+      msdata$MS1 <- msdata$MS1[order(mz)]
+      msdata$MS2 <- msdata$MS2[order(premz)]
+    }
+    if(ordered=="rt"){
+      scan_info <- scan_info[order(rt)]
+      msdata$MS1 <- msdata$MS1[order(rt)]
[…]

### 2025-03-19 converters/turn_mzml_sqlite.R

@@ -1,0 +1,74 @@
+
+library(RaMS)
+library(DBI)
+library(RSQLite)
+library(data.table)
+library(xml2)
+
+turn_mzml_sqlite <- function(ms_files, outfile, ordered=NULL){
+  msdata <- grabMSdata(ms_files)
+  
+  simple_cols <- setdiff(names(msdata$metadata), "config_data")
+  file_info <- msdata$metadata[,simple_cols,with = FALSE]
+  
+  ms_levels <- lapply(ms_files, grabAccessionData, "MS:1000511")
+  ms_file_vec <- rep(basename(ms_files), sapply(ms_levels, nrow))
+  ms_levels <- do.call(what = "rbind", ms_levels)
+  colnames(ms_levels) <- c("name", "ms_level")
+  ms_levels$filename <- ms_file_vec
+  ms_levels$ms_level <- as.numeric(ms_levels$ms_level)
+  ms_levels <- as.data.table(ms_levels)
+  
+  getScanNum <- function(filename){
+    xml_data <- read_xml(filename)
+    cont_nums <- xml_attr(xml_find_all(xml_data, "//d1:spectrum"), "id")
+    scan_nums <- as.numeric(gsub(".*scan=", "", cont_nums))
+    data.table(filename=basename(filename), scan_num=scan_nums)
+  }
+  scan_nums <- lapply(ms_files, getScanNum)
+  scan_nums <- do.call(what = "rbind", scan_nums)
+  scan_levels <- cbind(ms_levels, scan_nums[,filename:=NULL])
+  scan_levels[,name:=NULL]
+  ms1_nums <- scan_levels[ms_level==1,"scan_num"]
+  tic <- cbind(ms1_nums, msdata$TIC)
+  setnames(tic, c("scan_num", "rt", "TIC", "filename"))
+  bpc <- cbind(ms1_nums, msdata$BPC)
+  setnames(bpc, c("scan_num", "rt", "BPC", "filename"))
+  scan_info <- merge(scan_levels, merge(tic, bpc), all.x = TRUE)
+  
+  conn <- dbConnect(RSQLite::SQLite(), outfile)
+  dbWriteTable(conn, "file_info", file_info)
+  dbWriteTable(conn, "scan_info", scan_info)
+  dbWriteTable(conn, "MS1", msdata$MS1)
[…]

### 2025-03-19 converters/turn_mzml_duckdb.R
Finished R and Python converters with scan_info and file_info tables

@@ -56,10 +56,12 @@ turn_mzml_sqlite <- function(ms_files, outfile, ordered=NULL){
   return(outfile)
 }
 
+
+# Example usage:
 ms_files <- list.files("demo_data", pattern = "mzML", full.names = TRUE)
 turn_mzml_sqlite(ms_files, "demo_data/msdata.duckdb")
 
-
+# Example chromatogram extraction
 library(DBI)
 pmppm <- function(mass, ppm)c(mass * (1 - ppm/1e+06), mass * (1 + ppm/1e+06))
 

### 2025-03-19 converters/turn_mzml_duckdb.py

@@ -1,0 +1,111 @@
+import duckdb
+import pandas as pd
+import pyteomics.mzml
+import os
+def pmppm(mz, ppm):
+    return(mz*(1-ppm/1000000), mz*(1+ppm/1000000))
+
+def turn_mzml_duckdb(files, outfile, ordered=None):
+    conn = duckdb.connect(outfile)
+    conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS2")
+    conn.execute("DROP TABLE IF EXISTS file_info")
+    conn.execute("DROP TABLE IF EXISTS scan_info")
+    empty_ms1 = pd.DataFrame(columns=["filename", "id", "mz", "int", "rt"]).astype(
+        {"filename": "string", "id": "int64", "mz": "float64", "int": "float64", "rt": "float64"}
+    )
+    conn.execute("CREATE TABLE MS1 AS SELECT * FROM empty_ms1")
+    empty_ms2 = pd.DataFrame(columns=["filename", "id", "premz", "fragmz", "int", "rt"]).astype(
+        {"filename": "string", "id": "int64", "premz": "float64", "fragmz": "float64", "int": "float64", "rt": "float64"}
+    )
+    conn.execute("CREATE TABLE MS2 AS SELECT * FROM empty_ms2")
+    empty_file_info = pd.DataFrame(columns=['filename', 'n_scans', 'rt_start', 'rt_end']).astype(
+        {'filename':'string', 'n_scans':'int64', 'rt_start':'float64', 'rt_end':'float64'}
+    )
+    conn.execute("CREATE TABLE file_info AS SELECT * FROM empty_file_info")
+    empty_scan_info = pd.DataFrame(columns=[
+      'filename', 'rt', 'scan_num', 'min_mz', 'max_mz', 'ms_level', 'centroided', 'polarity'
+    ]).astype(
+        {'filename':"string", 'rt':"float64",
+         'scan_num':"int64", 'min_mz':"float64", 
+         'max_mz':"float64",
+         'ms_level':"int64",
+         'centroided':"bool",
+         'polarity':"string"}
+    )
+    conn.execute("CREATE TABLE scan_info AS SELECT * FROM empty_scan_info")
+
+    
+    
+    if isinstance(files, str):
+        files = [files]
+    for file in files:
+        open_file = pyteomics.mzml.MzML(file)
+        file_info = []
+        n_scans = len(open_file)
[…]

### 2025-03-19 converters/turn_mzml_sqlite.R

@@ -58,16 +58,15 @@ turn_mzml_sqlite <- function(ms_files, outfile, ordered=NULL){
   return(outfile)
 }
 
+# Example usage:
 ms_files <- list.files("demo_data", pattern = "mzML", full.names = TRUE)
 turn_mzml_sqlite(ms_files, "demo_data/msdata.sqlite", ordered = "mz")
 
-
-
-
+# Example chromatogram extraction
 library(DBI)
 pmppm <- function(mass, ppm)c(mass * (1 - ppm/1e+06), mass * (1 + ppm/1e+06))
 
-conn <- dbConnect(RSQLite::SQLite(), "demo_data/demo_data.sqlite")
+conn <- dbConnect(RSQLite::SQLite(), "demo_data/msdata.sqlite")
 chr_statement <- "SELECT * FROM MS1 WHERE mz BETWEEN ? AND ? ORDER BY filename, rt"
 chrom <- dbGetQuery(conn, chr_statement, params=pmppm(118.0865, 10))
 dbDisconnect(conn)

### 2025-03-19 converters/turn_mzml_sqlite.py

@@ -1,0 +1,80 @@
+import sqlite3
+import pandas as pd
+import pyteomics.mzml
+import os
+def pmppm(mz, ppm):
+    return(mz*(1-ppm/1000000), mz*(1+ppm/1000000))
+
+def turn_mzml_sqlite(files, outfile, ordered=None):
+    conn = sqlite3.connect(outfile)
+    conn.execute("DROP TABLE IF EXISTS MS1")
+    conn.execute("DROP TABLE IF EXISTS MS2")
+    if isinstance(files, str):
+        files = [files]
+    for file in files:
+        open_file = pyteomics.mzml.MzML(file)
+        file_info = []
+        n_scans = len(open_file)
+        min_rt = open_file[0]["scanList"]["scan"][0]["scan start time"]
+        max_rt = open_file[n_scans-1]["scanList"]["scan"][0]["scan start time"]
+        file_info.append(pd.DataFrame({'filename':[os.path.basename(file)], 'n_scans':n_scans,
+                                       'rt_start':min_rt, 'rt_end':max_rt}))
+        # Might need to switch to pymzml for more metadata...
+    
+        scan_info = []
+        MS1_dfs = []
+        MS2_dfs = []
+        for spectrum in open_file:
+            rt_val = spectrum['scanList']['scan'][0]['scan start time']
+            idx = int(spectrum['id'].split("scan=")[-1].split()[0])
+            scan_info.append(pd.DataFrame({'filename':[os.path.basename(file)], 'rt':rt_val,
+                                           'scan_num':idx, 'min_mz':spectrum['lowest observed m/z'], 
+                                           'max_mz':spectrum['highest observed m/z'],
+                                           'ms_level':spectrum["ms level"],
+                                           'centroided':'centroid spectrum' in demo_lib,
+                                           'polarity':("negative", "positive")['positive scan' in demo_lib]}))
+            
+            if spectrum['ms level'] == 1:
+                mz_vals = spectrum['m/z array']
+                int_vals = spectrum['intensity array']
+                df_scan = pd.DataFrame({'filename': os.path.basename(file), 'id': idx, 'mz': mz_vals, 
+                                        'int': int_vals, 'rt': [rt_val] * len(mz_vals)})
+                MS1_dfs.append(df_scan)
+            if spectrum['ms level'] == 2:
[…]

### 2025-03-20 README.qmd
Updating README with new examples

@@
 
-Ensure `RaMS`, `DBI`, and `duckdb`/`RSQLite` are all installed from CRAN.
+Some simple demo code shown below for MS1 and MS2 table creation to illustrate what the converters do:
 
 ```{r make database, eval=FALSE}
+# Ensure `RaMS`, `DBI`, and `duckdb`/`RSQLite` are all installed from CRAN.
 library(RaMS)
@@
 # conn <- dbConnect(duckdb::duckdb(), "demo_data/msdata.duckdb")
-chr_statement <- "SELECT * FROM MS1 WHERE mz BETWEEN ? AND ?"
+chr_statement <- "SELECT * FROM MS1 WHERE mz BETWEEN ? AND ? ORDER BY filename, rt"
 chrom <- dbGetQuery(conn, chr_statement, params=pmppm(118.0865, 10))

### 2025-06-10 metsoc_poster/chrom_script.jl
Adding metabolomics 2025 poster

@@ -1,0 +1,26 @@
+using SQLite, DataFrames, StatsPlots
+
+db = SQLite.DB("chrom_data/chrom_data.sqlite")
+targets = [147.0764, 148.0734, 149.0705]
+window(m) = (m*(1 - 2/1e6), m*(1 + 2/1e6))
+clause = join(["(mz BETWEEN $(mzmin) AND $(mzmax))" for (mzmin, mzmax) in window.(targets)], " OR ")
+query = "SELECT * FROM MS1 WHERE ($clause) AND (rt BETWEEN 10 AND 12)"
+df = DBInterface.execute(db, query) |> DataFrame
+SQLite.close(db)
+
+p1 = @df df[round.(df.mz) .== 147, :] plot(:rt, :int, group=:filename, color="black", title="Unlabeled")
+p2 = @df df[round.(df.mz) .== 148, :] plot(:rt, :int, group=:filename, color="#0b505c", title="15Nx1")
+p3 = @df df[round.(df.mz) .== 149, :] plot(:rt, :int, group=:filename, color="#028e34", title="15Nx2")
+plot(p1, p2, p3, layout = (3,1), dpi=600)
+
+
+using LaTeXStrings, Plots
+default(fontfamily="Computer Modern")
+p1 = @df df[round.(df.mz) .== 147, :] plot(:rt, :int, group=:filename, color="black", legend=false, 
+title="Unlabeled", xlims=(10, 12), titlefontsize = 10)
+p2 = @df df[round.(df.mz) .== 148, :] plot(:rt, :int, group=:filename, color="#0b505c", legend=false, 
+title=L"\textrm{^{15}N_1\ isotope}", xlims=(10, 12), titlefontsize = 10)
+p3 = @df df[round.(df.mz) .== 149, :] plot(:rt, :int, group=:filename, color="#028e34", legend=false, 
+title=L"\textrm{^{15}N_2\ isotope}", xlims=(10, 12), titlefontsize = 10)
+plot(p1, p2, p3, layout = (3,1), dpi=600, size=(600, 400))
+savefig("chrom_sqlite_julia_fig.png")

### 2025-06-10 metsoc_poster/db_maker.R

@@ -1,0 +1,177 @@
+
+library(tidyverse)
+library(RaMS)
+
+# Write imzML data to Arrow for R ----
+# 10.1007/s00216-012-5841-x
+# Maybe also https://www.ebi.ac.uk/metabolights/editor/MTBLS313/descriptors?
+library(Cardinal)
+msi_files <- list.files("imzml_data", pattern = ".imzML", full.names = TRUE)[1:3]
+all_msi <- pbapply::pblapply(msi_files, function(file_i){
+  msi <- readImzML(file_i)
+  
+  coords <- coord(msi)
+  mz_list <- as.list(mz(msi))
+  int_list <- as.list(intensity(msi))
+  
+  result <- data.frame(
+    filename = basename(file_i),
+    scan = rep(seq_along(mz_list), lengths(mz_list)),
+    x = rep(coords$x, lengths(mz_list)),
+    y = rep(coords$y, lengths(mz_list)),
+    mz = unlist(mz_list),
+    int = unlist(int_list)
+  )
+  return(result)
+})
+v <- do.call("rbind", all_msi)
+write_dataset(v, "imzml_data/pqds")
+
+lapply(all_msi, function(df_i){
+  df_i %>%
+    arrange(desc(int)) %>%
+    head()
+})
+
+# Write ms3 data to DuckDB for Python ----
+# https://www.metabolomicsworkbench.org/data/DRCCStudySummary.php?Mode=SetupRawDataDownload&StudyID=ST001661
+# 10.1021/acs.analchem.0c04895
+library(DBI)
+ms3_files <- list.files("ms3_data", pattern = "mzML", full.names = TRUE)
+new_grabMzmlMS3 <- function (xml_data, rtrange, file_metadata, incl_polarity) {
+  ms3_xpath <- "//d1:spectrum[d1:cvParam[@name=\"ms level\" and @value=\"3\"]]"
+  ms3_nodes <- xml2::xml_find_all(xml_data, ms3_xpath)
+  rt_vals <- RaMS:::grabSpectraRt(ms3_nodes)
+  fsnodes <- xml2::xml_find_all(ms3_nodes, "d1:scanList/d1:scan/d1:cvParam[@name=\"filter string\"]")
+  shortstrings <- gsub("FTMS \\+ c NSI d Full ms3 |@hcd30.00.*", "", xml_attr(fsnodes, "value"))
[…]

### 2025-06-10 metsoc_poster/imzml_script.R

@@ -1,0 +1,28 @@
+library(tidyverse)
+library(RaMS)
+arrow::open_dataset("imzml_data/pqds") %>%
+  filter(mz%between%pmppm(421.3111, 10) | mz%between%pmppm(281.2484, 10)) %>%
+  mutate(mz=paste("m/z =", round(mz, 2))) %>%
+  collect() %>%
+  ggplot(aes(x, y, fill=int)) + geom_raster() + facet_grid(mz~filename)
+
+
+out_gp <- arrow::open_dataset("imzml_data/pqds") %>%
+  filter(mz%between%pmppm(421.3111, 10) | mz%between%pmppm(281.2484, 10)) %>%
+  mutate(mz=paste("m/z =", round(mz, 2))) %>%
+  mutate(filename=str_remove(filename, "-centroid")) %>%
+  mutate(filename=str_replace(filename, " S", "\n S")) %>%
+  collect() %>%
+  ggplot() +
+  geom_raster(aes(x, y, fill=log10(int))) +
+  ggh4x::facet_grid2(mz~filename, scales="free", independent="y") +
+  scale_fill_gradientn(colors=viridis::viridis(10, direction = -1), 
+                       breaks=c(1:4), labels=10^(1:4), name="Intensity") +
+  scale_x_continuous(expand = expansion(), name = "DESI IMS X-coordinate") +
+  scale_y_continuous(expand = expansion(), name = "DESI IMS Y-coordinate") +
+  theme_bw() +
+  theme(legend.position = "inside",
+        legend.position.inside = c(1, 0.5),
+        legend.justification = c(1, 0.5),
+        legend.background = element_rect(fill="#FFFFFF", color="black"))
+ggsave("imzml_arrow_R_fig.png", plot = out_gp, device = "png", width = 6, height = 5, units = "in", dpi = 600)

### 2025-06-10 metsoc_poster/ms3_script.py

@@ -1,0 +1,59 @@
+import duckdb, matplotlib.pyplot as plt, pandas as pd
+
+con = duckdb.connect("ms3_data/ms3_data.duckdb")
+ms1 = con.execute("SELECT * FROM MS1 WHERE mz BETWEEN 282.1186 AND 282.1208").fetchdf()
+ms2 = con.execute("SELECT * FROM MS2 WHERE premz BETWEEN 282.1186 AND 282.1208 ORDER BY int DESC").fetchdf()
+ms3 = con.execute("""SELECT * FROM MS3 WHERE prepremz BETWEEN 282.1186 AND 282.1208 AND 
+                  premz BETWEEN 166.0718 AND 166.0731""").fetchdf()
+con.close()
+
+# fig, axs = plt.subplots(3, 1, figsize=(9, 6))
+# axs[0].plot(ms1['rt'], ms1['int'])
+# axs[0].vlines(ms2['rt'], ymin=0, ymax=ms1['int'].max())
+# axs[1].stem(ms2['fragmz'], ms2['int'])
+# axs[2].stem(ms3['fragmz'], ms3['int'])
+# plt.show()
+
+ms1["int"] /= 1e6
+ms2["int"] /= 1e6
+ms3["int"] /= 1e6
+
+fig, axs = plt.subplots(3, 1, figsize=(9, 6), sharex=False)
+fig.subplots_adjust(hspace=0.4)
+
+axs[0].plot(ms1['rt'], ms1['int'], color="black", lw=1)
+axs[0].vlines(ms2['rt'], ymin=0, ymax=ms1['int'].max(), color="#a41118", lw=1.5)
+axs[0].vlines(sorted(ms2['rt'].unique())[2], ymin=0, ymax=ms1['int'].max(), color="#e0670b", lw=1.5)
+ms1_lab = """MS$^1$ Extracted
+Ion Chromatogram"""
+axs[0].text(0.01, 0.95, ms1_lab, va="top", ha="left", transform=axs[0].transAxes)
+axs[0].set_xlabel("Retention time")
+axs[0].set_ylabel(r"Intensity ($10^6$)")
+
+axs[1].stem(ms2['fragmz'], ms2['int'], linefmt="#e0670b", basefmt='k')
+axs[1].stem(ms2['fragmz'][0], ms2['int'][0], linefmt="#f4bb23", basefmt='k')
[…]

### 2025-06-10 metsoc_poster/timing_size_script.R

@@ -1,0 +1,65 @@
+
+library(tidyverse)
+library(ggtext)
+library(patchwork)
+
+method_names <- c("pyteomics", "pymzml", "pyopenms", "2DPeak",
+                  "mzMLb", "mzDB", "MZA", "mz5", 
+                  "MZTree", "mzMD", 
+                  "SQLite", "DuckDB", "Parquet")
+method_colors <- c("#393B79", "#5254A3", "#6B6ECF", "#9C9EDE", 
+                   "#0F8299", "#637939", "#E6550D", "#843C39", 
+                   "#7B4173", "#CE6DBD", 
+                   "#8C6D31", "#E7BA52", "#efddb9")
+backup_colors <- c("#393B79", "#5254A3", "#6B6ECF", "#9C9EDE", 
+                   "#0F8299", "#637939", "#E6550D", "#843C39", 
+                   "#7B4173", "#CE6DBD", 
+                   "#8C6D31", "#E7BA52", "grey20")
+
+read_csv("timing_data/singlefile_times.csv") %>% 
+  mutate(method=str_remove(method, "pyopenms_")) %>%
+  mutate(method=factor(method, levels=method_names)) %>%
+  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
+                                      "ms2_spec", "premz", "fragmz"),
+                      labels=c("A. Full scan", "B. Chromatogram", "C. RT Range",
+                               "D. MS/MS scan", "E. Precursor search",
+                               "F. Fragment search"))) %>%
+  ggplot() +
+  geom_boxplot(aes(x=method, y=time, color=method)) +
+  facet_wrap(~query) +
+  scale_y_log10() +
+  scale_color_manual(breaks = method_names, values=method_colors) +
+  labs(x=NULL, y="Time (seconds)", color=NULL) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
+        legend.position = "none",
+        strip.text.x = element_text(hjust = 0),
+        text=element_text(size=24))
+ggsave("timing_boxplots.png", width = 9.5, height = 6, dpi=600)
+
+read_csv("timing_data/file_sizes.csv") %>%
+  mutate(method=str_remove(method, "pyopenms_")) %>%
+  mutate(method=factor(method, levels=method_names)) %>%
[…]

### 2025-06-10 metsoc_poster/db_maker.R
First draft edits, sent to collabs

@@ -3,7 +3,7 @@ library(tidyverse)
 library(RaMS)
 
 # Write imzML data to Arrow for R ----
-# 10.1007/s00216-012-5841-x
+# dx.doi.org/10.1007/s00216-012-5841-x
 # Maybe also https://www.ebi.ac.uk/metabolights/editor/MTBLS313/descriptors?
 library(Cardinal)
 msi_files <- list.files("imzml_data", pattern = ".imzML", full.names = TRUE)[1:3]
@@ -35,7 +35,7 @@ lapply(all_msi, function(df_i){
 
 # Write ms3 data to DuckDB for Python ----
 # https://www.metabolomicsworkbench.org/data/DRCCStudySummary.php?Mode=SetupRawDataDownload&StudyID=ST001661
-# 10.1021/acs.analchem.0c04895
+# dx.doi.org/10.1021/acs.analchem.0c04895
 library(DBI)
 ms3_files <- list.files("ms3_data", pattern = "mzML", full.names = TRUE)
 new_grabMzmlMS3 <- function (xml_data, rtrange, file_metadata, incl_polarity) {

### 2025-06-10 manuscript/manuscript.Rmd
Adding Bryna edits

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, include=FALSE, eval=FALSE)
+knitr::opts_chunk$set(echo = TRUE, include=FALSE, eval=TRUE)
 library(tidyverse)
@@
 
-method_names <- c("pyteomics", "pymzml", "pyopenms", "pyopenms_2DPeak",
+method_names <- c("pyteomics", "pymzml", "pyopenms", "2DPeak",
                   "mzMLb", "mzDB", "MZA", "mz5", 
@@
 Relational databases also have several distinct advantages over hierarchical or text-based systems, particularly in performing searches for subsets of data via indices.
-Importantly, this indexing differs from the byte-offset indexes that already exist in the indexed mzML and HDF5 formats because the search for a particular subset cannot be done efficiently with a byte-offset index when the *m/z* data is encoded, though access to a particular scan can be incredibly rapid.
-Additionally, data from multiple samples can be stored together in a single database table, differing from formats like mzDB, mzTree, and mzMD and allowing queries of all dataset samples to be performed without looping through each file in turn, thereby avoiding the associated computational overhead and query complexity.
+Importantly, this indexing differs from the byte-offset indexes that already exist in the indexed mzML and HDF5 formats because the search for a particular subset cannot be done efficiently with a byte-offset index when the *m/z* data is encoded (whether by numpress, zlib, or just base64), though access to a particular scan can be incredibly rapid.
+Additionally, data from multiple samples can be stored together in a single database table to permit queries of all dataset samples to be performed without looping through each file in turn.
+This differs from existing formats like mzDB, mzTree, and mzMD and thereby avoids the associated computational overhead and query complexity.
 
@@
 Full scan queries consisted of 1) single scan extraction by scan number, 2) retention time range extraction of all scans within a specified retention time range, and 3) chromatogram extraction, which collects the ions within a specified parts-per-million (PPM) error of a known mass.
-These queries generally correspond to the methods used in @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser.
+These queries generally correspond to the methods used in Bouyssié et al.
+(2015) @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser.
 Note that the chromatogram extraction does not extract a precompiled chromatogram of the sort commonly found at the end of mzML files or as a result of SIM/PRM analysis but instead refers to sifting through the raw data for data tuples with an *m/z* value between specified bounds.
@@
 SQL databases were built using Python 3.11.11 with SQLite (v3.48.0) via the Python sqlite3 package (v2.6.0), DuckDB via the duckdb package (v1.1.3), and Parquet files via the pyarrow package (v19.0.0).
-We were additionally able to download and install [blah]
 
@@
 Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure 1).
-We do not force compression of any of these fields because we believe that decoding compressed data is both a slow and unintuitive step, though automatic compression is supplied by the DuckDB and Parquet file types.
+We do not force compression of any of these fields because we find that decoding compressed data is both a slow and unintuitive step, though automatic compression is supplied by the DuckDB and Parquet file types.
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "manuscript/figures/fileinfo_table.png", vwidth=1000)
+  gtsave(filename = "figures/fileinfo_table.png", vwidth=1000)
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "manuscript/figures/scaninfo_table.png", vwidth=1000)
+  gtsave(filename = "figures/scaninfo_table.png", vwidth=1000)
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "manuscript/figures/ms1_table.png", vwidth=1000)
+  gtsave(filename = "figures/ms1_table.png", vwidth=1000)
 
@@
   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "manuscript/figures/ms2_table.png", vwidth=1000)
+  gtsave(filename = "figures/ms2_table.png", vwidth=1000)
 ```
@@
 ```{r}
-
+library(gt)
+# data.frame(
+#   filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+#   scan_num=c(1,1,651501,651501),
+#   rt=c(0.10, 0.10, 12.37, 12.37),
+#   dt=c(),
+#   mz=c(60.0452, 60.0532, 60.0456, 60.0531),
+#   int=c(6618, 2657, 158084, 4673)
+# ) %>%
+#   gt() %>%
+#   cols_label(filename=html("filename</br><small>Source file</small>"),
+#              scan_num=html("scan_num</br><small>Scan number</small>"),
+#              rt=html("rt</br><small>Retention time</small>"), 
+#              mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
+#              int=html("int</br><small>Intensity</small>")) %>%
+#   tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
+#   tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+#   tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+#   opt_table_outline() %>%
+#   tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+#   cols_align(align = "center") %>%
+#   cols_move_to_start("filename") %>%
+#   tab_options(row_group.padding = 3) %>%
+#   tab_header(title = "Table: MS1") %>%
+#   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+#   gtsave(filename = "figures/ms1_table.png", vwidth=1000)
 ```
@@
 
-We settled on a large dataset of gut microbiota LC-MS files published in @Portlock2025 and available on Metabolights under accession number MTBLS10066.
+We settled on a large dataset of gut microbiota LC-MS files published in Portlock et al. (2025) @Portlock2025 and available on Metabolights under accession number MTBLS10066.
 
@@
 
-Despite the relative simplicity and relevance of our queries, none of the available mass spectrometry (MS) formats had existing functions or documented examples of all six queries.
+Despite the relative simplicity and relevance of our queries, none of the available mass spectrometry (MS) formats had existing functions or examples of all six queries.
 Unsurprisingly, the mzML file type had the most extensive coverage but documentation and prebuilt functionality was still disappointingly sparse.
@@
 In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D).
-Here, we found that the mzML and mzMLb file types were consistently the slowest to parse the mzML file type and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
+Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
 `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX).
@@
 all_timings <- read_csv("data/singlefile_times.csv") %>% 
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
   mutate(method=factor(method, levels=method_names)) %>%
@@
 all_sizes <- read_csv("data/file_sizes.csv") %>%
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
   mutate(method=factor(method, levels=method_names))
@@
   geom_boxplot(aes(x=method, y=time, color=method)) +
-  facet_wrap(~query) +
+  facet_wrap(~query, ncol=2) +
   scale_y_log10() +
@@
 size_barplot <- all_sizes %>%
-  filter(!method%in%c("pyteomics", "pyopenms", "pyopenms_2DPeak")) %>%
+  filter(!method%in%c("pyteomics", "pyopenms", "2DPeak")) %>%
   mutate(method=as.character(method)) %>%
@@
 *Figure 2: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper.
-The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum (A), an ion chromatogram (B), all data within a retention time range (C), an MS/MS scan (D), the fragments of a specified precursor (E), and all precursors with a specified fragment (F).
+The left six panels show boxplots representing the time required in seconds to extract a full scan spectrum (A), an ion chromatogram (B), all data within a retention time range (C), an MS/MS scan (D), the fragments of a specified precursor (E), and all precursors with a specified fragment (F).
 The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z.
@@
                "<span style='color:#0b505c'>Extraction only</span>")
-init_outside_gp <- read_csv("manuscript/data/singlefile_supp.csv") %>%
+init_outside_gp <- read_csv("data/singlefile_supp.csv") %>%
   pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
@@
 
-pyopenms_precomp_gp <- read_csv("manuscript/data/pyopenms_precomp.csv") %>%
+pyopenms_precomp_gp <- read_csv("data/pyopenms_precomp.csv") %>%
   pivot_longer(rep1:rep3) %>%
@@
 init_outside_gp / pyopenms_precomp_gp
-ggsave("manuscript/figures/supp_preload_fig.png", width = 6.5, height=5)
+ggsave("figures/supp_preload_fig.png", width = 6.5, height=5)
 ```
@@
 ```{r}
-multichrom_times <- read_csv("manuscript/data/multichrom_times.csv")
+multichrom_times <- read_csv("data/multichrom_times.csv")
 timelabs <- c("0.01 seconds", "One second", "One minute", "One hour")
 multichrom_times %>%
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
   mutate(method=factor(method, levels=method_names)) %>%
   ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
-  geom_abline(slope=1, intercept=-3:3, color="black") +
+  geom_abline(slope=1, intercept=-3:3, color="grey50") +
   geom_smooth(method="lm", formula="y~x", alpha=0.2, fill="grey50") +
@@
   theme(panel.grid.minor = element_blank())
-ggsave("manuscript/figures/multichrom_fig.png", width = 6.5, height=5)
+ggsave("figures/multichrom_fig.png", width = 6.5, height=5)
 ```
@@
 ```{r}
-multichrom_times <- read_csv("manuscript/data/multichrom_times.csv")
+multichrom_times <- read_csv("data/multichrom_times.csv")
 multichrom_baseline_df <- multichrom_times %>%
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
   filter(n_chroms==1) %>%
@@
 multichrom_times %>%
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
   mutate(method=factor(method, levels=method_names)) %>%
@@
   theme(legend.position = "none")
-ggsave("manuscript/figures/supp_multichrom_facets.png", width = 6.5, height=5)
+ggsave("figures/supp_multichrom_facets.png", width = 6.5, height=5)
 ```
@@
                "<span style='color:#147C8F'>Non-equi join</span>")
-read_csv("manuscript/data/dbloopjoinuni_times.csv") %>%
+read_csv("data/dbloopjoinuni_times.csv") %>%
   separate(method, into = c("method", "type")) %>%
@@
         plot.caption.position =  "plot")
-ggsave("manuscript/figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
+ggsave("figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
 ```
@@
 
+```{r}
+read_csv("data/multifile_timings.csv") %>%
+  mutate(size=size/1e6) %>%
+  summarise(size=mean(size), .by=c(method, n_files, db_sort)) %>%
+  pivot_wider(names_from = db_sort, values_from=size) %>%
+  mutate(diff=ordered/unordered) %>%
+  arrange(method)
+```
+
 We found that SQLite queries benefitted significantly from the construction of an index on the *m/z* column when extracting chromatograms, improving lookup times by an order of magnitude (dropping from [0.3] seconds to [0.03 seconds], Figure 4).
-However, because the SQLite index is stored on disk alongside the data, this improvement also increased the file size by 33% from [150MB] to [210MB].
-DuckDB also improved significantly with ordered data but to a smaller degree and surprisingly required additional space to do so, likely due to the reordering resulting in a different compression strategy (Figure 4).
-Parquet files had the smallest improvement upon data ordering but remained the same small size they were when unordered.
+However, because the SQLite index is stored on disk alongside the data, this improvement also increased the file size by 24%.
+DuckDB also improved significantly with ordered data but to a smaller degree and required additional space to do so, likely due to the reordering resulting in worse compression in other columns such as filename or retention time (Figure 4).
+Parquet files had the smallest improvement upon data ordering and also required 50% more space.
 
@@
 SQLite had much higher variance and slower extraction times with datasets consisting of more than one file, typically an order of magnitude slower than DuckDB, while Parquet fell between the two.
-Importantly, only DuckDB had a slope significantly less than one.
+Importantly, only DuckDB had a slope much less than one.
 This is what would be expected if the database was performing a simple binary search on the index, with an expected time efficiency of O(log(# of files)).
-However, DuckDB's performance degrades at larger database sizes and approaches a 1:1 slope, possibly due to the overhead of reading large amounts of data into memory after it's found.
+However, DuckDB's performance degrades at larger database sizes and approaches a 1:1 slope, possibly due to the time required to read large amounts of data into memory after it's found.
 We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure 4).
@@
 
-@Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers.
+Wilhelm et al.
+(2012) @Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers.
 Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of two times smaller (50%).
 
-@Bouyssie2015 also found that mz5 was about 80% smaller than the mzML and on par with their mzDB format.
+Bouyssié et al.
+(2015) @Bouyssie2015 also found that mz5 was about 80% smaller than the mzML and on par with their mzDB format.
 They reported query times of around 30 seconds for a wide (5 Da) ion chromatogram extraction and found that mz5 was about 40 times slower and that mzML was 200 times slower, in contrast to our parser which was 10 times slower for mz5 - possibly due to the large EIC width.
@@
 
-Just as the original database paper from @Codd1970 argued that the same problems were being solved over and over again, mass spectrometry data scientists are re-solving problems that have been more elegantly ironed out by dedicated teams in computer science and industry with much more extensive support.
+Just as the original database paper from Codd (1970) @Codd1970 argued that the same problems were being solved repeatedly, mass spectrometry data scientists are re-solving problems that have been more elegantly ironed out by dedicated teams in computer science and industry with much more extensive support.
 By leveraging existing optimizations in SQLite and DuckDB, we were able to create a highly performant system for storage of MS data that does not come with significant trade-offs between data extraction methods.
@@
 full scan only versus MS/MS or metadata requirements), 4) whether the entire file will be processed or only a subset, and 5) how well a file type interfaces with downstream software.
-We intend to show with this manuscript that there is significant overlap between the goals of organizations much larger than any individual lab and that MS as a whole can benefit significantly from co-opting their development.
+We intend to show with this manuscript that there is significant overlap between the goals of organizations much larger than any individual lab and that mass spectrometrists can benefit significantly from co-opting their development.
 
@@
 
-[add R/Julia comparison?]
-
 # Acknowledgements

### 2025-06-10 manuscript/manuscript.Rmd
Adding edits from Sam

@@
 We compare the time and space required to extract a representative data subset under six conditions and perform these tests on multiple databases as well as mzML and other MS data formats.
+Our specific questions were:
+
+1. Is there a simple database schema that enables exploratory MS queries with basic SQL statements?
+2. How expensive in time and (disk) space is it to access MS data in SQLite, DuckDB, and Parquet formats?
+3. How does this cost compare to more complex formats that have been previously proposed?
 
@@
 The pyteomics package provides four "combined examples" that focus on the spectrum visualization and annotation common to proteomics research but provide minimal guidance about chromatogram or retention time range extraction.
-pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb.
+Pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb.
 The pyopenms package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS<sup>1</sup> data.
 Additionally, pyOpenMS required installing an old version of the package (3.0.0), Python itself (3.11) and the numpy package (<2.0) due to more recent builds requiring AVX support which was unavailable on our hardware.
-pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing (February 2025).
+Pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing (February 2025).
 
@@
 This required that we deduce the SQLite BLOB type compression format from scratch when writing a parser and spend extensive time reading through the documentation to determine how best to link the various tables provided in the mzDB file.
-Scan metadata in this file type is stored as raw XML strings, producing the worst of both worlds in requiring both SQLite knowledge in their extraction and XML processing to obtain the relevant information.
-Additionally, its failure to implement the clever bounding box and run slice scheme for MS/MS data negated with our ability to avoid parsing every MS/MS spectrum when performing precursor and fragment searches.
+Scan metadata in this file type is stored as raw XML strings, producing the worst of both worlds in requiring both SQLite knowledge in their extraction and XML processing to obtain the necessary information.
+Additionally, mzDB seems to dump all MS/MS data into a single bounding box, meaning that we were unable to use the scheme to avoid parsing every MS/MS spectrum when performing precursor and fragment searches.
 
@@
 `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX).
-It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number as scans are not always consecutive (e.g.
-during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan.
+It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number.
+Scans are not always consecutive (e.g.
+during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan number.
 
@@
 Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3).
-pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
+Pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
 Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).
@@
 Finally, a complexity penalty must be noted for formats and packages requiring complex installation procedures.
-pyopenms appears to be the worst offender here, with its bindings to the OpenMS C++ libraries requiring us to step back to Python 3.11 and numpy 1.26 to successfully access the data on our setup and required direct input from the maintainers.
+Pyopenms appears to be the worst offender here, with its bindings to the OpenMS C++ libraries requiring us to step back to Python 3.11 and numpy 1.26 to successfully access the data on our setup and required direct input from the maintainers.
 The Java applications for MzTree and, more egregiously mzMD, provided essentially zero documentation on installation consisting of a single README file without an intact HTTP request example and have not seen updates in years.
@@
 (2012) @Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers.
-Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of two times smaller (50%).
+Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of half the mzML size.
 
@@
 The mzMLb group [@Bhamber2021] reports only info for spectrum access at approximately 15ms/scan which agrees with our 5-100ms/scan estimates only if the data is loaded ahead of time.
-They perform extensive comparisons to mzML at varying compression methods and levels but our default options of 1024 KB chunk sizes for the mzMLb file and zlib but not numpress compression for the mzML resulting in timing values very similar between mzML and mzMLb using the pyteomics library for access.
+They perform extensive comparisons to mzML at varying compression methods and levels but we stuck with the default options of 1024 KB chunk sizes for the mzMLb file and zlib-only compression for the mzML.
+This resulted in timing values very similar to those they reported when using the pyteomics library for access.
 They also compared to the mz5 file type and we are able to validate their results of the mzML+zlib occupying significantly more space, though our mz5 parser outperformed theirs for full scan and MS/MS data by two orders of magnitude in time.
@@
 We were also unable to test several other recent and promising formats.
-Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF, @Beagley2009) from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
+Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
 

### 2025-06-10 manuscript/manuscript.Rmd
Numbering supplemental figures

@@
 
-```{r}
+```{r supp fig 1 gtables, eval=FALSE}
 library(gt)
@@
 
-*Supplemental figure XX: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
+*Supplemental figure 1: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
 
@@
 Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
-`pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX).
+`pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure 2).
 It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number.
@@
 init_outside_gp / pyopenms_precomp_gp
-ggsave("figures/supp_preload_fig.png", width = 6.5, height=5)
+ggsave("figures/supp_2_preload_fig.png", width = 6.5, height=5)
 ```
 
-![](figures/supp_preload_fig.png)
+![](figures/supp_2_preload_fig.png)
 
-*Supplemental figure XX: Timing information with and without pre-loading the object into Python prior to extraction.
+*Supplemental figure 2: Timing information with and without pre-loading the object into Python prior to extraction.
 Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types.
@@
 Pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
-Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).
+Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure 3).
 
-```{r}
+```{r supp fig 3 multichrom}
 multichrom_times <- read_csv("data/multichrom_times.csv")
@@
   theme(legend.position = "none")
-ggsave("figures/supp_multichrom_facets.png", width = 6.5, height=5)
+ggsave("figures/supp_3_multichrom_facets.png", width = 6.5, height=5)
 ```
 
-![](figures/supp_multichrom_facets.png)
+![](figures/supp_3_multichrom_facets.png)
 
-*Supplemental figure XX: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
+*Supplemental figure 3: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
 Best-fit linear models have been added for each method and are shown behind triplicate timing measurements, while the black line shows a 1:1 slope intercepting the first data point to show nonlinear behavior with additional chromatograms.
@@
 
-We also explored whether database queries could be improved via the use of either a unified query (single SQL statement with multiple OR clauses for each ion's *m/z* range) or a non-equi join between a peak table with *m/z* minimum and maximum columns (Supplemental Figure XX).
+We also explored whether database queries could be improved via the use of either a unified query (single SQL statement with multiple OR clauses for each ion's *m/z* range) or a non-equi join between a peak table with *m/z* minimum and maximum columns (Supplemental Figure 4).
 SQLite and Parquet performed ~3-5 times faster with the unified query than with the loop method despite the necessity of and additional processing step for the looped query to correctly assign each data point to its original peak information.
@@
 
-```{r}
+```{r supp fig 4 dbloopjoin}
 type_labels = c("<span style='color:#a41118'>For loop</span>", 
@@
         plot.caption.position =  "plot")
-ggsave("figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
+ggsave("figures/supp_4_dbloopjoinuni.png", width = 6.5, height=5)
 ```
 
-![](figures/supp_dbloopjoinuni.png)
+![](figures/supp_4_dbloopjoinuni.png)
 
-*Supplemental figure XX: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color.
+*Supplemental figure 4: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color.
 Ten replicates are shown for each data point.*
@@
 Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number.
-The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supp Fig XX).
+The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supplemental Figure 1).
 

### 2025-06-11 manuscript/manuscript.Rmd
Finished supp. figure 1

@@
 library(gt)
-# data.frame(
-#   filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
-#   scan_num=c(1,1,651501,651501),
-#   rt=c(0.10, 0.10, 12.37, 12.37),
-#   dt=c(),
-#   mz=c(60.0452, 60.0532, 60.0456, 60.0531),
-#   int=c(6618, 2657, 158084, 4673)
-# ) %>%
-#   gt() %>%
-#   cols_label(filename=html("filename</br><small>Source file</small>"),
-#              scan_num=html("scan_num</br><small>Scan number</small>"),
-#              rt=html("rt</br><small>Retention time</small>"), 
-#              mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
-#              int=html("int</br><small>Intensity</small>")) %>%
-#   tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
-#   tab_row_group(label = "", rows = rt<=mean(rt)) %>%
-#   tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
-#   opt_table_outline() %>%
-#   tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
-#   cols_align(align = "center") %>%
-#   cols_move_to_start("filename") %>%
-#   tab_options(row_group.padding = 3) %>%
-#   tab_header(title = "Table: MS1") %>%
-#   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-#   gtsave(filename = "figures/ms1_table.png", vwidth=1000)
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,651501,651501),
+  rt=c(0.10, 0.10, 12.37, 12.37),
+  dt=c(0.07, 0.07, 6.16, 6.16),
+  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
+  int=c(6618, 2657, 158084, 4673)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             rt=html("rt</br><small>Retention time</small>"),
+             dt=html("dt</br><small>Drift time</small>"),
+             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
+             int=html("int</br><small>Intensity</small>")) %>%
+  tab_row_group(label = "...billions of additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: MS1") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/ionmob_table.png", vwidth=1000)
+
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,651501,651501),
+  x_coord=c(1, 1, 678, 678),
+  y_coord=c(1, 1, 327, 327),
+  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
+  int=c(6618, 2657, 158084, 4673)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             x_coord=html("x_coord</br><small>X coordinate</small>"),
+             y_coord=html("y_coord</br><small>Y coordinate</small>"),
+             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
+             int=html("int</br><small>Intensity</small>")) %>%
+  tab_row_group(label = "...millions of additional entries...", rows = x_coord>=mean(x_coord)) %>%
+  tab_row_group(label = "", rows = x_coord<=mean(x_coord)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: MS1") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/imaging_table.png", vwidth=1000)
+
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,23338,23338),
+  rt=c(0.10, 0.10, 12.37, 12.37),
+  rt_cor=c(0.10, 0.10, 13.5, 13.5)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             rt=html("rt</br><small>Retention time</small>"),
+             rt_cor=html("rt_cor</br><small>Corrected RT</small>")) %>%
+  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: rt_cor") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/rtcor_table.png", vwidth=1000)
+
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
+  rt_min=c(6.5, 12.1, 23.7, 29.5),
+  rt_max=c(7.6, 13.0, 25.0, 30.6),
+  mz_min=c(118.0862, 76.0760, 231.1005, 254.0325),
+  mz_max=c(118.0865, 76.0762, 231.1006, 254.0328),
+  peak_area=c(1104948, 1729482, 7953192, 2397.9),
+  quality=c(0.97, 0.87, 0.88, 0.24)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             feature=html("feature</br><small>Feature</small>"),
+             rt_min=html("rt_min</br><small>Min peak RT</small>"),
+             rt_max=html("rt_max</br><small>Max peak RT</small>"),
+             mz_min=html("mz_min</br><small>Min peak m/z</small>"),
+             mz_max=html("mz_max</br><small>Max peak m/z</small>"),
+             peak_area=html("peak_area</br><small>Area</small>"),
+             quality=html("qscore</br><small>Quality</small>")) %>%
+  tab_row_group(label = "...additional entries...", rows = rt_min>=mean(rt_min)) %>%
+  tab_row_group(label = "", rows = rt_min<=mean(rt_min)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: chrom_peaks") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/peak_table.png", vwidth=1000)
+
+data.frame(
+  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
+  name=c("Valine", "TMAO", "Unknown", "Unknown"),
+  formula=c("C5H11NO2", "C3H9NO", "NA", "NA"),
+  rt=c(6.5, 12.1, 23.7, 29.5),
+  mz=c(118.0862, 76.0760, 231.1005, 254.0325),
+  avg_area=c(1104948, 1729482, 7953192, 2397.9),
+  med_snr=c(8.79, 7.80, 1.72, 3.19),
+  med_qscore=c(0.97, 0.87, 0.88, 0.24)
+) %>%
+  gt() %>%
+  cols_label(feature=html("feature</br><small>Feature</small>"),
+             name=html("name</br><small>Feature ID</small>"),
+             formula=html("formula</br><small>Formula</small>"),
+             rt=html("rt</br><small>Retention time</small>"),
+             mz=html("mz</br><small>m/z</small>"),
+             avg_area=html("avg_area</br><small>Average area</small>"),
+             med_snr=html("med_snr</br><small>Median SNR</small>"),
+             med_qscore=html("med_qscore</br><small>Median quality</small>")) %>%
+  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("feature") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: features") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/feature_table.png", vwidth=1000)
+
 ```
 
-*Supplemental figure 1: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
+![](figures/supp_db_fig.png)
+
+*Supplemental figure 1: Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
 
@@
 Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number.
-The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supplemental Figure 1).
+The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supplemental Figure 2).
 

### 2025-06-11 manuscript/manuscript.Rmd
Moving supplement to end

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, include=FALSE, eval=TRUE)
+knitr::opts_chunk$set(echo = TRUE, include=FALSE, eval=FALSE)
 library(tidyverse)
@@
 2. How expensive in time and (disk) space is it to access MS data in SQLite, DuckDB, and Parquet formats?
-3. How does this cost compare to more complex formats that have been previously proposed?
+3. How does this cost compare to more complex MS storage formats that have been previously proposed?
 
@@
 
-```{r supp fig 1 gtables, eval=FALSE}
-library(gt)
-data.frame(
-  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
-  scan_num=c(1,1,651501,651501),
-  rt=c(0.10, 0.10, 12.37, 12.37),
-  dt=c(0.07, 0.07, 6.16, 6.16),
-  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
-  int=c(6618, 2657, 158084, 4673)
-) %>%
-  gt() %>%
-  cols_label(filename=html("filename</br><small>Source file</small>"),
-             scan_num=html("scan_num</br><small>Scan number</small>"),
-             rt=html("rt</br><small>Retention time</small>"),
-             dt=html("dt</br><small>Drift time</small>"),
-             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
-             int=html("int</br><small>Intensity</small>")) %>%
-  tab_row_group(label = "...billions of additional entries...", rows = rt>=mean(rt)) %>%
-  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
-  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
-  opt_table_outline() %>%
-  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
-  cols_align(align = "center") %>%
-  cols_move_to_start("filename") %>%
-  tab_options(row_group.padding = 3) %>%
-  tab_header(title = "Table: MS1") %>%
-  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/ionmob_table.png", vwidth=1000)
-
-data.frame(
-  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
-  scan_num=c(1,1,651501,651501),
-  x_coord=c(1, 1, 678, 678),
-  y_coord=c(1, 1, 327, 327),
-  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
-  int=c(6618, 2657, 158084, 4673)
-) %>%
-  gt() %>%
-  cols_label(filename=html("filename</br><small>Source file</small>"),
-             scan_num=html("scan_num</br><small>Scan number</small>"),
-             x_coord=html("x_coord</br><small>X coordinate</small>"),
-             y_coord=html("y_coord</br><small>Y coordinate</small>"),
-             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
-             int=html("int</br><small>Intensity</small>")) %>%
-  tab_row_group(label = "...millions of additional entries...", rows = x_coord>=mean(x_coord)) %>%
-  tab_row_group(label = "", rows = x_coord<=mean(x_coord)) %>%
-  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
-  opt_table_outline() %>%
-  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
-  cols_align(align = "center") %>%
-  cols_move_to_start("filename") %>%
-  tab_options(row_group.padding = 3) %>%
-  tab_header(title = "Table: MS1") %>%
-  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/imaging_table.png", vwidth=1000)
-
-data.frame(
-  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
-  scan_num=c(1,1,23338,23338),
-  rt=c(0.10, 0.10, 12.37, 12.37),
-  rt_cor=c(0.10, 0.10, 13.5, 13.5)
-) %>%
-  gt() %>%
-  cols_label(filename=html("filename</br><small>Source file</small>"),
-             scan_num=html("scan_num</br><small>Scan number</small>"),
-             rt=html("rt</br><small>Retention time</small>"),
-             rt_cor=html("rt_cor</br><small>Corrected RT</small>")) %>%
-  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
-  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
-  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
-  opt_table_outline() %>%
-  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
-  cols_align(align = "center") %>%
-  cols_move_to_start("filename") %>%
-  tab_options(row_group.padding = 3) %>%
-  tab_header(title = "Table: rt_cor") %>%
-  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/rtcor_table.png", vwidth=1000)
-
-data.frame(
-  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
-  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
-  rt_min=c(6.5, 12.1, 23.7, 29.5),
-  rt_max=c(7.6, 13.0, 25.0, 30.6),
-  mz_min=c(118.0862, 76.0760, 231.1005, 254.0325),
-  mz_max=c(118.0865, 76.0762, 231.1006, 254.0328),
-  peak_area=c(1104948, 1729482, 7953192, 2397.9),
-  quality=c(0.97, 0.87, 0.88, 0.24)
-) %>%
-  gt() %>%
-  cols_label(filename=html("filename</br><small>Source file</small>"),
-             feature=html("feature</br><small>Feature</small>"),
-             rt_min=html("rt_min</br><small>Min peak RT</small>"),
-             rt_max=html("rt_max</br><small>Max peak RT</small>"),
-             mz_min=html("mz_min</br><small>Min peak m/z</small>"),
-             mz_max=html("mz_max</br><small>Max peak m/z</small>"),
-             peak_area=html("peak_area</br><small>Area</small>"),
-             quality=html("qscore</br><small>Quality</small>")) %>%
-  tab_row_group(label = "...additional entries...", rows = rt_min>=mean(rt_min)) %>%
-  tab_row_group(label = "", rows = rt_min<=mean(rt_min)) %>%
-  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
-  opt_table_outline() %>%
-  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
-  cols_align(align = "center") %>%
-  cols_move_to_start("filename") %>%
-  tab_options(row_group.padding = 3) %>%
-  tab_header(title = "Table: chrom_peaks") %>%
-  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/peak_table.png", vwidth=1000)
-
-data.frame(
-  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
-  name=c("Valine", "TMAO", "Unknown", "Unknown"),
-  formula=c("C5H11NO2", "C3H9NO", "NA", "NA"),
-  rt=c(6.5, 12.1, 23.7, 29.5),
-  mz=c(118.0862, 76.0760, 231.1005, 254.0325),
-  avg_area=c(1104948, 1729482, 7953192, 2397.9),
-  med_snr=c(8.79, 7.80, 1.72, 3.19),
-  med_qscore=c(0.97, 0.87, 0.88, 0.24)
-) %>%
-  gt() %>%
-  cols_label(feature=html("feature</br><small>Feature</small>"),
-             name=html("name</br><small>Feature ID</small>"),
-             formula=html("formula</br><small>Formula</small>"),
-             rt=html("rt</br><small>Retention time</small>"),
-             mz=html("mz</br><small>m/z</small>"),
-             avg_area=html("avg_area</br><small>Average area</small>"),
-             med_snr=html("med_snr</br><small>Median SNR</small>"),
-             med_qscore=html("med_qscore</br><small>Median quality</small>")) %>%
-  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
-  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
-  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
-  opt_table_outline() %>%
-  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
-  cols_align(align = "center") %>%
-  cols_move_to_start("feature") %>%
-  tab_options(row_group.padding = 3) %>%
-  tab_header(title = "Table: features") %>%
-  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
-  gtsave(filename = "figures/feature_table.png", vwidth=1000)
-
-```
-
-![](figures/supp_db_fig.png)
-
-*Supplemental figure 1: Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
-
 ## Time and space testing
@@
 
-```{r singlefile_supp}
-fun_levels = c("<span style='color:#a41118'>Including initial load</span>", 
-               "<span style='color:#0b505c'>Extraction only</span>")
-init_outside_gp <- read_csv("data/singlefile_supp.csv") %>%
-  pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
-  mutate(method=factor(method, levels=method_names)) %>%
-  mutate(fun_type=factor(fun_type, levels=c("inclusive", "minimal"),
-                         labels=fun_levels)) %>%
-  ggplot() +
-  geom_boxplot(aes(x=method, y=time, color=fun_type)) +
-  scale_y_log10() +
-  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
-  labs(x=NULL, y="Time (seconds)", color=NULL) +
-  theme_bw() +
-  theme(legend.text=element_markdown(), legend.position = "inside",
-        legend.justification.inside = c(0.98, 0.95),
-        legend.background = element_rect(color='black'),
-        plot.title.position = "plot", 
-        plot.caption.position =  "plot") +
-  ggtitle("A) Scan extraction with and without pre-loading")
-
-pyopenms_precomp_gp <- read_csv("data/pyopenms_precomp.csv") %>%
-  pivot_longer(rep1:rep3) %>%
-  mutate(type=factor(type, levels=c("init", "precomp"),
-                       labels=fun_levels)) %>%
-  mutate(method=factor(method, levels=c("spec", "chrom", "chrom_2D", "rtrange",
-                                        "rtrange_2D", "premz", "fragmz"),
-                       labels = c("Spectrum", "EIC", "EIC\n(2D Peak method)",
-                                  "RT Range", "RT Range\n(2D Peak method)",
-                                  "Precursor\nsearch", "Fragment\nsearch"))) %>%
-  ggplot() +
-  geom_boxplot(aes(x=method, y=value, color=type)) +
-  scale_y_log10() +
-  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
-  theme_bw() +
-  theme(legend.position = "none", 
-        plot.title.position = "plot", 
-        plot.caption.position =  "plot") +
-  labs(x=NULL, y="Time (seconds)", color=NULL) +
-  ggtitle("B) pyopenms metric timings with and without pre-loading")
-
-init_outside_gp / pyopenms_precomp_gp
-ggsave("figures/supp_2_preload_fig.png", width = 6.5, height=5)
-```
-
-![](figures/supp_2_preload_fig.png)
-
-*Supplemental figure 2: Timing information with and without pre-loading the object into Python prior to extraction.
-Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types.
-Panel B) shows the number of seconds required for pyopenms to perform each of the queries if the initial data load was removed from the timing count.
-Error bars show 3x replicate queries of 5x different targets.*
-
 The `pymzml` package was able to extract both MS<sup>1</sup> and MS<sup>2</sup> spectra from the mzML file nearly two orders of magnitude faster than the other mzML parsers, largely due to its use of naming the scans by their number and thus avoiding the expensive scan number extraction step.
@@
 
-```{r supp fig 3 multichrom}
-multichrom_times <- read_csv("data/multichrom_times.csv")
-multichrom_baseline_df <- multichrom_times %>%
-  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
-  filter(n_chroms==1) %>%
-  summarise(time=mean(time), .by=method) %>%
-  mutate(method=factor(method, levels=method_names))
-multichrom_times %>%
-  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
-  mutate(method=factor(method, levels=method_names)) %>%
-  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
-  geom_abline(aes(slope=1, intercept=log10(time)), data=multichrom_baseline_df) +
-  geom_smooth(method="lm", formula="y~x", color="black") +
-  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
-  scale_y_log10(expand=expansion(0.1)) +
-  scale_x_log10(breaks=unique(multichrom_times$n_chroms), expand=expansion(0.1)) +
-  scale_fill_manual(breaks = method_names, values=method_colors, 
-                    aesthetics = c("fill", "color")) +
-  facet_wrap(~method, scales="free_y") +
-  labs(x="Number of chromatograms extracted", y="Time required (seconds)") +
-  theme_bw() +
-  theme(legend.position = "none")
-ggsave("figures/supp_3_multichrom_facets.png", width = 6.5, height=5)
-```
-
-![](figures/supp_3_multichrom_facets.png)
-
-*Supplemental figure 3: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
-Best-fit linear models have been added for each method and are shown behind triplicate timing measurements, while the black line shows a 1:1 slope intercepting the first data point to show nonlinear behavior with additional chromatograms.
-Transparent intervals around each best-fit line show a single standard error of the mean.*
-
 We also explored whether database queries could be improved via the use of either a unified query (single SQL statement with multiple OR clauses for each ion's *m/z* range) or a non-equi join between a peak table with *m/z* minimum and maximum columns (Supplemental Figure 4).
@@
 The opposite was true for DuckDB likely due to its optimized reader, with the unified query consistently outperformed by the non-equi join when 100 chromatograms were extracted.
-
-```{r supp fig 4 dbloopjoin}
-type_labels = c("<span style='color:#a41118'>For loop</span>", 
-               "<span style='color:#f4bb23'>Unified query</span>",
-               "<span style='color:#147C8F'>Non-equi join</span>")
-read_csv("data/dbloopjoinuni_times.csv") %>%
-  separate(method, into = c("method", "type")) %>%
-  mutate(type=case_when(
-    type=="join"~"Non-equi join",
-    type=="loop"~"For loop",
-    type=="unified"~"Unified query"
-  )) %>%
-  mutate(type=factor(type, levels=c("For loop", "Unified query", "Non-equi join"),
-                     labels=type_labels)) %>%
-  mutate(method=factor(method, levels=c("sqlite", "duckdb", "parquet"),
-                       labels=c("SQLite", "DuckDB", "Parquet"))) %>%
-  ggplot(aes(x=n_chroms, y=time, fill=type)) +
-  geom_point(position=position_dodge(width = 0.1), color="black", size=3, pch=21) +
-  facet_wrap(~method, nrow=1) +
-  scale_y_log10() +
-  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
-  scale_fill_manual(breaks=type_labels, values = c("#a41118", "#f4bb23", "#147C8F")) +
-  labs(x="Number of chromatograms", y="Time required (seconds)", fill=NULL) +
-  theme_bw() +
-  theme(legend.position = "inside", legend.justification.inside = c(0.5, 0.95),
-        legend.background = element_rect(color='black'),
-        legend.text=element_markdown(),
-        plot.title.position = "plot", 
-        plot.caption.position =  "plot")
-ggsave("figures/supp_4_dbloopjoinuni.png", width = 6.5, height=5)
-```
-
-![](figures/supp_4_dbloopjoinuni.png)
-
-*Supplemental figure 4: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color.
-Ten replicates are shown for each data point.*
 
@@
 # Supplement
+
+```{r supp fig 1 gtables, eval=FALSE}
+library(gt)
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,651501,651501),
+  rt=c(0.10, 0.10, 12.37, 12.37),
+  dt=c(0.07, 0.07, 6.16, 6.16),
+  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
+  int=c(6618, 2657, 158084, 4673)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             rt=html("rt</br><small>Retention time</small>"),
+             dt=html("dt</br><small>Drift time</small>"),
+             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
+             int=html("int</br><small>Intensity</small>")) %>%
+  tab_row_group(label = "...billions of additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: MS1") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/ionmob_table.png", vwidth=1000)
+
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,651501,651501),
+  x_coord=c(1, 1, 678, 678),
+  y_coord=c(1, 1, 327, 327),
+  mz=c(60.0452, 60.0532, 60.0456, 60.0531),
+  int=c(6618, 2657, 158084, 4673)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             x_coord=html("x_coord</br><small>X coordinate</small>"),
+             y_coord=html("y_coord</br><small>Y coordinate</small>"),
+             mz=html("mz</br><small><emph>m/z</emph> ratio</small>"),
+             int=html("int</br><small>Intensity</small>")) %>%
+  tab_row_group(label = "...millions of additional entries...", rows = x_coord>=mean(x_coord)) %>%
+  tab_row_group(label = "", rows = x_coord<=mean(x_coord)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: MS1") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/imaging_table.png", vwidth=1000)
+
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  scan_num=c(1,1,23338,23338),
+  rt=c(0.10, 0.10, 12.37, 12.37),
+  rt_cor=c(0.10, 0.10, 13.5, 13.5)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             scan_num=html("scan_num</br><small>Scan number</small>"),
+             rt=html("rt</br><small>Retention time</small>"),
+             rt_cor=html("rt_cor</br><small>Corrected RT</small>")) %>%
+  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: rt_cor") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/rtcor_table.png", vwidth=1000)
+
+data.frame(
+  filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
+  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
+  rt_min=c(6.5, 12.1, 23.7, 29.5),
+  rt_max=c(7.6, 13.0, 25.0, 30.6),
+  mz_min=c(118.0862, 76.0760, 231.1005, 254.0325),
+  mz_max=c(118.0865, 76.0762, 231.1006, 254.0328),
+  peak_area=c(1104948, 1729482, 7953192, 2397.9),
+  quality=c(0.97, 0.87, 0.88, 0.24)
+) %>%
+  gt() %>%
+  cols_label(filename=html("filename</br><small>Source file</small>"),
+             feature=html("feature</br><small>Feature</small>"),
+             rt_min=html("rt_min</br><small>Min peak RT</small>"),
+             rt_max=html("rt_max</br><small>Max peak RT</small>"),
+             mz_min=html("mz_min</br><small>Min peak m/z</small>"),
+             mz_max=html("mz_max</br><small>Max peak m/z</small>"),
+             peak_area=html("peak_area</br><small>Area</small>"),
+             quality=html("qscore</br><small>Quality</small>")) %>%
+  tab_row_group(label = "...additional entries...", rows = rt_min>=mean(rt_min)) %>%
+  tab_row_group(label = "", rows = rt_min<=mean(rt_min)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("filename") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: chrom_peaks") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/peak_table.png", vwidth=1000)
+
+data.frame(
+  feature=c("FT0001", "FT0002", "FT2311", "FT2312"),
+  name=c("Valine", "TMAO", "Unknown", "Unknown"),
+  formula=c("C5H11NO2", "C3H9NO", "NA", "NA"),
+  rt=c(6.5, 12.1, 23.7, 29.5),
+  mz=c(118.0862, 76.0760, 231.1005, 254.0325),
+  avg_area=c(1104948, 1729482, 7953192, 2397.9),
+  med_snr=c(8.79, 7.80, 1.72, 3.19),
+  med_qscore=c(0.97, 0.87, 0.88, 0.24)
+) %>%
+  gt() %>%
+  cols_label(feature=html("feature</br><small>Feature</small>"),
+             name=html("name</br><small>Feature ID</small>"),
+             formula=html("formula</br><small>Formula</small>"),
+             rt=html("rt</br><small>Retention time</small>"),
+             mz=html("mz</br><small>m/z</small>"),
+             avg_area=html("avg_area</br><small>Average area</small>"),
+             med_snr=html("med_snr</br><small>Median SNR</small>"),
+             med_qscore=html("med_qscore</br><small>Median quality</small>")) %>%
+  tab_row_group(label = "...additional entries...", rows = rt>=mean(rt)) %>%
+  tab_row_group(label = "", rows = rt<=mean(rt)) %>%
+  tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
+  opt_table_outline() %>%
+  tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
+  cols_align(align = "center") %>%
+  cols_move_to_start("feature") %>%
+  tab_options(row_group.padding = 3) %>%
+  tab_header(title = "Table: features") %>%
+  tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
+  gtsave(filename = "figures/feature_table.png", vwidth=1000)
+
+```
+
+![](figures/supp_db_fig.png)
+
+*Supplemental figure 1: Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*
+
+```{r singlefile_supp}
+fun_levels = c("<span style='color:#a41118'>Including initial load</span>", 
+               "<span style='color:#0b505c'>Extraction only</span>")
+init_outside_gp <- read_csv("data/singlefile_supp.csv") %>%
+  pivot_longer(pyteomics:mz5, names_to = "method", values_to="time") %>%
+  mutate(method=factor(method, levels=method_names)) %>%
+  mutate(fun_type=factor(fun_type, levels=c("inclusive", "minimal"),
+                         labels=fun_levels)) %>%
+  ggplot() +
+  geom_boxplot(aes(x=method, y=time, color=fun_type)) +
+  scale_y_log10() +
+  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
+  labs(x=NULL, y="Time (seconds)", color=NULL) +
+  theme_bw() +
+  theme(legend.text=element_markdown(), legend.position = "inside",
+        legend.justification.inside = c(0.98, 0.95),
+        legend.background = element_rect(color='black'),
+        plot.title.position = "plot", 
+        plot.caption.position =  "plot") +
+  ggtitle("A) Scan extraction with and without pre-loading")
+
+pyopenms_precomp_gp <- read_csv("data/pyopenms_precomp.csv") %>%
+  pivot_longer(rep1:rep3) %>%
+  mutate(type=factor(type, levels=c("init", "precomp"),
+                       labels=fun_levels)) %>%
+  mutate(method=factor(method, levels=c("spec", "chrom", "chrom_2D", "rtrange",
+                                        "rtrange_2D", "premz", "fragmz"),
+                       labels = c("Spectrum", "EIC", "EIC\n(2D Peak method)",
+                                  "RT Range", "RT Range\n(2D Peak method)",
+                                  "Precursor\nsearch", "Fragment\nsearch"))) %>%
+  ggplot() +
+  geom_boxplot(aes(x=method, y=value, color=type)) +
+  scale_y_log10() +
+  scale_color_manual(breaks=fun_levels, values = c("#a41118", "#0b505c")) +
+  theme_bw() +
+  theme(legend.position = "none", 
+        plot.title.position = "plot", 
+        plot.caption.position =  "plot") +
+  labs(x=NULL, y="Time (seconds)", color=NULL) +
+  ggtitle("B) pyopenms metric timings with and without pre-loading")
+
+init_outside_gp / pyopenms_precomp_gp
+ggsave("figures/supp_2_preload_fig.png", width = 6.5, height=5)
+```
+
+![](figures/supp_2_preload_fig.png)
+
+*Supplemental figure 2: Timing information with and without pre-loading the object into Python prior to extraction.
+Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types.
+Panel B) shows the number of seconds required for pyopenms to perform each of the queries if the initial data load was removed from the timing count.
+Error bars show 3x replicate queries of 5x different targets.*
+
+```{r supp fig 3 multichrom}
+multichrom_times <- read_csv("data/multichrom_times.csv")
+multichrom_baseline_df <- multichrom_times %>%
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
+  filter(n_chroms==1) %>%
+  summarise(time=mean(time), .by=method) %>%
+  mutate(method=factor(method, levels=method_names))
+multichrom_times %>%
+  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
+  mutate(method=factor(method, levels=method_names)) %>%
+  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
+  geom_abline(aes(slope=1, intercept=log10(time)), data=multichrom_baseline_df) +
+  geom_smooth(method="lm", formula="y~x", color="black") +
+  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
+  scale_y_log10(expand=expansion(0.1)) +
+  scale_x_log10(breaks=unique(multichrom_times$n_chroms), expand=expansion(0.1)) +
+  scale_fill_manual(breaks = method_names, values=method_colors, 
+                    aesthetics = c("fill", "color")) +
+  facet_wrap(~method, scales="free_y") +
+  labs(x="Number of chromatograms extracted", y="Time required (seconds)") +
+  theme_bw() +
+  theme(legend.position = "none")
+ggsave("figures/supp_3_multichrom_facets.png", width = 6.5, height=5)
+```
+
+![](figures/supp_3_multichrom_facets.png)
+
+*Supplemental figure 3: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes.
+Best-fit linear models have been added for each method and are shown behind triplicate timing measurements, while the black line shows a 1:1 slope intercepting the first data point to show nonlinear behavior with additional chromatograms.
+Transparent intervals around each best-fit line show a single standard error of the mean.*
+
+```{r supp fig 4 dbloopjoin}
+type_labels = c("<span style='color:#a41118'>For loop</span>", 
+               "<span style='color:#f4bb23'>Unified query</span>",
+               "<span style='color:#147C8F'>Non-equi join</span>")
+read_csv("data/dbloopjoinuni_times.csv") %>%
+  separate(method, into = c("method", "type")) %>%
+  mutate(type=case_when(
+    type=="join"~"Non-equi join",
+    type=="loop"~"For loop",
+    type=="unified"~"Unified query"
+  )) %>%
+  mutate(type=factor(type, levels=c("For loop", "Unified query", "Non-equi join"),
+                     labels=type_labels)) %>%
+  mutate(method=factor(method, levels=c("sqlite", "duckdb", "parquet"),
+                       labels=c("SQLite", "DuckDB", "Parquet"))) %>%
+  ggplot(aes(x=n_chroms, y=time, fill=type)) +
+  geom_point(position=position_dodge(width = 0.1), color="black", size=3, pch=21) +
+  facet_wrap(~method, nrow=1) +
+  scale_y_log10() +
+  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
+  scale_fill_manual(breaks=type_labels, values = c("#a41118", "#f4bb23", "#147C8F")) +
+  labs(x="Number of chromatograms", y="Time required (seconds)", fill=NULL) +
+  theme_bw() +
+  theme(legend.position = "inside", legend.justification.inside = c(0.5, 0.95),
+        legend.background = element_rect(color='black'),
+        legend.text=element_markdown(),
+        plot.title.position = "plot", 
+        plot.caption.position =  "plot")
+ggsave("figures/supp_4_dbloopjoinuni.png", width = 6.5, height=5)
+```
+
+![](figures/supp_4_dbloopjoinuni.png)
+
+*Supplemental figure 4: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color.
+Ten replicates are shown for each data point.*

### 2025-06-11 manuscript/manuscript.Rmd
Adding Anitra edits, abstract and acknowledgements

@@
 
-(200 words max)
+Current mass spectrometry (MS) data formats lack accessibility, interoperability, and performance.
+This study evaluates 10 recent MS file formats and readers across several exploratory MS analysis metrics and compares them to a simple database representation implemented in SQLite, DuckDB, and Parquet.
+We found that most existing formats severely lack the documentation required for adoption and that no existing format offers a balanced combination of speed, storage space, and simplicity.
+In contrast, our data storage schema improved data discovery and extraction by multiple orders of magnitude with minimal overhead.
+We argue that these database systems offer a performant and transparent way to store MS data for exploratory analysis while reducing technical debt and allowing mass spectrometrists to leverage recent advances in data science as our own computational complexity continues to grow.
 
@@
 
-As a result, alternative file formats are proposed practically every year.
+Alternative file formats aimed at improving data access are proposed nearly every year.
 These include direct improvements to the mzML format with indexing [@Rost2015] and better internal encoding of the data [@Bhamber2021], HDF5-based alternatives [@Bhamber2021; @Wilhelm2012; @Bilbao2023; @Tully2020; @Askenazi2017], relational databases [@Shah2010; @Bouyssie2015; @Handy2017; @Yang2022; @Beagley2009], or fully custom alternatives [@Rompp2011; @Lu2022].
-Fundamentally, these alternatives implement tradeoffs between user sanity in exchange for access speed and/or size on disk with clever compression algorithms and modern data structures that move away from the human-readable format of the mzML.
+Fundamentally, these alternatives exchange ease of use for access speed and/or size on disk with clever compression algorithms and modern data structures that move away from the human-readable format of the mzML.
 These optimized formats are inherently more difficult to understand and usually lack comprehensive documentation or examples (particularly across programming languages) making it difficult for new users to enjoy their benefits or extend their functionality.
@@
 
-Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage.
-Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity pairs is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
+Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage on disk.
+Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw *m/z* and intensity pairs is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control.
 We compare the time and space required to extract a representative data subset under six conditions and perform these tests on multiple databases as well as mzML and other MS data formats.
@@
 
+We chose to focus on liquid-chromatography mass-spectrometry (LC-MS) data given its widespread use and fundamentally simple raw data structure as tuples consisting of retention time, *m/z*, and intensity, though the tidy framework here can be extended easily to other MS data (Supplemental Figure 1).
 We performed a literature search for mass-spectrometry data formats that have been published in the last 15 years and attempted to find or construct parsers for each format in Python, a popular high-level interpreted language.
@@
 
-We browsed Metabolights [@Haug2019] for suitable LCMS datasets, looking for studies that included 100+ gigabytes of data from both full scan and MS/MS analysis.
+We browsed Metabolights [@Haug2019] for suitable LC-MS datasets, looking for studies that included 100+ gigabytes of data from both full scan and MS/MS analysis.
 We were also restricted to the Thermo Scientific .raw file format, as it was the most widely supported by alternative MS storage methods.
@@
 This decision was made after discussion with a wide variety of experts, all of whom preserved the original MS files even after conversion to another file type, indicating that a highly-performant addition is more important than direct replacement.
-Here, we map MS concepts (retention time, *m/z*, intensity, etc.) directly to database fields to make downstream processing as intuitive as possible.
+Here, we map MS concepts (retention time, drift time, spatial coordinate, *m/z*, intensity, etc.) directly to database fields to make downstream processing as intuitive as possible.
 Metadata is stored separately in `file_info` and `scan_info` tables that are linked by filename and scan number (Figure 1).
@@
 We were able to obtain or write parsers for seven different existing mass spectrometry (MS) data formats: mzML, mzMLb, mz5, mzDB, MZA, MzTree, and mzMD.
-Multiple Python packages existed for the mzML data format so we used each of the three dominant packages (pyteomics, pyOpenMS, and pymzml) and compared their timing results as well.
+Multiple Python packages exist for the mzML data format so we used each of the three dominant packages (pyteomics, pyOpenMS, and pymzml) and compared their timing results as well.
 We failed to produce parsers for the YAFMS and Shaduf file types due to complete deprecation (links to these no longer exist), the toffee file type due to its application solely to time-of-flight (TOF) data-independent acquisition (DIA) data, the Aird file type due to its current deprecation in Python and C#, and the UIMF format due to a complete lack of interface documentation.
@@
 
-mzDB and MZA both had reasonably good documentation, both formats notably providing self-contained extensions to `msconvert` for ease of conversion.
-However, both converters provide limited coverage, with mzDB notably missing support for Waters and Agilent .d files while MZA currently lacks support for AB Sciex .wiff and Bruker .baf files.
+mzDB and MZA both had extensive documentation, providing self-contained extensions to `msconvert` for ease of conversion.
+However, both converters provide limited coverage, with mzDB missing support for Waters and Agilent .d files while MZA currently lacks support for AB Sciex .wiff and Bruker .baf files.
 Both converters are only available via binary executable (.exe), restricting their use to Windows platforms.
@@
 
-The Aird file type was straightforward to convert on Windows via the executable available on Github (v6.0.0) but does not seem to be available for other operating systems, much like MZA and mzDB.
-Unfortunately the Python package designed to allow an interface to the file type has been deprecated and we were unable to install or use it in a meaningful way and were unable to reverse-engineer the file type sufficiently to compare it here.
+The Aird file type was straightforward to convert on Windows via the executable available on Github (v6.0.0) but was not available for other operating systems, much like MZA and mzDB.
+The Python package designed to allow an interface to the file type has been deprecated and we were unable to install or use it and were unable to reverse-engineer the file type sufficiently to compare it here.
 The UIMF file type from the Pacific Northwest National Lab (PNNL) provided documentation exclusively in the form of C# commands and did not supply instructions for file conversion, making it unclear what input formats were supported.
-The toffee format also provided zero documentation for conversion from other formats and was restricted to time-of-flight (TOF) data independent acquisition (DIA) MS data.
+The toffee format provided no documentation for conversion from other formats and was restricted to time-of-flight (TOF) data independent acquisition (DIA) MS data.
 Thus, we were unable to directly compare any of these three file types to the others.
@@
 Despite the relative simplicity and relevance of our queries, none of the available mass spectrometry (MS) formats had existing functions or examples of all six queries.
-Unsurprisingly, the mzML file type had the most extensive coverage but documentation and prebuilt functionality was still disappointingly sparse.
+The mzML file type had the most extensive coverage but documentation and prebuilt functionality was still sparse.
 The pyteomics package provides four "combined examples" that focus on the spectrum visualization and annotation common to proteomics research but provide minimal guidance about chromatogram or retention time range extraction.
@@
 
-mz5's documentation was remarkably sparse, especially considering it was one of the earliest mzML formats and is supported by Proteowizard.
-Crucially, the original paper [@Wilhelm2012] contains links to a website (https://software.steenlab.org/mz5) which currently returns an HTTP error 500.
-A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2) and hasn't been updated in 12 years and is predominately a simple fork of h5py [@Collette2017] with three mz5-specific commits on top.
+mz5's documentation was sparse, especially for one of the earliest mzML formats with support from Proteowizard.
+The original paper [@Wilhelm2012] contains links to a website (https://software.steenlab.org/mz5) which currently returns an HTTP error 500.
+A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2), has not been updated in 12 years, and is predominately a simple fork of h5py [@Collette2017] with three mz5-specific commits on top.
 Most problematically, we were unable to determine how mz5 stores precursor *m/z* ratios, making the fragment and precursor searches impossible.
@@
 
-MZTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server that is then be queried via an HTTP API.
+MZTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server that can then be queried via an HTTP API.
 For users without prior knowledge of HTTP request methods or exposure to programming APIs, the README is entirely unhelpful because it simply documents the API's endpoints and provides no complete query strings as examples to guide the user.
@@
 We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS<sup>1</sup>) and MS/MS (MS<sup>2</sup>) data.
-The MS<sup>1</sup> table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
+The MS<sup>1</sup> table consisted exclusively of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
 The MS2 table consisted of the same fields except that the *m/z* column was separated into precursor and fragment *m/z*.
@@
 It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number.
-Scans are not always consecutive (e.g.
-during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan number.
+Scans are not always consecutive and no metadata was obviously available that would have allowed using the index directly to a specific scan number.
 
@@
 Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3).
-Pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
-Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure 3).
+Pyteomics and pyopenms both had significant overhead upon initial load resulting in faster subsequent queries that performed better than predicted from a 1:1 extrapolation, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30.
+The methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number worse than expected from a predicted fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure 3).
 
@@
 
-```{r}
+```{r multifile timing data}
 read_csv("data/multifile_timings.csv") %>%
@@
 
-We found that SQLite queries benefitted significantly from the construction of an index on the *m/z* column when extracting chromatograms, improving lookup times by an order of magnitude (dropping from [0.3] seconds to [0.03 seconds], Figure 4).
+We found that SQLite queries benefitted significantly from the construction of an index on the *m/z* column when extracting chromatograms, improving lookup times by an order of magnitude (dropping from 0.5 seconds to 0.03 seconds, Figure 4).
 However, because the SQLite index is stored on disk alongside the data, this improvement also increased the file size by 24%.
-DuckDB also improved significantly with ordered data but to a smaller degree and required additional space to do so, likely due to the reordering resulting in worse compression in other columns such as filename or retention time (Figure 4).
-Parquet files had the smallest improvement upon data ordering and also required 50% more space.
+Parquet files had the smallest improvement upon data ordering and required 50% more space, likely due to the reordering resulting in worse compression in other columns such as filename or retention time.
+DuckDB also improved significantly with ordered data but to a smaller degree and in contrast to SQLite or Parquet sometimes actually decreased in size when ordered.
 
-Crucially, these improvements also persisted when multiple files were stored in a single database.
+These improvements also persisted when multiple files were stored in a single database.
 We built databases consisting of between 1 and 100 individual MS files and tested the time required to extract ion chromatograms from each after an index was constructed (Figure 4).
@@
 We additionally compared these values to the timings obtained from converting each file into its own database and looping over each of those to confirm the linearity of that response (Figure 4).
-Interestingly, SQLite timings were slightly better with multiple database files instead of one combined database (Figure 4).
 
-```{r}
+```{r fig 4 multifile}
 timelabs <- c("0.01 seconds", "0.1 seconds", "One second", "Ten seconds", "One minute")
@@
   scale_shape_manual(breaks=c("No index/ordering", "With index/ordering"), values=c(22, 21)) +
-  labs(x="Database size", y="Time taken", color=NULL, fill=NULL, shape=NULL) +
+  labs(x="Database size", y="Time required to extract\na random chromatogram", color=NULL, fill=NULL, shape=NULL) +
   theme_bw() +
@@
 However, performant scan-based methods struggled significantly with chromatogram extraction because these axes are inherently orthogonal to each other.
-Here, mzDB performed quite well and was competitive with the specifically optimized mzTree and mzMD formats and pyopenms performed incredibly well when extracting multiple chromatograms via its 2DPeak method.
-Chromatograms (and retention time range queries) are of course only relevant for chromatography-based workflows but this type of analysis has become increasingly popular and we expect it to continue to do so, making ion extraction increasingly important.
+mzDB was competitive with the specifically optimized mzTree and mzMD formats, illustrating its success as an axis-agnostic structure.
+We were especially impressed with the 2DPeak method in pyopenms when extracting multiple chromatograms, as it had essentially a single setup time cost after which any number of chromatograms could be extracted for free and can therefore be highly recommended for visualization applications.
+Chromatograms (and retention time range queries) are of course only relevant for chromatography-based workflows but this type of analysis has become increasingly popular, making ion extraction increasingly important.
 None of the existing MS formats we tested performed very well on MS/MS data, despite the growing availability of fragmentation data, though our use-case is oriented more towards exploration instead of comprehensive analysis.
@@
 We were also unable to test several other recent and promising formats.
-Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
+Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chromatograms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
 
@@
 First, scan metadata that was encoded within a scan instead of in a separate unit required looping over each scan to see whether it contained the information requested.
-Scan number, MS level, and retention time were all useful bits of information that could be included in a file header or footer to relate the three to the data location within the file and allow index use instead of looping over every scan.
+Scan number, MS level, and retention time were all necessary bits of information that could be included in a file header or footer to relate the three to the data location within the file and allow index use instead of looping over every scan.
 Second, needing to decode or parse a compressed *m/z*/intensity array in each scan introduced an additional overhead that was especially punishing during ion chromatogram extraction and MS/MS search.
@@
 Additionally, this method often confuses the scan number with the scan's indexes in the data structure.
-Scan number is not always consecutive (i.e.
+Scan number is not always consecutive (e.g.
 during polarity switching, multi-experiment samples, or if any filtering is performed during processing), so even if the first or second item in the structure can be queried speedily this is no guarantee that the item will contain the information of interest.
@@
 
-Of course, to claim that existing frameworks should be discarded in favor of a novel method is to ignore decades of discussion and conversation.
+Of course, to claim that existing frameworks should be discarded in favor of a novel method is to ignore decades of discussion.
 We acknowledge that our use case, that of largely exploratory and quality-control steps, is not a universal need and our lack of perfect metadata preservation in particular indicates that databases should become an auxiliary data structure alongside the vendor files or mzMLs, not substitute for them directly.
@@
 
+We would like to acknowledge the University of Washington's eScience Institute and especially Bryna Hazelton and Dave Beck for their guidance and support during this project.
+We are also grateful to Theo Portlock and the other authors of their 2025 manuscript for posting their metabolomics data to Metabolights and allowing us to reuse it.
+Finally, we would like to acknowledge Josh Sacks and other members of the Ingalls Lab for their helpful discussions and for beta-testing many parts of the project.
+
 # Data availability
 
+All data and code are available on the Github repository associated with this project at https://github.com/wkumler/mzsql under the `manuscript_things` branch.
+
 # References
+
+<div id="refs"></div>
 
@@
         plot.caption.position =  "plot")
-ggsave("figures/supp_4_dbloopjoinuni.png", width = 6.5, height=5)
+ggsave("figures/supp_4_dbloopjoinuni.png", width = 6.5, height=3)
 ```

### 2025-06-20 manuscript/manuscript.Rmd
Fixing author list

@@
 author:
-  - William Kumler
-  - Sam LaRue
-  - Anitra E. Ingalls
+  William Kumler, Sam LaRue, Anitra E. Ingalls
 output: 

### 2025-07-18 manuscript/manuscript.Rmd
Fixing MzTree capitalization

@@
 Additionally, data from multiple samples can be stored together in a single database table to permit queries of all dataset samples to be performed without looping through each file in turn.
-This differs from existing formats like mzDB, mzTree, and mzMD and thereby avoids the associated computational overhead and query complexity.
+This differs from existing formats like mzDB, MzTree, and mzMD and thereby avoids the associated computational overhead and query complexity.
 
@@
 
-MZTree and its derivative, mzMD, provided significantly less documentation about the conversion process than the other file types.
+MzTree and its derivative, mzMD, provided significantly less documentation about the conversion process than the other file types.
 This documentation consisted solely of the README available in the associated Github repositories and their installation and deployment required rebuilding the Java applet, of which the bare-bones instructions make several assumptions about the user's PATH environmental variable.
-In the case of mzMD, no documentation for installation and build was provided and this instead needed to be deduced from MZTree.
+In the case of mzMD, no documentation for installation and build was provided and this instead needed to be deduced from MzTree.
 Additionally, we ran into issues with hardware acceleration once the GUI was launched that required extensive debugging.
@@
 
-MZTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server that can then be queried via an HTTP API.
+MzTree and mzMD provide a slightly strange interface to MS data, requiring a separate Java server that can then be queried via an HTTP API.
 For users without prior knowledge of HTTP request methods or exposure to programming APIs, the README is entirely unhelpful because it simply documents the API's endpoints and provides no complete query strings as examples to guide the user.
@@
 all_timings <- read_csv("data/singlefile_times.csv") %>% 
+  print(n=Inf)
   mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
@@
 However, these comparisons are not perfect because not all files store exactly the same data.
-MZTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS<sup>1</sup> scans was 58 MB, a 23% size reduction.
+MzTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS<sup>1</sup> scans was 58 MB, a 23% size reduction.
 The SQLite, DuckDB, and Parquet formats also lack the extensive scan and file metadata that's present in the other file types, though it is difficult to estimate the fraction of disk space allocated for this (and which will depend upon the precise definition of metadata).
@@
 However, performant scan-based methods struggled significantly with chromatogram extraction because these axes are inherently orthogonal to each other.
-mzDB was competitive with the specifically optimized mzTree and mzMD formats, illustrating its success as an axis-agnostic structure.
+mzDB was competitive with the specifically optimized MzTree and mzMD formats, illustrating its success as an axis-agnostic structure.
 We were especially impressed with the 2DPeak method in pyopenms when extracting multiple chromatograms, as it had essentially a single setup time cost after which any number of chromatograms could be extracted for free and can therefore be highly recommended for visualization applications.

### 2025-07-18 manuscript/manuscript.Rmd
Fixing MS level superscripts

@@
 Pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb.
-The pyopenms package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS<sup>1</sup> data.
+The pyopenms package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS^1^ data.
 Additionally, pyOpenMS required installing an old version of the package (3.0.0), Python itself (3.11) and the numpy package (<2.0) due to more recent builds requiring AVX support which was unavailable on our hardware.
@@
 
-We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS<sup>1</sup>) and MS/MS (MS<sup>2</sup>) data.
-The MS<sup>1</sup> table consisted exclusively of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
+We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS^1^) and MS/MS (MS^2^) data.
+The MS^1^ table consisted exclusively of fields for filename, scan index, retention time, *m/z* ratio, and intensity.
 The MS2 table consisted of the same fields except that the *m/z* column was separated into precursor and fragment *m/z*.
 Although we did not extend these databases to include the metadata associated with each file and scan, the logical framework could be easily extended in future work and the metadata typically represents a small fraction of the total space within the file, allowing us to make reasonable comparisons about file size between the databases and the metadata-rich other file types.
-We also converted each file's MS<sup>1</sup> and MS2 table into Parquet representations for comparison using the same field/column schema.
+We also converted each file's MS^1^ and MS2 table into Parquet representations for comparison using the same field/column schema.
 
@@
 After establishing a connection to the database, the six queries could be asked using nearly human-readable SQL syntax.
-Requesting the thousandth MS<sup>1</sup> scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the `pandas.read_sql_query` function.
+Requesting the thousandth MS^1^ scan by number consisted simply of `SELECT * FROM MS1 WHERE id = 1000` passed along to the `pandas.read_sql_query` function.
 More complicated queries such as retention time range (`SELECT * FROM MS1 WHERE rt BETWEEN 6 AND 8`) and a precursor mass search (`SELECT * FROM MS2 WHERE premz BETWEEN 118.086 AND 118.087`) were similarly intuitive.
@@
 
-The `pymzml` package was able to extract both MS<sup>1</sup> and MS<sup>2</sup> spectra from the mzML file nearly two orders of magnitude faster than the other mzML parsers, largely due to its use of naming the scans by their number and thus avoiding the expensive scan number extraction step.
-mzDB had the only notable difference between MS<sup>1</sup> and MS<sup>2</sup> scans, performing slightly better than `pyteomics` and `pyopenms` methods for MS<sup>1</sup> data and significantly better for MS<sup>2</sup> data, placing it approximately on par with `pymzml` in taking about a tenth of a second.
+The `pymzml` package was able to extract both MS^1^ and MS^2^ spectra from the mzML file nearly two orders of magnitude faster than the other mzML parsers, largely due to its use of naming the scans by their number and thus avoiding the expensive scan number extraction step.
+mzDB had the only notable difference between MS^1^ and MS^2^ scans, performing slightly better than `pyteomics` and `pyopenms` methods for MS^1^ data and significantly better for MS^2^ data, placing it approximately on par with `pymzml` in taking about a tenth of a second.
 The simple database methods (SQLite, DuckDB, and Parquet) also fell in this ~0.1 second range, with SQLite performing most poorly and DuckDB ~10x faster.
@@
 However, these comparisons are not perfect because not all files store exactly the same data.
-MzTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS<sup>1</sup> scans was 58 MB, a 23% size reduction.
+MzTree and mzMD appear to entirely lack the MS/MS information in the sample DDA file, representing a potentially significant size reduction that's difficult to estimate though the extraction of the same file via `msconvert` containing only MS^1^ scans was 58 MB, a 23% size reduction.
 The SQLite, DuckDB, and Parquet formats also lack the extensive scan and file metadata that's present in the other file types, though it is difficult to estimate the fraction of disk space allocated for this (and which will depend upon the precise definition of metadata).

### 2025-07-18 manuscript/manuscript.Rmd
ce from thesis committee

@@
 
-We chose to focus on liquid-chromatography mass-spectrometry (LC-MS) data given its widespread use and fundamentally simple raw data structure as tuples consisting of retention time, *m/z*, and intensity, though the tidy framework here can be extended easily to other MS data (Supplemental Figure 1).
+We chose to focus on liquid-chromatography mass-spectrometry (LC-MS) data given its widespread use and fundamentally simple raw data structure with entries consisting of retention time, *m/z*, and intensity, though the tidy framework here can be extended easily to other MS data (Supplemental Figure 1).
 We performed a literature search for mass-spectrometry data formats that have been published in the last 15 years and attempted to find or construct parsers for each format in Python, a popular high-level interpreted language.
@@
 
-Timing data was obtained on an Intel Xeon CPU with two X5650 (\@2.67 GHz) processors and 24 total cores running Windows 10 Pro (64 bit version).
+Timing data was obtained on an Intel Xeon CPU with two X5650 (\@2.67 GHz) processors and 24 total cores running Windows 10 Pro (64-bit version).
 96 gigabytes of RAM (DDR3 @ 1333 MHz) were available and a solid-state drive was used for disk storage.
@@
 The simplest and most abundantly documented query was the extraction of a single spectrum.
-In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D).
+In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulable data (Figure 2A and 2D).
 Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types.
@@
 We were also unable to test several other recent and promising formats.
-Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chromatograms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
+Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzML + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chromatograms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
 
